---
title: "bone_marrow_aging"
format: html
editor: visual
---

```{r}
library(seneR)
library(Seurat)
library(EnsDb.Mmusculus.v79)
library(dplyr)
library(readr)
library(glue)
library(scDblFinder)
library(ggplot2)
library(SoupX)
library(glmGamPoi)
library(patchwork)
library(SoupX)

set.seed(42)
```

# Male bone marrow

## Loading in data

```{r}
sample_id <- "MCC8209A3_Male_BM_3"

raw_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_raw_feature_bc_matrix")

filtered_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_filtered_feature_bc_matrix")

raw_data <- Read10X(raw_path)
filtered_data <- Read10X(filtered_path)
```

```{r}
raw_gex <- raw_data$`Gene Expression`
filtered_gex <- filtered_data$`Gene Expression`

# Explicit zero-pad instead of intersect
missing_genes <- setdiff(rownames(raw_gex), rownames(filtered_gex))
if (length(missing_genes) > 0) {
  pad <- Matrix::Matrix(0,
                        nrow = length(missing_genes),
                        ncol = ncol(filtered_gex),
                        sparse = TRUE,
                        dimnames = list(missing_genes, colnames(filtered_gex)))
  filtered_gex <- rbind(filtered_gex, pad)
}

filtered_gex <- filtered_gex[rownames(raw_gex), ]
stopifnot(identical(rownames(raw_gex), rownames(filtered_gex)))
```

RUnning SoupX deconamination correction

```{r}
sc <- SoupChannel(tod = raw_gex, toc = filtered_gex, channelName = sample_id)

# Quick clustering to satisfy SoupX's requirement
quick_srt <- CreateSeuratObject(counts = filtered_gex)
quick_srt <- NormalizeData(quick_srt, verbose = FALSE)
quick_srt <- FindVariableFeatures(quick_srt, verbose = FALSE)
quick_srt <- ScaleData(quick_srt, verbose = FALSE)
quick_srt <- RunPCA(quick_srt, npcs = 30, verbose = FALSE)
quick_srt <- FindNeighbors(quick_srt, dims = 1:30, verbose = FALSE)
quick_srt <- FindClusters(quick_srt, resolution = 0.5, verbose = FALSE)

# Pull clusters as a named vector matching SoupChannel barcodes
clusters <- setNames(as.character(quick_srt$seurat_clusters), colnames(quick_srt))

#Set clusters and run estimation
sc <- setClusters(sc, clusters)
sc <- autoEstCont(sc)

sc$fit$rhoEst  # estimated contamination fraction
head(sc$soupProfile[order(-sc$soupProfile$est), ], 20)  # top contaminating genes

# Adjust counts
adj_counts <- adjustCounts(sc, roundToInt = TRUE)

soup_male_bm <- CreateSeuratObject(counts = adj_counts,
                                   project = sample_id)
```

```{r}
male_bm <- CreateSeuratObject(counts = filtered_gex,
                              project = sample_id)
```

Normalising the ADT data and adding the hashtags to a new assay

```{r}
library(Matrix)
library(matrixStats)

raw_adt <- raw_data$`Antibody Capture`
filtered_adt <- filtered_data$`Antibody Capture`

isotype_features <- grep("^Isotype", rownames(raw_adt), value = TRUE)
hto_features     <- grep("Hashtag",  rownames(raw_adt), value = TRUE)
adt_features     <- setdiff(rownames(raw_adt), hto_features)

stopifnot(all(isotype_features %in% rownames(raw_adt)))

cell_adt <- as.matrix(filtered_adt[adt_features, colnames(soup_male_bm)])

adt_log <- log1p(cell_adt)
isotype_vals <- as.matrix(adt_log[isotype_features, , drop = FALSE])
cell_isotype_median <- colMedians(isotype_vals)
adt_dsb <- sweep(adt_log, 2, cell_isotype_median, "-")
```

```{r}
hto_features <- c("Ms.Hashtag1", "Ms.Hashtag2","Ms.Hashtag3", "Ms.Hashtag4")

adt_all <- filtered_data$`Antibody Capture`[, colnames(male_bm)]
adt_features <- setdiff(rownames(adt_all), hto_features)

male_bm[["ADT"]] <- CreateAssayObject(counts = adt_all[adt_features, ])
male_bm[["ADT"]]$data <- adt_dsb

male_bm[["HTO"]] <- CreateAssayObject(counts = filtered_data$`Antibody Capture`[hto_features,
                                                                                colnames(male_bm)])
```

```{r}
male_bm <- NormalizeData(male_bm, 
                              assay = "HTO", 
                              normalization.method = "CLR", 
                              margin = 2)

male_bm <- MULTIseqDemux(male_bm, 
                              assay = "HTO", 
                              autoThresh = TRUE)

table(male_bm$MULTI_ID)
```

```{r}
# Ensure this is character, not factor, before doing anything else
male_bm$MULTI_ID <- as.character(male_bm$MULTI_ID)

hto_to_age_male <- c(
  "Ms.Hashtag1" = "3m_M",
  "Ms.Hashtag2" = "9m_M",
  "Ms.Hashtag3" = "12m_M",
  "Ms.Hashtag4" = "18m_M"
)
male_bm$age_sex <- unname(hto_to_age_male[match(male_bm$MULTI_ID, names(hto_to_age_male))])

hto_to_age <- c(
  "Ms.Hashtag1" = "3m",
  "Ms.Hashtag2" = "9m",
  "Ms.Hashtag3" = "12m",
  "Ms.Hashtag4" = "18m"
)
male_bm$age <- unname(hto_to_age[match(male_bm$MULTI_ID, names(hto_to_age))])

rm(hto_to_age_male, hto_to_age)
```

```{r}
rm(adj_counts, adt_all, adt_dsb, adt_log, cell_adt, filtered_adt, filtered_data, 
   filtered_gex, isotype_vals, pad, quick_srt, raw_adt, raw_data, raw_gex, sc,
   adt_features, cell_isotype_median, clusters, filtered_path, missing_genes)
```

## Standard QC metrics

```{r}
# Transcriptional complexity
male_bm$log10GenesPerUMI <- log10(male_bm$nFeature_RNA) / log10(male_bm$nCount_RNA)

# Mitochondrial fraction (PercentageFeatureSet returns 0–100; convert to 0–1)
male_bm$mitoRatio <- PercentageFeatureSet(male_bm, pattern = "^mt-") / 100

# Ribosomal fraction
male_bm$riboRatio <- PercentageFeatureSet(male_bm, pattern = "^Rp[sl]") / 100

# Haemoglobin fraction — marker of ambient red blood cell RNA
hb_genes <- grep("^(Hba|Hbb)", rownames(male_bm), value = TRUE)
male_bm[["percent.hb"]] <- PercentageFeatureSet(male_bm, features = hb_genes)
```

```{r}
VlnPlot(male_bm, features = "log10GenesPerUMI", group.by = "age")
VlnPlot(male_bm, features = "mitoRatio", group.by = "age")
VlnPlot(male_bm, features = "riboRatio", group.by = "age")
VlnPlot(male_bm, features = "percent.hb", group.by = "age")
```

Checking for doublets

```{r}
sce <- as.SingleCellExperiment(male_bm)
sce <- scDblFinder(sce, samples = male_bm$MULTI_ID)

# Copy the doublet classification back into the Seurat metadata
male_bm$scDblFinder.class <- sce$scDblFinder.class
male_bm$scDblFinder.score <- sce$scDblFinder.score

message(sprintf(
  "%d doublets detected (%.1f%% of cells).",
  sum(male_bm$scDblFinder.class == "doublet"),
  mean(male_bm$scDblFinder.class == "doublet") * 100
))

library(dittoSeq)

dittoBarPlot(male_bm,
             var = "scDblFinder.class",
             group.by = "MULTI_ID",
             scale = "count")
```

Looking at doublet distribution

```{r}
meta <- male_bm@meta.data

ggplot(meta, aes(x = scDblFinder.score,
                         fill = scDblFinder.class)) +
  geom_histogram(bins = 80, alpha = 0.8, position = "identity") +
  scale_fill_manual(values = c("singlet" = "#4393c3", "doublet" = "#d73027"),
                    name = "Call") +
  labs(
    title = "scDblFinder Score Distribution",
    x = "Doublet Score",
    y = "Cell Count"
  ) +
  theme_classic(base_size = 12)
```

Filtering

```{r}
n_before <- ncol(male_bm)

male_bm_filt <- subset(
  male_bm,
  scDblFinder.class == "singlet" &
    MULTI_ID %in% c("Ms.Hashtag1", "Ms.Hashtag2", "Ms.Hashtag3", "Ms.Hashtag4") &
    log10GenesPerUMI > 0.8 &
    mitoRatio < 0.2 &
    percent.hb < 10
    
)

n_after <- ncol(male_bm_filt)
message(sprintf(
  "Cell filtering: %d → %d cells retained (%.1f%% kept).",
  n_before, n_after, n_after / n_before * 100
))
```

```{r}
male_bm <- male_bm_filt
```

## Analysis of data

### Wrangling gene expression data

```{r}
DefaultAssay(male_bm) <- "RNA"
male_bm <- SCTransform(male_bm, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

male_bm <- RunPCA(male_bm,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(male_bm, reduction = "sct_pca", ndims = 30)
DimPlot(male_bm, reduction = "sct_pca")
```

```{r}
male_bm <- RunUMAP(male_bm,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:30,
                        reduction.name = "sct_umap")

DimPlot(male_bm, reduction = "sct_umap")
```

```{r}
DefaultAssay(male_bm) <- "RNA"
male_bm <- NormalizeData(male_bm)
male_bm <- FindVariableFeatures(male_bm)
male_bm <- ScaleData(male_bm, features = VariableFeatures(male_bm))
male_bm <- RunPCA(male_bm, assay = "RNA")

ElbowPlot(male_bm)
DimPlot(male_bm, reduction = "pca")

male_bm <- RunUMAP(male_bm,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(male_bm, reduction = "umap")
```

Assigning cell references based on RNA

```{r}
library(celldex)
library(SingleR)
ref <- fetchReference("immgen", "2024-02-26")
sce <- as.SingleCellExperiment(male_bm, assay = "RNA")

cell_identity <- SingleR(test = sce, ref = ref, labels = ref$label.main)
cell_identity2 <- SingleR(test = sce, ref = ref, labels = ref$label.fine)

male_bm$singleR_main_label <- cell_identity$pruned.labels
male_bm$singleR_fine_label <- cell_identity2$pruned.labels
male_bm$main_label_score <- cell_identity$scores

# Replace NAs (cells that couldn't be confidently labelled) with the string "NA"
# so they appear explicitly in plots and tables
male_bm$singleR_main_label[is.na(male_bm$singleR_main_label)] <- "NA"
male_bm$singleR_fine_label[is.na(male_bm$singleR_fine_label)] <- "NA"

rm(ref, sce, cell_identity, cell_identity2)

DimPlot(male_bm, reduction = "sct_umap", group.by = "singleR_main_label")
```

## Stem cells and neutrophils subsethgn 

```{r}
neustem_male_bm <- subset(male_bm,
                           grepl("^Neutrophils|^Stem cells", singleR_fine_label))

DimPlot(neustem_male_bm, reduction = "sct_umap", group.by = "singleR_fine_label")
```

```{r}
DefaultAssay(neustem_male_bm) <- "RNA"
neustem_male_bm <- NormalizeData(neustem_male_bm)
neustem_male_bm <- FindVariableFeatures(neustem_male_bm)
neustem_male_bm <- ScaleData(neustem_male_bm, 
                                 features = VariableFeatures(neustem_male_bm))
neustem_male_bm <- RunPCA(neustem_male_bm, assay = "RNA")

ElbowPlot(neustem_male_bm)
DimPlot(neustem_male_bm, reduction = "pca")

neustem_male_bm <- RunUMAP(neustem_male_bm,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:10,
                        reduction.name = "umap")

DimPlot(neustem_male_bm, reduction = "umap", group.by = "singleR_fine_label")
```

### Wrangling antibody data

```{r}
DefaultAssay(neustem_male_bm) <- "ADT"

neustem_male_bm <- ScaleData(neustem_male_bm, assay = "ADT")

neustem_male_bm <- RunPCA(neustem_male_bm, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(neustem_male_bm[["ADT"]]), 
                       verbose = FALSE)

DimPlot(neustem_male_bm, reduction = "adt_pca")
ElbowPlot(neustem_male_bm, reduction = "adt_pca")
```

```{r}
neustem_male_bm <- FindMultiModalNeighbors(
  neustem_male_bm,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:20, 1:20)                # adjust dims to your data
)
```

```{r}
ElbowPlot(neustem_male_bm, reduction = "adt_pca")

neustem_male_bm <- RunUMAP(neustem_male_bm, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(neustem_male_bm, reduction = "adt_umap", group.by = "singleR_main_label")

neustem_male_bm <- RunTSNE(neustem_male_bm, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(neustem_male_bm, reduction = "adt_tsne", group.by = "singleR_main_label")
```

Finding the appropriate cluster resolution and assigning it to objects

```{r}
library(clustree)
res_grid <- c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6)

for (r in res_grid) {
  neustem_male_bm <- FindClusters(neustem_male_bm, 
                                 resolution = r, 
                                 algorithm = 4, 
                                 verbose = FALSE,
                                 graph.name = "wsnn", 
                                 cluster.name = paste0("clust_", r))
}

# Plot clustree
clustree(neustem_male_bm@meta.data, prefix = "clust_")

DimPlot(neustem_male_bm, reduction = "umap", group.by = "clust_0.4")
```

```{r}
neustem_male_bm$clusters <- neustem_male_bm$clust_0.4

DimPlot(neustem_male_bm, reduction = "umap", group.by = "clusters")
DimPlot(neustem_male_bm, reduction = "umap", group.by = "age")
DimPlot(neustem_male_bm, reduction = "umap", group.by = "singleR_fine_label")

library(dittoSeq)
dittoBarPlot(neustem_male_bm,
             group.by = "clusters",
             var = "age",
             scale = "percent")
```

```{r}
library(CytoTRACE2)

expr_matrix <- as.matrix(neustem_male_bm[["RNA"]]$counts)  # or GetAssayData with layer= if needed for your version

cytotrace2_result <- cytotrace2(
  expr_matrix,
  is_seurat = FALSE,
  species   = 'mouse'
)

# making an annotation dataframe that matches input requirements for plotData function
annotation <- data.frame(cell_type = male_bm@meta.data$singleR_main_label) %>% 
  set_rownames(., colnames(male_bm))

neustem_male_bm <- AddMetaData(
  object   = neustem_male_bm,
  metadata = cytotrace2_result
)

DimPlot(neustem_male_bm,
        reduction = "umap",
        group.by = "CytoTRACE2_Potency")
```

```{r}
immgen_label_map <- c(
  "Stem cells (SC.MEP)"         = "Megakaryocyte-Erythroid Progenitor (MEP)",
  "Stem cells (MLP)"            = "Multipotent Lymphoid Progenitor",
  "Stem cells (SC.MDP)"         = "Macrophage-DC Progenitor (MDP)",
  "Stem cells (SC.CMP.DR)"      = "Common Myeloid Progenitor (CMP)",
  "Stem cells (GMP)"            = "Granulocyte-Monocyte Progenitor (GMP)",
  "Stem cells (SC.CDP)"         = "Common Dendritic Cell Progenitor (CDP)",
  "Neutrophils (GN.ARTH)"       = "Neutrophils",
  "Stem cells (SC.ST34F)"       = "Short-Term HSC (CD34+ Flt3-)",
  "Stem cells (SC.CD150-CD48-)" = "HSC subset (CD150- CD48-, SLAM MPP)",
  "Neutrophils (GN)"            = "Neutrophils",
  "Stem cells (SC.LT34F)"       = "Long-Term HSC (CD34- Flt3-)",
  "Stem cells (SC.MPP34F)"      = "Multipotent Progenitor (CD34+ Flt3-)",
  "Stem cells (SC.STSL)"        = "Short-Term HSC (Sca1+ Lin-, SLAM)",
  "Neutrophils (GN.URAC)"       = "Neutrophils",
  "Stem cells (LTHSC)"          = "Long-Term HSC (LT-HSC)",
  "Neutrophils (GN.Thio)"       = "Neutrophils",
  "Stem cells (proB.CLP)"       = "Common Lymphoid Progenitor/pro-B cell"
)

# Assuming your existing fine labels are in a metadata column, e.g. seurat_obj$immgen_fine
neustem_male_bm$finel_label_readable <- unname(immgen_label_map[neustem_male_bm$singleR_fine_label])

DimPlot(neustem_male_bm, reduction = "umap", group.by = "finel_label_readable")
```

### Trajectory analysis

```{r}
neustem_male_bm <- readRDS("aging_male_bm_neustem.rds")
```

```{r}
library(monocle3)
library(SeuratWrappers)

find_traj_genes <- function(obj){
  orig_umap <- reducedDims(obj)$UMAP
  obj <- choose_graph_segments(obj)
  reducedDims(obj)$UMAP <- orig_umap[colnames(obj), ]
  obj <- cluster_cells(obj, reduction_method = "UMAP")
  obj <- learn_graph(obj, close_loop = FALSE, use_partition = FALSE)
  obj <- order_cells(obj, reduction_method = "UMAP")
  return(obj)
}
```

```{r}
compute_pseudotime_spearman <- function(cds) {
  
  pseudotime <- cds@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
  
  expr_matrix <- exprs(cds)  # genes x cells, sparse matrix
  expr_matrix <- expr_matrix[, names(pseudotime)]  # align cell order
  
  # Remove cells with NA/Inf pseudotime (common at root-adjacent or disconnected cells)
  valid_cells <- is.finite(pseudotime)
  expr_matrix <- expr_matrix[, valid_cells]
  pt <- pseudotime[valid_cells]
  
  # Spearman correlation per gene (loop version, safest for sparse matrices)
  spearman_results <- apply(expr_matrix, 1, function(x) {
    test <- cor.test(as.numeric(x), pt, method = "spearman")
    c(rho = test$estimate, p_value = test$p.value)
  })
  
  spearman_df <- as.data.frame(t(spearman_results))
  colnames(spearman_df) <- c("rho", "p_value")
  spearman_df$gene <- rownames(spearman_df)
  spearman_df$padj <- p.adjust(spearman_df$p_value, method = "BH")
  
  # Sort by strongest correlation
  spearman_df <- spearman_df[order(-abs(spearman_df$rho)), ]
  
  return(spearman_df)
}
```

```{r}
neustem_3m_male_bm <- subset(neustem_male_bm, age == "3m")
DimPlot(neustem_3m_male_bm, reduction = "umap", group.by = "age")

neustem_3m_male_cds <- as.cell_data_set(neustem_3m_male_bm, assay = "RNA")
neustem_3m_male_cds <- preprocess_cds(neustem_3m_male_cds, num_dim = 50)
neustem_3m_male_cds <- cluster_cells(neustem_3m_male_cds)

neustem_3m_male_cds <- learn_graph(
  neustem_3m_male_cds,
  use_partition = TRUE,
  close_loop = FALSE,
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_3m_male_cds <- order_cells(neustem_3m_male_cds, reduction_method = "UMAP")
neustem_3m_male_cds <- find_traj_genes(neustem_3m_male_cds)

neustem_3m_male_genes <- graph_test(neustem_3m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_3m_male_genes <- neustem_3m_male_genes[rev(order(neustem_3m_male_genes$morans_test_statistic )), ]

spearman_3m  <- compute_pseudotime_spearman(neustem_3m_male_cds)
write.csv(spearman_3m, "male_bm_3m_spearman.csv")
```

```{r}
neustem_9m_male_bm <- subset(neustem_male_bm, age == "9m")
DimPlot(neustem_9m_male_bm, reduction = "umap", group.by = "age")

neustem_9m_male_cds <- as.cell_data_set(neustem_9m_male_bm, assay = "RNA")
neustem_9m_male_cds <- preprocess_cds(neustem_9m_male_cds, num_dim = 50)
neustem_9m_male_cds <- cluster_cells(neustem_9m_male_cds)

neustem_9m_male_cds <- learn_graph(
  neustem_9m_male_cds,
  use_partition = TRUE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_9m_male_cds <- order_cells(neustem_9m_male_cds, reduction_method = "UMAP")
neustem_9m_male_cds <- find_traj_genes(neustem_9m_male_cds)

neustem_9m_male_genes <- graph_test(neustem_9m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_9m_male_genes <- neustem_9m_male_genes[rev(order(neustem_9m_male_genes$morans_test_statistic )), ]

spearman_9m  <- compute_pseudotime_spearman(neustem_9m_male_cds)
write.csv(spearman_9m, "male_bm_9m_spearman.csv")
```

```{r}
neustem_18m_male_bm <- subset(neustem_male_bm, age == "18m")
DimPlot(neustem_18m_male_bm, reduction = "umap", group.by = "age")

neustem_18m_male_cds <- as.cell_data_set(neustem_18m_male_bm, assay = "RNA")
neustem_18m_male_cds <- preprocess_cds(neustem_18m_male_cds, num_dim = 50)
neustem_18m_male_cds <- cluster_cells(neustem_18m_male_cds)

neustem_18m_male_cds <- learn_graph(
  neustem_18m_male_cds,
  use_partition = TRUE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_18m_male_cds <- order_cells(neustem_18m_male_cds, reduction_method = "UMAP")
neustem_18m_male_cds <- find_traj_genes(neustem_18m_male_cds)

neustem_18m_male_genes <- graph_test(neustem_18m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_18m_male_genes <- neustem_18m_male_genes[rev(order(neustem_18m_male_genes$morans_test_statistic )), ]

print(neustem_18m_male_genes)

intersect(rownames(neustem_3m_male_genes), rownames(neustem_18m_male_genes))

spearman_18m <- compute_pseudotime_spearman(neustem_18m_male_cds)
write.csv(spearman_18m, "male_bm_18m_spearman.csv")
```

```{r}
print(neustem_3m_male_genes)
print(neustem_9m_male_genes)
print(neustem_18m_male_genes)

# Genes unique to each timepoint (not found in either of the other two lists)
unique_3m  <- setdiff(rownames(neustem_3m_male_genes),  
                      union(rownames(neustem_9m_male_genes), rownames(neustem_18m_male_genes)))

unique_9m  <- setdiff(rownames(neustem_9m_male_genes),  
                      union(rownames(neustem_3m_male_genes), rownames(neustem_18m_male_genes)))

unique_18m  <- setdiff(rownames(neustem_18m_male_genes),  
                      union(rownames(neustem_9m_male_genes), rownames(neustem_3m_male_genes)))

write.csv(neustem_3m_male_genes, "trajectory_3m_genes.csv")
write.csv(neustem_9m_male_genes, "trajectory_9m_genes.csv")
write.csv(neustem_18m_male_genes, "trajectory_18m_genes.csv")
```

```{r}
library(ggVennDiagram)
library(ggplot2)

gene_lists <- list(
  "3m" = rownames(neustem_3m_male_genes),
  "9m" = rownames(neustem_9m_male_genes),
  "18m" = rownames(neustem_18m_male_genes)
)

ggVennDiagram(gene_lists) +
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "")
```

```{r}
saveRDS(neustem_male_bm, "aging_male_bm_neustem.rds")

neustem_male_bm <- readRDS("aging_male_bm_neustem.rds")
```

## Neutrophils

```{r}
male_bm_no_9m <- subset(male_bm,age != "9m")
male_bm_neutrophils <- subset(male_bm_no_9m,
                              singleR_main_label == "Neutrophils" &
                                CytoTRACE2_Potency == "Differentiated")

DimPlot(male_bm_neutrophils, reduction = "umap", group.by = "age")
DimPlot(male_bm_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(male_bm_neutrophils) <- "RNA"
male_bm_neutrophils <- SCTransform(male_bm_neutrophils, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

male_bm_neutrophils <- RunPCA(male_bm_neutrophils,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(male_bm_neutrophils, reduction = "sct_pca")
DimPlot(male_bm_neutrophils, reduction = "sct_pca")
```

```{r}
male_bm_neutrophils <- RunUMAP(male_bm_neutrophils,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:5,
                        reduction.name = "sct_umap")

DimPlot(male_bm_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(male_bm_neutrophils) <- "RNA"
male_bm_neutrophils <- NormalizeData(male_bm_neutrophils)
male_bm_neutrophils <- FindVariableFeatures(male_bm_neutrophils)
male_bm_neutrophils <- ScaleData(male_bm_neutrophils, 
                                 features = VariableFeatures(male_bm_neutrophils))
male_bm_neutrophils <- RunPCA(male_bm_neutrophils, assay = "RNA")

ElbowPlot(male_bm_neutrophils)
DimPlot(male_bm_neutrophils, reduction = "pca")

male_bm_neutrophils <- RunUMAP(male_bm_neutrophils,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:10,
                        reduction.name = "umap")

DimPlot(male_bm_neutrophils, reduction = "umap", group.by = "age")
FeaturePlot(male_bm_neutrophils, reduction = "umap", features = "CytoTRACE2_Score")
```

### Wrangling antibody data

```{r}
DefaultAssay(male_bm_neutrophils) <- "ADT"

male_bm_neutrophils <- ScaleData(male_bm_neutrophils, assay = "ADT")

male_bm_neutrophils <- RunPCA(male_bm_neutrophils, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(male_bm_neutrophils[["ADT"]]), 
                       verbose = FALSE)

DimPlot(male_bm_neutrophils, reduction = "adt_pca")
ElbowPlot(male_bm_neutrophils, reduction = "adt_pca")
```

```{r}
ElbowPlot(male_bm_neutrophils, reduction = "adt_pca")

male_bm_neutrophils <- RunUMAP(male_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(male_bm_neutrophils, reduction = "adt_umap", group.by = "age")

male_bm_neutrophils <- RunTSNE(male_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(male_bm_neutrophils, reduction = "adt_tsne", group.by = "age")
```

```{r}
male_bm_neutrophils <- FindMultiModalNeighbors(
  male_bm_neutrophils,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:10, 1:10)                # adjust dims to your data
)
```

### DEG testing

```{r}
Idents(male_bm_neutrophils) <- "age"
male_age_markers <- FindAllMarkers(male_bm_neutrophils,
                                     assay = "RNA")
male_age_markers <- subset(male_age_markers,
                             p_val_adj < 0.05)

write_csv(male_age_markers, "/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_bm_analysis/male/male_age_markers.csv")
```

# Female bone marrow (v2)


## Loading in data

```{r}
sample_id <- "MCC8209A4_Female_BM_4"

raw_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_raw_feature_bc_matrix")

filtered_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_filtered_feature_bc_matrix")

raw_data <- Read10X(raw_path)
filtered_data <- Read10X(filtered_path)
```

```{r}
raw_gex <- raw_data$`Gene Expression`
filtered_gex <- filtered_data$`Gene Expression`

# Explicit zero-pad instead of intersect
missing_genes <- setdiff(rownames(raw_gex), rownames(filtered_gex))
if (length(missing_genes) > 0) {
  pad <- Matrix::Matrix(0,
                        nrow = length(missing_genes),
                        ncol = ncol(filtered_gex),
                        sparse = TRUE,
                        dimnames = list(missing_genes, colnames(filtered_gex)))
  filtered_gex <- rbind(filtered_gex, pad)
}

filtered_gex <- filtered_gex[rownames(raw_gex), ]
stopifnot(identical(rownames(raw_gex), rownames(filtered_gex)))
```

RUnning SoupX deconamination correction

```{r}
sc <- SoupChannel(tod = raw_gex, toc = filtered_gex, channelName = sample_id)

# Quick clustering to satisfy SoupX's requirement
quick_srt <- CreateSeuratObject(counts = filtered_gex)
quick_srt <- NormalizeData(quick_srt, verbose = FALSE)
quick_srt <- FindVariableFeatures(quick_srt, verbose = FALSE)
quick_srt <- ScaleData(quick_srt, verbose = FALSE)
quick_srt <- RunPCA(quick_srt, npcs = 30, verbose = FALSE)
quick_srt <- FindNeighbors(quick_srt, dims = 1:30, verbose = FALSE)
quick_srt <- FindClusters(quick_srt, resolution = 0.5, verbose = FALSE)

# Pull clusters as a named vector matching SoupChannel barcodes
clusters <- setNames(as.character(quick_srt$seurat_clusters), colnames(quick_srt))

#Set clusters and run estimation
sc <- setClusters(sc, clusters)
sc <- autoEstCont(sc)

sc$fit$rhoEst  # estimated contamination fraction
head(sc$soupProfile[order(-sc$soupProfile$est), ], 20)  # top contaminating genes

# Adjust counts
adj_counts <- adjustCounts(sc, roundToInt = TRUE)

soup_female_bm <- CreateSeuratObject(counts = adj_counts,
                                   project = sample_id)
```

```{r}
female_bm <- CreateSeuratObject(counts = filtered_gex,
                              project = sample_id)
```

Normalising the ADT data and adding the hashtags to a new assay

```{r}
library(Matrix)
library(matrixStats)

raw_adt <- raw_data$`Antibody Capture`
filtered_adt <- filtered_data$`Antibody Capture`

isotype_features <- grep("^Isotype", rownames(raw_adt), value = TRUE)
hto_features     <- grep("Hashtag",  rownames(raw_adt), value = TRUE)
adt_features     <- setdiff(rownames(raw_adt), hto_features)

stopifnot(all(isotype_features %in% rownames(raw_adt)))

cell_adt <- as.matrix(filtered_adt[adt_features, colnames(soup_female_bm)])

adt_log <- log1p(cell_adt)
isotype_vals <- as.matrix(adt_log[isotype_features, , drop = FALSE])
cell_isotype_median <- colMedians(isotype_vals)
adt_dsb <- sweep(adt_log, 2, cell_isotype_median, "-")
```

```{r}
hto_features <- c("Ms.Hashtag1", "Ms.Hashtag2","Ms.Hashtag3", "Ms.Hashtag4")

adt_all <- filtered_data$`Antibody Capture`[, colnames(soup_female_bm)]
adt_features <- setdiff(rownames(adt_all), hto_features)

female_bm[["ADT"]] <- CreateAssayObject(counts = adt_all[adt_features, ])
female_bm[["ADT"]]$data <- adt_dsb

female_bm[["HTO"]] <- CreateAssayObject(counts = filtered_data$`Antibody Capture`[hto_features,
                                                                                colnames(female_bm)])
```

```{r}
female_bm <- NormalizeData(female_bm, 
                              assay = "HTO", 
                              normalization.method = "CLR", 
                              margin = 2)

female_bm <- MULTIseqDemux(female_bm, 
                              assay = "HTO", 
                              autoThresh = TRUE)

table(female_bm$MULTI_ID)
```

```{r}
# Ensure this is character, not factor, before doing anything else
female_bm$MULTI_ID <- as.character(female_bm$MULTI_ID)

hto_to_age_female <- c(
  "Ms.Hashtag1" = "3m_F",
  "Ms.Hashtag2" = "9m_F",
  "Ms.Hashtag3" = "12m_F",
  "Ms.Hashtag4" = "18m_F"
)
female_bm$age_sex <- unname(hto_to_age_female[match(female_bm$MULTI_ID, names(hto_to_age_female))])

hto_to_age <- c(
  "Ms.Hashtag1" = "3m",
  "Ms.Hashtag2" = "9m",
  "Ms.Hashtag3" = "12m",
  "Ms.Hashtag4" = "18m"
)
female_bm$age <- unname(hto_to_age[match(female_bm$MULTI_ID, names(hto_to_age))])

rm(hto_to_age_female, hto_to_age)
```

```{r}
rm(adj_counts, adt_all, adt_dsb, adt_log, cell_adt, filtered_adt, filtered_data, 
   filtered_gex, isotype_vals, pad, quick_srt, raw_adt, raw_data, raw_gex, sc,
   adt_features, cell_isotype_median, clusters, filtered_path, missing_genes)
```

## Standard QC metrics

```{r}
# Transcriptional complexity
female_bm$log10GenesPerUMI <- log10(female_bm$nFeature_RNA) / log10(female_bm$nCount_RNA)

# Mitochondrial fraction (PercentageFeatureSet returns 0–100; convert to 0–1)
female_bm$mitoRatio <- PercentageFeatureSet(female_bm, pattern = "^mt-") / 100

# Ribosomal fraction
female_bm$riboRatio <- PercentageFeatureSet(female_bm, pattern = "^Rp[sl]") / 100

# Haemoglobin fraction — marker of ambient red blood cell RNA
hb_genes <- grep("^(Hba|Hbb)", rownames(female_bm), value = TRUE)
female_bm[["percent.hb"]] <- PercentageFeatureSet(female_bm, features = hb_genes)
```

```{r}
VlnPlot(female_bm, features = "log10GenesPerUMI", group.by = "age")
VlnPlot(female_bm, features = "mitoRatio", group.by = "age")
VlnPlot(female_bm, features = "riboRatio", group.by = "age")
VlnPlot(female_bm, features = "percent.hb", group.by = "age")
```

Checking for doublets

```{r}
sce <- as.SingleCellExperiment(female_bm)
sce <- scDblFinder(sce, samples = female_bm$MULTI_ID)

# Copy the doublet classification back into the Seurat metadata
female_bm$scDblFinder.class <- sce$scDblFinder.class
female_bm$scDblFinder.score <- sce$scDblFinder.score

message(sprintf(
  "%d doublets detected (%.1f%% of cells).",
  sum(female_bm$scDblFinder.class == "doublet"),
  mean(female_bm$scDblFinder.class == "doublet") * 100
))

library(dittoSeq)

dittoBarPlot(female_bm,
             var = "scDblFinder.class",
             group.by = "MULTI_ID",
             scale = "count")
```

Looking at doublet distribution

```{r}
meta <- female_bm@meta.data

ggplot(meta, aes(x = scDblFinder.score,
                         fill = scDblFinder.class)) +
  geom_histogram(bins = 80, alpha = 0.8, position = "identity") +
  scale_fill_manual(values = c("singlet" = "#4393c3", "doublet" = "#d73027"),
                    name = "Call") +
  labs(
    title = "scDblFinder Score Distribution",
    x = "Doublet Score",
    y = "Cell Count"
  ) +
  theme_classic(base_size = 12)
```

Filtering

```{r}
n_before <- ncol(female_bm)

female_bm_filt <- subset(
  female_bm,
  scDblFinder.class == "singlet" &
    MULTI_ID %in% c("Ms.Hashtag1", "Ms.Hashtag2", "Ms.Hashtag3", "Ms.Hashtag4") &
    log10GenesPerUMI > 0.8 &
    mitoRatio < 0.2 &
    percent.hb < 10
    
)

n_after <- ncol(female_bm_filt)
message(sprintf(
  "Cell filtering: %d → %d cells retained (%.1f%% kept).",
  n_before, n_after, n_after / n_before * 100
))
```

```{r}
female_bm <- female_bm_filt
```

## Analysis of data

### Wrangling gene expression data

```{r}
DefaultAssay(female_bm) <- "RNA"
female_bm <- SCTransform(female_bm, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

female_bm <- RunPCA(female_bm,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(female_bm, reduction = "sct_pca", ndims = 30)
DimPlot(female_bm, reduction = "sct_pca")
```

```{r}
female_bm <- RunUMAP(female_bm,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:30,
                        reduction.name = "sct_umap")

DimPlot(female_bm, reduction = "sct_umap")
```

```{r}
DefaultAssay(female_bm) <- "RNA"
female_bm <- NormalizeData(female_bm)
female_bm <- FindVariableFeatures(female_bm)
female_bm <- ScaleData(female_bm, features = VariableFeatures(female_bm))
female_bm <- RunPCA(female_bm, assay = "RNA")

ElbowPlot(female_bm)
DimPlot(female_bm, reduction = "pca")

female_bm <- RunUMAP(female_bm,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(female_bm, reduction = "umap")
```

Assigning cell references based on RNA

```{r}
library(celldex)
library(SingleR)
ref <- fetchReference("immgen", "2024-02-26")
sce <- as.SingleCellExperiment(female_bm, assay = "RNA")

cell_identity <- SingleR(test = sce, ref = ref, labels = ref$label.main)
cell_identity2 <- SingleR(test = sce, ref = ref, labels = ref$label.fine)

female_bm$singleR_main_label <- cell_identity$pruned.labels
female_bm$singleR_fine_label <- cell_identity2$pruned.labels
female_bm$main_label_score <- cell_identity$scores

# Replace NAs (cells that couldn't be confidently labelled) with the string "NA"
# so they appear explicitly in plots and tables
female_bm$singleR_main_label[is.na(female_bm$singleR_main_label)] <- "NA"
female_bm$singleR_fine_label[is.na(female_bm$singleR_fine_label)] <- "NA"

rm(ref, sce, cell_identity, cell_identity2)

DimPlot(female_bm, reduction = "umap", group.by = "singleR_main_label")
```

## Stem cells and neutrophils subset

```{r}
neustem_female_bm <- subset(female_bm,
                           grepl("^Neutrophils|^Stem cells", singleR_fine_label))

DimPlot(neustem_female_bm, reduction = "umap", group.by = "singleR_fine_label")
```

```{r}
DefaultAssay(neustem_female_bm) <- "RNA"
neustem_female_bm <- NormalizeData(neustem_female_bm)
neustem_female_bm <- FindVariableFeatures(neustem_female_bm)
neustem_female_bm <- ScaleData(neustem_female_bm, 
                                 features = VariableFeatures(neustem_female_bm))
neustem_female_bm <- RunPCA(neustem_female_bm, assay = "RNA")

ElbowPlot(neustem_female_bm)
DimPlot(neustem_female_bm, reduction = "pca")

neustem_female_bm <- RunUMAP(neustem_female_bm,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:5,
                        reduction.name = "umap")

DimPlot(neustem_female_bm, reduction = "umap", group.by = "singleR_fine_label")
```

```{r}
DefaultAssay(neustem_female_bm) <- "RNA"
neustem_female_bm <- SCTransform(neustem_female_bm, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

neustem_female_bm <- RunPCA(neustem_female_bm,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(neustem_female_bm, reduction = "sct_pca", ndims = 30)
DimPlot(neustem_female_bm, reduction = "sct_pca")

neustem_female_bm <- RunUMAP(neustem_female_bm,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:20,
                        reduction.name = "sct_umap")

DimPlot(neustem_female_bm, reduction = "sct_umap")
```

### Wrangling antibody data

```{r}
DefaultAssay(neustem_female_bm) <- "ADT"

neustem_female_bm <- ScaleData(neustem_female_bm, assay = "ADT")

neustem_female_bm <- RunPCA(neustem_female_bm, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(neustem_female_bm[["ADT"]]), 
                       verbose = FALSE)

DimPlot(neustem_female_bm, reduction = "adt_pca")
ElbowPlot(neustem_female_bm, reduction = "adt_pca")
```

```{r}
neustem_female_bm <- FindMultiModalNeighbors(
  neustem_female_bm,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:5, 1:10)                # adjust dims to your data
)
```

```{r}
ElbowPlot(neustem_female_bm, reduction = "adt_pca")

neustem_female_bm <- RunUMAP(neustem_female_bm, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(neustem_female_bm, reduction = "adt_umap", group.by = "singleR_fine_label")

neustem_female_bm <- RunTSNE(neustem_female_bm, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(neustem_female_bm, reduction = "adt_tsne", group.by = "singleR_fine_label")
```

Finding the appropriate cluster resolution and assigning it to objects

```{r}
library(clustree)
res_grid <- c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6)

for (r in res_grid) {
  neustem_female_bm <- FindClusters(neustem_female_bm, 
                                 resolution = r, 
                                 algorithm = 4, 
                                 verbose = FALSE,
                                 graph.name = "wsnn", 
                                 cluster.name = paste0("clust_", r))
}

# Plot clustree
clustree(neustem_female_bm@meta.data, prefix = "clust_")

DimPlot(neustem_female_bm, reduction = "umap", group.by = "clust_0.4")
```

```{r}
neustem_female_bm$clusters <- neustem_female_bm$clust_0.4

DimPlot(neustem_female_bm, reduction = "umap", group.by = "clusters")
DimPlot(neustem_female_bm, reduction = "umap", group.by = "age")
DimPlot(neustem_female_bm, reduction = "umap", group.by = "singleR_fine_label")

library(dittoSeq)
dittoBarPlot(neustem_female_bm,
             group.by = "clusters",
             var = "age",
             scale = "percent")
```

```{r}
library(CytoTRACE2)

expr_matrix <- as.matrix(neustem_female_bm[["RNA"]]$counts)  # or GetAssayData with layer= if needed for your version

cytotrace2_result <- cytotrace2(
  expr_matrix,
  is_seurat = FALSE,
  species   = 'mouse'
)

# making an annotation dataframe that matches input requirements for plotData function
annotation <- data.frame(cell_type = female_bm@meta.data$singleR_main_label) %>% 
  set_rownames(., colnames(female_bm))

neustem_female_bm <- AddMetaData(
  object   = neustem_female_bm,
  metadata = cytotrace2_result
)

DimPlot(neustem_female_bm,
        reduction = "umap",
        group.by = "CytoTRACE2_Potency")
```

```{r}
immgen_label_map <- c(
  "Stem cells (SC.MEP)"         = "Megakaryocyte-Erythroid Progenitor (MEP)",
  "Stem cells (MLP)"            = "Multipotent Lymphoid Progenitor",
  "Stem cells (SC.MDP)"         = "Macrophage-DC Progenitor (MDP)",
  "Stem cells (SC.CMP.DR)"      = "Common Myeloid Progenitor (CMP)",
  "Stem cells (GMP)"            = "Granulocyte-Monocyte Progenitor (GMP)",
  "Stem cells (SC.CDP)"         = "Common Dendritic Cell Progenitor (CDP)",
  "Neutrophils (GN.ARTH)"       = "Neutrophils",
  "Stem cells (SC.ST34F)"       = "Short-Term HSC (CD34+ Flt3-)",
  "Stem cells (SC.CD150-CD48-)" = "HSC subset (CD150- CD48-, SLAM MPP)",
  "Neutrophils (GN)"            = "Neutrophils",
  "Stem cells (SC.LT34F)"       = "Long-Term HSC (CD34- Flt3-)",
  "Stem cells (SC.MPP34F)"      = "Multipotent Progenitor (CD34+ Flt3-)",
  "Stem cells (SC.STSL)"        = "Short-Term HSC (Sca1+ Lin-, SLAM)",
  "Neutrophils (GN.URAC)"       = "Neutrophils",
  "Stem cells (LTHSC)"          = "Long-Term HSC (LT-HSC)",
  "Neutrophils (GN.Thio)"       = "Neutrophils",
  "Stem cells (proB.CLP)"       = "Common Lymphoid Progenitor/pro-B cell"
)

# Assuming your existing fine labels are in a metadata column, e.g. seurat_obj$immgen_fine
neustem_female_bm$finel_label_readable <- unname(immgen_label_map[neustem_female_bm$singleR_fine_label])

DimPlot(neustem_female_bm, reduction = "umap", group.by = "finel_label_readable")
```

### Trajectory analysis

```{r}
saveRDS(neustem_female_bm, "aging_female_bm_neustem.rds")
neustem_female_bm <- readRDS("aging_female_bm_neustem.rds")
```

```{r}
library(monocle3)
library(SeuratWrappers)

find_traj_genes <- function(obj){
  orig_umap <- reducedDims(obj)$UMAP
  obj <- choose_graph_segments(obj)
  reducedDims(obj)$UMAP <- orig_umap[colnames(obj), ]
  obj <- cluster_cells(obj, reduction_method = "UMAP")
  obj <- learn_graph(obj, close_loop = FALSE, use_partition = FALSE)
  obj <- order_cells(obj, reduction_method = "UMAP")
  return(obj)
}
```

```{r}
compute_pseudotime_spearman <- function(cds) {
  
  pseudotime <- cds@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
  
  expr_matrix <- exprs(cds)  # genes x cells, sparse matrix
  expr_matrix <- expr_matrix[, names(pseudotime)]  # align cell order
  
  # Remove cells with NA/Inf pseudotime (common at root-adjacent or disconnected cells)
  valid_cells <- is.finite(pseudotime)
  expr_matrix <- expr_matrix[, valid_cells]
  pt <- pseudotime[valid_cells]
  
  # Spearman correlation per gene (loop version, safest for sparse matrices)
  spearman_results <- apply(expr_matrix, 1, function(x) {
    test <- cor.test(as.numeric(x), pt, method = "spearman")
    c(rho = test$estimate, p_value = test$p.value)
  })
  
  spearman_df <- as.data.frame(t(spearman_results))
  colnames(spearman_df) <- c("rho", "p_value")
  spearman_df$gene <- rownames(spearman_df)
  spearman_df$padj <- p.adjust(spearman_df$p_value, method = "BH")
  
  # Sort by strongest correlation
  spearman_df <- spearman_df[order(-abs(spearman_df$rho)), ]
  
  return(spearman_df)
}
```

```{r}
neustem_3m_female_bm <- subset(neustem_female_bm, age == "3m")
DimPlot(neustem_3m_female_bm, reduction = "umap", group.by = "age")

neustem_3m_female_cds <- as.cell_data_set(neustem_3m_female_bm, assay = "RNA")
neustem_3m_female_cds <- preprocess_cds(neustem_3m_female_cds, num_dim = 50)
neustem_3m_female_cds <- cluster_cells(neustem_3m_female_cds)

neustem_3m_female_cds <- learn_graph(
  neustem_3m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_3m_female_cds <- order_cells(neustem_3m_female_cds, reduction_method = "UMAP")
neustem_3m_female_cds <- find_traj_genes(neustem_3m_female_cds)

neustem_3m_female_genes <- graph_test(neustem_3m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_3m_female_genes <- neustem_3m_female_genes[rev(order(neustem_3m_female_genes$morans_test_statistic )), ]

spearman_3m  <- compute_pseudotime_spearman(neustem_3m_female_cds)
write.csv(spearman_3m, "female_bm_3m_spearman.csv")
```

```{r}
neustem_12m_female_bm <- subset(neustem_female_bm, age == "12m")
DimPlot(neustem_12m_female_bm, reduction = "umap", group.by = "age")

neustem_12m_female_cds <- as.cell_data_set(neustem_12m_female_bm, assay = "RNA")
neustem_12m_female_cds <- preprocess_cds(neustem_12m_female_cds, num_dim = 50)
neustem_12m_female_cds <- cluster_cells(neustem_12m_female_cds)

neustem_12m_female_cds <- learn_graph(
  neustem_12m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_12m_female_cds <- order_cells(neustem_12m_female_cds, reduction_method = "UMAP")
neustem_12m_female_cds <- find_traj_genes(neustem_12m_female_cds)

neustem_12m_female_genes <- graph_test(neustem_12m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_12m_female_genes <- neustem_12m_female_genes[rev(order(neustem_12m_female_genes$morans_test_statistic )), ]

spearman_12m  <- compute_pseudotime_spearman(neustem_12m_female_cds)
write.csv(spearman_12m, "female_bm_12m_spearman.csv")
```

```{r}
neustem_18m_female_bm <- subset(neustem_female_bm, age == "18m")
DimPlot(neustem_18m_female_bm, reduction = "umap", group.by = "age")

neustem_18m_female_cds <- as.cell_data_set(neustem_18m_female_bm, assay = "RNA")
neustem_18m_female_cds <- preprocess_cds(neustem_18m_female_cds, num_dim = 50)
neustem_18m_female_cds <- cluster_cells(neustem_18m_female_cds)

neustem_18m_female_cds <- learn_graph(
  neustem_18m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_18m_female_cds <- order_cells(neustem_18m_female_cds, reduction_method = "UMAP")
neustem_18m_female_cds <- find_traj_genes(neustem_18m_female_cds)

neustem_18m_female_genes <- graph_test(neustem_18m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_18m_female_genes <- neustem_18m_female_genes[rev(order(neustem_18m_female_genes$morans_test_statistic )), ]

print(neustem_18m_female_genes)

intersect(rownames(neustem_3m_female_genes), rownames(neustem_18m_female_genes))

spearman_18m <- compute_pseudotime_spearman(neustem_18m_female_cds)
write.csv(spearman_18m, "female_bm_18m_spearman.csv")
```

```{r}
print(neustem_3m_female_genes)
print(neustem_12m_female_genes)
print(neustem_18m_female_genes)

# Genes unique to each timepoint (not found in either of the other two lists)
unique_3m  <- setdiff(rownames(neustem_3m_female_genes),  
                      union(rownames(neustem_12m_female_genes), rownames(neustem_18m_female_genes)))

unique_12m  <- setdiff(rownames(neustem_12m_female_genes),  
                      union(rownames(neustem_3m_female_genes), rownames(neustem_18m_female_genes)))

unique_18m  <- setdiff(rownames(neustem_18m_female_genes),  
                      union(rownames(neustem_12m_female_genes), rownames(neustem_3m_female_genes)))

write.csv(neustem_3m_female_genes, "trajectory_3m_genes.csv")
write.csv(neustem_12m_female_genes, "trajectory_12m_genes.csv")
write.csv(neustem_18m_female_genes, "trajectory_18m_genes.csv")
```

```{r}
library(ggVennDiagram)
library(ggplot2)

neustem_3m_female_genes <- subset(neustem_3m_female_genes,
                                  p_value < 0.05)

neustem_12m_female_genes <- subset(neustem_12m_female_genes,
                                  p_value < 0.05)

neustem_18m_female_genes <- subset(neustem_18m_female_genes,
                                  p_value < 0.05)

gene_lists <- list(
  "3m" = rownames(neustem_3m_female_genes),
  "12m" = rownames(neustem_12m_female_genes),
  "18m" = rownames(neustem_18m_female_genes)
)

ggVennDiagram(gene_lists) +
  scale_fill_gradient(low = "white", high = "lightblue") +
  labs(title = "")
```

```{r}
saveRDS(neustem_female_bm, "aging_female_bm_neustem.rds")
```

## Neutrophils

```{r}
female_bm_no_9m <- subset(female_bm,age != "9m")
female_bm_neutrophils <- subset(female_bm_no_9m,
                              singleR_main_label == "Neutrophils" &
                                CytoTRACE2_Potency == "Differentiated")

DimPlot(female_bm_neutrophils, reduction = "umap", group.by = "age")
DimPlot(female_bm_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(female_bm_neutrophils) <- "RNA"
female_bm_neutrophils <- SCTransform(female_bm_neutrophils, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

female_bm_neutrophils <- RunPCA(female_bm_neutrophils,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(female_bm_neutrophils, reduction = "sct_pca")
DimPlot(female_bm_neutrophils, reduction = "sct_pca")
```

```{r}
female_bm_neutrophils <- RunUMAP(female_bm_neutrophils,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:5,
                        reduction.name = "sct_umap")

DimPlot(female_bm_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(female_bm_neutrophils) <- "RNA"
female_bm_neutrophils <- NormalizeData(female_bm_neutrophils)
female_bm_neutrophils <- FindVariableFeatures(female_bm_neutrophils)
female_bm_neutrophils <- ScaleData(female_bm_neutrophils, 
                                 features = VariableFeatures(female_bm_neutrophils))
female_bm_neutrophils <- RunPCA(female_bm_neutrophils, assay = "RNA")

ElbowPlot(female_bm_neutrophils)
DimPlot(female_bm_neutrophils, reduction = "pca")

female_bm_neutrophils <- RunUMAP(female_bm_neutrophils,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:10,
                        reduction.name = "umap")

DimPlot(female_bm_neutrophils, reduction = "umap", group.by = "age")
FeaturePlot(female_bm_neutrophils, reduction = "umap", features = "CytoTRACE2_Score")
```

### Wrangling antibody data

```{r}
DefaultAssay(female_bm_neutrophils) <- "ADT"

female_bm_neutrophils <- ScaleData(female_bm_neutrophils, assay = "ADT")

female_bm_neutrophils <- RunPCA(female_bm_neutrophils, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(female_bm_neutrophils[["ADT"]]), 
                       verbose = FALSE)

DimPlot(female_bm_neutrophils, reduction = "adt_pca")
ElbowPlot(female_bm_neutrophils, reduction = "adt_pca")
```

```{r}
ElbowPlot(female_bm_neutrophils, reduction = "adt_pca")

female_bm_neutrophils <- RunUMAP(female_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(female_bm_neutrophils, reduction = "adt_umap", group.by = "age")

female_bm_neutrophils <- RunTSNE(female_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(female_bm_neutrophils, reduction = "adt_tsne", group.by = "age")
```

```{r}
female_bm_neutrophils <- FindMultiModalNeighbors(
  female_bm_neutrophils,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:10, 1:10)                # adjust dims to your data
)
```

### DEG testing

```{r}
Idents(female_bm_neutrophils) <- "age"
female_age_markers <- FindAllMarkers(female_bm_neutrophils,
                                     assay = "RNA")
female_age_markers <- subset(female_age_markers,
                             p_val_adj < 0.05)

write_csv(female_age_markers, "/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_bm_analysis/female/female_age_markers.csv")
```

# Female bone marrow

## Loading in data

```{r}
sample_id <- "MCC8209A4_Female_BM_4"

raw_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_raw_feature_bc_matrix")

filtered_path <- file.path("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/aging_BM/per_sample_outs", sample_id,
                           "sample_filtered_feature_bc_matrix")

raw_data <- Read10X(raw_path)
filtered_data <- Read10X(filtered_path)
```

```{r}
raw_gex<- raw_data$`Gene Expression`
filtered_gex <- filtered_data$`Gene Expression`

# Explicit zero-pad instead of intersect
missing_genes <- setdiff(rownames(raw_gex), rownames(filtered_gex))
if (length(missing_genes) > 0) {
  pad <- Matrix::Matrix(0,
                        nrow = length(missing_genes),
                        ncol = ncol(filtered_gex),
                        sparse = TRUE,
                        dimnames = list(missing_genes, colnames(filtered_gex)))
  filtered_gex <- rbind(filtered_gex, pad)
}

filtered_gex <- filtered_gex[rownames(raw_gex), ]
stopifnot(identical(rownames(raw_gex), rownames(filtered_gex)))

sc <- SoupChannel(tod = raw_gex, toc = filtered_gex, channelName = sample_id)
```

```{r}
# Quick clustering to satisfy SoupX's requirement
quick_srt <- CreateSeuratObject(counts = filtered_gex)
quick_srt <- NormalizeData(quick_srt, verbose = FALSE)
quick_srt <- FindVariableFeatures(quick_srt, verbose = FALSE)
quick_srt <- ScaleData(quick_srt, verbose = FALSE)
quick_srt <- RunPCA(quick_srt, npcs = 30, verbose = FALSE)
quick_srt <- FindNeighbors(quick_srt, dims = 1:30, verbose = FALSE)
quick_srt <- FindClusters(quick_srt, resolution = 0.5, verbose = FALSE)

# Pull clusters as a named vector matching SoupChannel barcodes
clusters <- setNames(as.character(quick_srt$seurat_clusters), colnames(quick_srt))

#Set clusters and run estimation
sc <- setClusters(sc, clusters)
sc <- autoEstCont(sc)

sc$fit$rhoEst  # estimated contamination fraction
head(sc$soupProfile[order(-sc$soupProfile$est), ], 20)  # top contaminating genes

# Adjust counts
adj_counts <- adjustCounts(sc, roundToInt = TRUE)

female_bm <- CreateSeuratObject(counts = adj_counts,
                                        project = sample_id)
```

```{r}
library(Matrix)
library(matrixStats)

raw_adt <- raw_data$`Antibody Capture`
filtered_adt <- filtered_data$`Antibody Capture`

isotype_features <- grep("^Isotype", rownames(raw_adt), value = TRUE)
hto_features     <- grep("Hashtag",  rownames(raw_adt), value = TRUE)
adt_features     <- setdiff(rownames(raw_adt), hto_features)

stopifnot(all(isotype_features %in% rownames(raw_adt)))

cell_adt <- as.matrix(filtered_adt[adt_features, colnames(soup_female_bm)])

adt_log <- log1p(cell_adt)
isotype_vals <- as.matrix(adt_log[isotype_features, , drop = FALSE])
cell_isotype_median <- colMedians(isotype_vals)
adt_dsb <- sweep(adt_log, 2, cell_isotype_median, "-")
```

```{r}
hto_features <- c("Ms.Hashtag1", "Ms.Hashtag2","Ms.Hashtag3", "Ms.Hashtag4")

adt_all <- filtered_data$`Antibody Capture`[, colnames(female_bm)]
adt_features <- setdiff(rownames(adt_all), hto_features)

soup_female_bm[["ADT"]] <- CreateAssayObject(counts = adt_all[adt_features, ])
soup_female_bm[["ADT"]]$data <- adt_dsb

soup_female_bm[["HTO"]] <- CreateAssayObject(counts = filtered_data$`Antibody Capture`[hto_features,
                                                                                    colnames(female_bm)])
```

```{r}
female_bm <- soup_female_bm

female_bm <- NormalizeData(female_bm, 
                              assay = "HTO", 
                              normalization.method = "CLR", 
                              margin = 2)

female_bm <- MULTIseqDemux(female_bm, 
                              assay = "HTO", 
                              autoThresh = TRUE)

table(female_bm$MULTI_ID)
```

```{r}
# Ensure this is character, not factor, before doing anything else
female_bm$MULTI_ID <- as.character(female_bm$MULTI_ID)

hto_to_age_female <- c(
  "Ms.Hashtag1" = "3m_F",
  "Ms.Hashtag2" = "9m_F",
  "Ms.Hashtag3" = "12m_F",
  "Ms.Hashtag4" = "18m_F"
)
female_bm$age_sex <- unname(hto_to_age_female[match(female_bm$MULTI_ID, names(hto_to_age_female))])

hto_to_age <- c(
  "Ms.Hashtag1" = "3m",
  "Ms.Hashtag2" = "9m",
  "Ms.Hashtag3" = "12m",
  "Ms.Hashtag4" = "18m"
)
female_bm$age <- unname(hto_to_age[match(female_bm$MULTI_ID, names(hto_to_age))])

rm(hto_to_age_female, hto_to_age)
```

## Standard QC metrics

```{r}
# Transcriptional complexity
female_bm$log10GenesPerUMI <- log10(female_bm$nFeature_RNA) / log10(female_bm$nCount_RNA)

# Mitochondrial fraction (PercentageFeatureSet returns 0–100; convert to 0–1)
female_bm$mitoRatio <- PercentageFeatureSet(female_bm, pattern = "^mt-") / 100

# Ribosomal fraction
female_bm$riboRatio <- PercentageFeatureSet(female_bm, pattern = "^Rp[sl]") / 100

# Haemoglobin fraction — marker of ambient red blood cell RNA
hb_genes <- grep("^(Hba|Hbb)", rownames(female_bm), value = TRUE)
female_bm[["percent.hb"]] <- PercentageFeatureSet(female_bm, features = hb_genes)
```

```{r}
VlnPlot(female_bm, features = "log10GenesPerUMI", group.by = "age")
VlnPlot(female_bm, features = "mitoRatio", group.by = "age")
VlnPlot(female_bm, features = "riboRatio", group.by = "age")
VlnPlot(female_bm, features = "percent.hb", group.by = "age")
```

Checking for doublets

```{r}
sce <- as.SingleCellExperiment(female_bm)
sce <- scDblFinder(sce, samples = female_bm$MULTI_ID)

# Copy the doublet classification back into the Seurat metadata
female_bm$scDblFinder.class <- sce$scDblFinder.class
female_bm$scDblFinder.score <- sce$scDblFinder.score

message(sprintf(
  "%d doublets detected (%.1f%% of cells).",
  sum(female_bm$scDblFinder.class == "doublet"),
  mean(female_bm$scDblFinder.class == "doublet") * 100
))

library(dittoSeq)

dittoBarPlot(female_bm,
             var = "scDblFinder.class",
             group.by = "MULTI_ID",
             scale = "count")
```

Looking at doublet distribution

```{r}
meta <- female_bm@meta.data

ggplot(meta, aes(x = scDblFinder.score,
                         fill = scDblFinder.class)) +
  geom_histogram(bins = 80, alpha = 0.8, position = "identity") +
  scale_fill_manual(values = c("singlet" = "#4393c3", "doublet" = "#d73027"),
                    name = "Call") +
  labs(
    title = "scDblFinder Score Distribution",
    x = "Doublet Score",
    y = "Cell Count"
  ) +
  theme_classic(base_size = 12)
```

Filtering

```{r}
n_before <- ncol(female_bm)

female_bm_filt <- subset(
  female_bm,
  scDblFinder.class == "singlet" &
    MULTI_ID %in% c("Ms.Hashtag1", "Ms.Hashtag2", "Ms.Hashtag3", "Ms.Hashtag4") &
    log10GenesPerUMI > 0.8 &
    mitoRatio < 0.2 &
    percent.hb < 10
    
)

n_after <- ncol(female_bm_filt)
message(sprintf(
  "Cell filtering: %d → %d cells retained (%.1f%% kept).",
  n_before, n_after, n_after / n_before * 100
))
```

```{r}
female_bm <- female_bm_filt
```

## Analysis of data

### Wrangling gene expression data

```{r}
DefaultAssay(female_bm) <- "RNA"
female_bm <- SCTransform(female_bm, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

female_bm <- RunPCA(female_bm,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(female_bm, reduction = "sct_pca", ndims = 30)
DimPlot(female_bm, reduction = "sct_pca")
```

```{r}
female_bm <- RunUMAP(female_bm,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:30,
                        reduction.name = "sct_umap")

DimPlot(female_bm, reduction = "sct_umap")
```

```{r}
DefaultAssay(female_bm) <- "RNA"
female_bm <- NormalizeData(female_bm)
female_bm <- FindVariableFeatures(female_bm)
female_bm <- ScaleData(female_bm, features = VariableFeatures(female_bm))
female_bm <- RunPCA(female_bm, assay = "RNA")

ElbowPlot(female_bm)
DimPlot(female_bm, reduction = "pca")

female_bm <- RunUMAP(female_bm,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(female_bm, reduction = "umap")
```

Assigning cell references based on RNA

```{r}
library(celldex)
library(SingleR)
ref <- fetchReference("immgen", "2024-02-26")
sce <- as.SingleCellExperiment(female_bm, assay = "RNA")

cell_identity <- SingleR(test = sce, ref = ref, labels = ref$label.main)
cell_identity2 <- SingleR(test = sce, ref = ref, labels = ref$label.fine)

female_bm$singleR_main_label <- cell_identity$pruned.labels
female_bm$singleR_fine_label <- cell_identity2$pruned.labels
female_bm$main_label_score <- cell_identity$scores

# Replace NAs (cells that couldn't be confidently labelled) with the string "NA"
# so they appear explicitly in plots and tables
female_bm$singleR_main_label[is.na(female_bm$singleR_main_label)] <- "NA"
female_bm$singleR_fine_label[is.na(female_bm$singleR_fine_label)] <- "NA"

rm(ref, sce, cell_identity, cell_identity2)

DimPlot(female_bm, reduction = "sct_umap", group.by = "singleR_main_label")
```

### Wrangling antibody data

```{r}
DefaultAssay(female_bm) <- "ADT"

female_bm <- ScaleData(female_bm, assay = "ADT")

female_bm <- RunPCA(female_bm, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(female_bm[["ADT"]]), 
                       verbose = FALSE)

DimPlot(female_bm, reduction = "adt_pca")
ElbowPlot(female_bm, reduction = "adt_pca")
```

```{r}
ElbowPlot(female_bm, reduction = "adt_pca")

female_bm <- RunUMAP(female_bm, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(female_bm, reduction = "adt_umap", group.by = "singleR_main_label")

female_bm <- RunTSNE(female_bm, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(female_bm, reduction = "adt_tsne", group.by = "singleR_main_label")
```

```{r}
female_bm <- FindMultiModalNeighbors(
  female_bm,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:20, 1:20)                # adjust dims to your data
)
```

Finding the appropriate cluster resolution and assigning it to objects

```{r}
library(clustree)
res_grid <- c(0.2, 0.4, 0.6, 0.8, 1.0, 1.2, 1.4, 1.6)

for (r in res_grid) {
  female_bm <- FindClusters(female_bm, 
                                 resolution = r, 
                                 algorithm = 4, 
                                 verbose = FALSE,
                                 graph.name = "wsnn", 
                                 cluster.name = paste0("clust_", r))
}

# Plot clustree
clustree(female_bm@meta.data, prefix = "clust_")

DimPlot(female_bm, reduction = "umap", group.by = "clust_0.4")
```

```{r}
female_bm$clusters <- female_bm$clust_0.4

DimPlot(female_bm, reduction = "umap", group.by = "clusters")
DimPlot(female_bm, reduction = "sct_umap", group.by = "clusters")

DimPlot(female_bm, reduction = "umap", group.by = "age")
DimPlot(female_bm, reduction = "sct_umap", group.by = "age")

DimPlot(female_bm, reduction = "umap", group.by = "singleR_main_label")
DimPlot(female_bm, reduction = "sct_umap", group.by = "singleR_main_label")

library(dittoSeq)
dittoBarPlot(female_bm,
             group.by = "clusters",
             var = "age",
             scale = "percent")
```

```{r}
library(CytoTRACE2)

expr_matrix <- as.matrix(female_bm[["RNA"]]$counts)  # or GetAssayData with layer= if needed for your version

cytotrace2_result <- cytotrace2(
  expr_matrix,
  is_seurat = FALSE,
  species   = 'mouse'
)

# making an annotation dataframe that matches input requirements for plotData function
annotation <- data.frame(cell_type = female_bm@meta.data$singleR_main_label) %>% 
  set_rownames(., colnames(female_bm))

female_bm <- AddMetaData(
  object   = female_bm,
  metadata = cytotrace2_result
)

DimPlot(female_bm,
        reduction = "umap",
        group.by = "CytoTRACE2_Potency")
```

```{r}
DimPlot(subset(neustem_male_bm,
                finel_label_readable %in% c("Neutrophils", "Granulocyte-Monocyte Progenitor (GMP)"))
               , reduction = "umap", group.by = "age")
```


### Neutrophils

```{r}
female_bm_no_9m <- subset(female_bm,age != "9m")
female_bm_neutrophils <- subset(female_bm_no_9m,
                              singleR_main_label == "Neutrophils" &
                                CytoTRACE2_Potency == "Differentiated")

DimPlot(female_bm_neutrophils, reduction = "umap", group.by = "age")
DimPlot(female_bm_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(female_bm_neutrophils) <- "RNA"
female_bm_neutrophils <- SCTransform(female_bm_neutrophils, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

female_bm_neutrophils <- RunPCA(female_bm_neutrophils,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(female_bm_neutrophils, reduction = "sct_pca")
DimPlot(female_bm_neutrophils, reduction = "sct_pca")
```

```{r}
female_bm_neutrophils <- RunUMAP(female_bm_neutrophils,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:5,
                        reduction.name = "sct_umap")

DimPlot(female_bm_neutrophils, reduction = "sct_umap", group.by = "age")
DimPlot(female_bm_neutrophils, reduction = "sct_pca", group.by = "age")
```

```{r}
DefaultAssay(female_bm_neutrophils) <- "RNA"
female_bm_neutrophils <- NormalizeData(female_bm_neutrophils)
female_bm_neutrophils <- FindVariableFeatures(female_bm_neutrophils)
female_bm_neutrophils <- ScaleData(female_bm_neutrophils, 
                                 features = VariableFeatures(female_bm_neutrophils))
female_bm_neutrophils <- RunPCA(female_bm_neutrophils, assay = "RNA")

ElbowPlot(female_bm_neutrophils)
DimPlot(female_bm_neutrophils, reduction = "pca")

female_bm_neutrophils <- RunUMAP(female_bm_neutrophils,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:10,
                        reduction.name = "umap")

DimPlot(female_bm_neutrophils, reduction = "umap", group.by = "age")
FeaturePlot(female_bm_neutrophils, reduction = "umap", features = "CytoTRACE2_Score")
```

### Wrangling antibody data

```{r}
DefaultAssay(female_bm_neutrophils) <- "ADT"

female_bm_neutrophils <- ScaleData(female_bm_neutrophils, assay = "ADT")

female_bm_neutrophils <- RunPCA(female_bm_neutrophils, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(female_bm_neutrophils[["ADT"]]), 
                       verbose = FALSE)

DimPlot(female_bm_neutrophils, reduction = "adt_pca")
ElbowPlot(female_bm_neutrophils, reduction = "adt_pca")
```

```{r}
ElbowPlot(female_bm_neutrophils, reduction = "adt_pca")

female_bm_neutrophils <- RunUMAP(female_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(female_bm_neutrophils, reduction = "adt_umap", group.by = "age")

female_bm_neutrophils <- RunTSNE(female_bm_neutrophils, 
                        dims = 1:10,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(female_bm_neutrophils, reduction = "adt_tsne", group.by = "age")
```

```{r}
female_bm_neutrophils <- FindMultiModalNeighbors(
  female_bm_neutrophils,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:10, 1:10)                # adjust dims to your data
)
```

# Merged data

```{r}
male_bm <- AddMetaData(male_bm,
                          metadata = "male",
                          col.name = "sex")

female_bm <- AddMetaData(female_bm,
                            metadata = "female",
                            col.name = "sex")

bm_merge <- merge(female_bm, 
                     male_bm,
                     merge.data = TRUE)
```

```{r}
DefaultAssay(bm_merge) <- "RNA"
bm_merge <- SCTransform(bm_merge, 
                            vst.flavor = "v2", 
                            conserve.memory = TRUE, 
                            method = "glmGamPoi") # SCT normalisation for clustering

bm_merge <- RunPCA(bm_merge,
                       assay = "SCT",
                       reduction.name = "sct_pca")

ElbowPlot(bm_merge, reduction = "sct_pca", ndims = 30)
DimPlot(bm_merge, reduction = "sct_pca", group.by = "sex")
```

```{r}
bm_merge <- RunUMAP(bm_merge,
                        assay = "SCT",
                        reduction = "sct_pca",
                        dims = 1:30,
                        reduction.name = "sct_umap")

DimPlot(bm_merge, reduction = "sct_umap", group.by = "sex", split.by = "age")
```

```{r}
DefaultAssay(bm_merge) <- "RNA"
bm_merge <- NormalizeData(bm_merge)
bm_merge <- FindVariableFeatures(bm_merge)
bm_merge <- ScaleData(bm_merge, features = VariableFeatures(bm_merge))
bm_merge <- RunPCA(bm_merge, assay = "RNA")

ElbowPlot(bm_merge)
DimPlot(bm_merge, reduction = "pca")

bm_merge <- RunUMAP(bm_merge,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(bm_merge, reduction = "umap", group.by = "sex")
DimPlot(bm_merge, reduction = "umap", group.by = "CytoTRACE2_Potency")
```

```{r}
DefaultAssay(bm_merge) <- "ADT"

bm_merge <- ScaleData(bm_merge, assay = "ADT")

bm_merge <- RunPCA(bm_merge, 
                       reduction.name = "adt_pca", 
                       assay = "ADT", 
                       features = rownames(bm_merge[["ADT"]]), 
                       verbose = FALSE)

DimPlot(bm_merge, reduction = "adt_pca")
ElbowPlot(bm_merge, reduction = "adt_pca")
```

```{r}
ElbowPlot(bm_merge, reduction = "adt_pca")

bm_merge <- RunUMAP(bm_merge, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_umap")

DimPlot(bm_merge, reduction = "adt_umap", group.by = "singleR_main_label")

bm_merge <- RunTSNE(bm_merge, 
                        dims = 1:15,
                        reduction = "adt_pca",
                        reduction.name = "adt_tsne",
                        perplexity = 50)

DimPlot(bm_merge, reduction = "adt_tsne", group.by = "sex")
```

```{r}
bm_merge <- FindMultiModalNeighbors(
  bm_merge,
  reduction.list = list("pca", "adt_pca"),      # RNA PCA + ADT PCA
  dims.list = list(1:20, 1:20)                # adjust dims to your data
)
```

## Stem cells and neutrophils

```{r}

```

## Neutrophils

```{r}
bm_merge_no_9m <- subset(bm_merge,age != "9m")
bm_merge_neutrophils <- subset(bm_merge_no_9m,
                              singleR_main_label == "Neutrophils" &
                                CytoTRACE2_Potency == "Differentiated")

DimPlot(bm_merge_neutrophils, reduction = "umap", group.by = "age")
DimPlot(bm_merge_neutrophils, reduction = "sct_umap", group.by = "age")
```

```{r}
DefaultAssay(bm_merge_neutrophils) <- "RNA"
bm_merge_neutrophils <- NormalizeData(bm_merge_neutrophils)
bm_merge_neutrophils <- FindVariableFeatures(bm_merge_neutrophils)
bm_merge_neutrophils <- ScaleData(bm_merge_neutrophils, 
                                  features = VariableFeatures(bm_merge_neutrophils))
bm_merge_neutrophils <- RunPCA(bm_merge_neutrophils, assay = "RNA")

ElbowPlot(bm_merge_neutrophils)
DimPlot(bm_merge_neutrophils, reduction = "pca")

bm_merge_neutrophils <- RunUMAP(bm_merge_neutrophils,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(bm_merge_neutrophils, reduction = "umap", group.by = "sex")
FeaturePlot(bm_merge_neutrophils, 
            reduction = "umap", 
            features = "CytoTRACE2_Score")
```

```{r}
Idents(bm_merge_neutrophils) <- bm_merge_neutrophils$age_sex
bm_merge_neutrophils <- JoinLayers(bm_merge_neutrophils)

DefaultAssay(bm_merge_neutrophils) <- "RNA"
bm_merge_neutrophils <- NormalizeData(bm_merge_neutrophils)

ages   <- c("3m", "12m", "18m")
assays <- c("RNA", "ADT")

degs_list <- list()

for (assay in assays) {
  DefaultAssay(bm_merge_neutrophils) <- assay
  degs_list[[assay]] <- list()
  
  for (age in ages) {
    degs <- FindMarkers(bm_merge_neutrophils,
                         ident.1 = paste0(age, "_M"),
                         ident.2 = paste0(age, "_F"),
                        min.pct = 0.05,
                        logfc.threshold = 0,
                        min.cells.feature = 3,
                        min.cells.group = 3)
  
    degs$gene <- rownames(degs)   # keep gene/ADT names as a column, not just rownames
    
    degs_list[[assay]][[age]] <- degs
    
    write.csv(degs,
              file = paste0("/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_bm_analysis/degs_", age, "_", assay, ".csv"),
              row.names = FALSE)
  }
}
```

```{r}
degs_3m <- degs_list[["RNA"]][["3m"]]

degs_3m_dedup <- degs_3m %>%
  group_by(gene) %>%
  slice_max(abs(avg_log2FC), n = 1, with_ties = FALSE) %>%
  ungroup()

# Split by direction (male = ident.1, so positive avg_log2FC = up in male)
sig_up_3m_male <- degs_3m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
  pull(gene)

sig_up_3m_female <- degs_3m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC < 0) %>%
  pull(gene)

length(sig_up_3m_male)
length(sig_up_3m_female)

# Universe stays the same for both — all genes tested, regardless of direction
universe_genes <- degs_3m_dedup$gene

# ORA: terms enriched among genes upregulated in MALE
ego_male_3m <- enrichGO(
  gene          = sig_up_3m_male,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_3m_male <- ego_male_3m@result %>%
  dplyr::filter(pvalue < 0.05)

# ORA: terms enriched among genes upregulated in FEMALE
ego_female_3m <- enrichGO(
  gene          = sig_up_3m_female,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_3m_female <- ego_female_3m@result %>%
  dplyr::filter(pvalue < 0.05)
```

```{r}
degs_12m <- degs_list[["RNA"]][["12m"]]

degs_12m_dedup <- degs_12m %>%
  group_by(gene) %>%
  slice_max(abs(avg_log2FC), n = 1, with_ties = FALSE) %>%
  ungroup()

# Split by direction (male = ident.1, so positive avg_log2FC = up in male)
sig_up_12m_male <- degs_12m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
  pull(gene)

sig_up_12m_female <- degs_12m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC < 0) %>%
  pull(gene)

length(sig_up_12m_male)
length(sig_up_12m_female)

# Universe stays the same for both — all genes tested, regardless of direction
universe_genes <- degs_12m_dedup$gene

# ORA: terms enriched among genes upregulated in MALE
ego_male_12m <- enrichGO(
  gene          = sig_up_12m_male,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_12m_male <- ego_male_12m@result %>%
  dplyr::filter(pvalue < 0.05)

# ORA: terms enriched among genes upregulated in FEMALE
ego_female_12m <- enrichGO(
  gene          = sig_up_12m_female,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_12m_female <- ego_female_12m@result %>%
  dplyr::filter(pvalue < 0.05)
```

```{r}
degs_18m <- degs_list[["RNA"]][["18m"]]

degs_18m_dedup <- degs_18m %>%
  group_by(gene) %>%
  slice_max(abs(avg_log2FC), n = 1, with_ties = FALSE) %>%
  ungroup()

# Split by direction (male = ident.1, so positive avg_log2FC = up in male)
sig_up_18m_male <- degs_18m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
  pull(gene)

sig_up_18m_female <- degs_18m_dedup %>%
  dplyr::filter(p_val_adj < 0.05, avg_log2FC < 0) %>%
  pull(gene)

length(sig_up_18m_male)
length(sig_up_18m_female)

# Universe stays the same for both — all genes tested, regardless of direction
universe_genes <- degs_18m_dedup$gene

# ORA: terms enriched among genes upregulated in MALE
ego_male_18m <- enrichGO(
  gene          = sig_up_18m_male,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_18m_male <- ego_male_18m@result %>%
  dplyr::filter(pvalue < 0.05)

# ORA: terms enriched among genes upregulated in FEMALE
ego_female_18m <- enrichGO(
  gene          = sig_up_18m_female,
  universe      = universe_genes,
  OrgDb         = org.Mm.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

terms_18m_female <- ego_female_18m@result %>%
  dplyr::filter(pvalue < 0.05)
```

```{r}
saveRDS(bm_merge_neutrophils, "bm_merge_neutrophils.rds")

DefaultAssay(bm_merge_neutrophils) <- "RNA"
```


### Merged trajectories

```{r}
gmp_neus_male <- subset(neustem_male_bm,
                        finel_label_readable %in% c("Granulocyte-Monocyte Progenitor (GMP)", "Neutrophils"))

DimPlot(gmp_neus_male, reduction = "umap")

gmp_neus_female <- subset(neustem_female_bm,
                        finel_label_readable %in% c("Granulocyte-Monocyte Progenitor (GMP)", "Neutrophils"))

DimPlot(gmp_neus_female, reduction = "umap")

gmp_neus_male <- AddMetaData(gmp_neus_male,
                               metadata = "male",
                               col.name = "sex")

gmp_neus_female <- AddMetaData(gmp_neus_female,
                            metadata = "female",
                            col.name = "sex")

gmp_neus_female$MULTI_ID <- as.character(gmp_neus_female$MULTI_ID)

hto_to_age_female <- c(
  "Ms.Hashtag1" = "3m_F",
  "Ms.Hashtag2" = "9m_F",
  "Ms.Hashtag3" = "12m_F",
  "Ms.Hashtag4" = "18m_F"
)
gmp_neus_female$age_sex <- unname(hto_to_age_female[match(gmp_neus_female$MULTI_ID, names(hto_to_age_female))])


gmps_neus_all <- merge(gmp_neus_male, gmp_neus_female)
```

```{r}
DefaultAssay(gmps_neus_all) <- "RNA"
gmps_neus_all <- NormalizeData(gmps_neus_all)
gmps_neus_all <- FindVariableFeatures(gmps_neus_all)
gmps_neus_all <- ScaleData(gmps_neus_all, 
                                  features = VariableFeatures(gmps_neus_all))
gmps_neus_all <- RunPCA(gmps_neus_all, assay = "RNA")

ElbowPlot(gmps_neus_all)
DimPlot(gmps_neus_all, reduction = "pca")

gmps_neus_all <- RunUMAP(gmps_neus_all,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(gmps_neus_all, reduction = "umap", group.by = "finel_label_readable")
DimPlot(gmps_neus_all, reduction = "umap", group.by = "sex")
DimPlot(gmps_neus_all, reduction = "umap", group.by = "age")
```

```{r}
neu_modules <- list(
  proNeu = c("Elane","Mpo","Prtn3","Ctsg","Ms4a3","Cebpe","Gfi1",
             "Fcnb","Rab44","Nkg7","Plac8","Cd34","Kit","Srgn"),

  proNeu2 = c("Il5ra","Cebpe","Fcnb","Ltf","Camp","Mki67"),

  preNeu = c("Ltf","Camp","Ngp","Lcn2","Chil3","Cebpe","Fcnb",
             "Anxa1","Hp","Ifitm6","Mki67"),

  immNeu = c("Ltf","Camp","Lcn2","Mmp8","Cd177","Ifitm1",
             "Chil3","Cd101"),

  matNeu = c("Cxcr2","Sell","Il1b","Csf3r","Mmp9","S100a8","S100a9",
             "Retnlg","Ifitm1","Msrb1","Slpi","Fpr1"),

  cycling = c("Mki67","Top2a","Ccnb1","Ccna2","Birc5","Ube2c","Cdk1")
)

gmps_neus_all <- AddModuleScore(
  gmps_neus_all,
  features = neu_modules,
  name     = "MOD_",
  ctrl     = 50,
  seed     = 42
)

# rename MOD_1..MOD_6 back to something readable
idx <- grep("^MOD_", colnames(gmps_neus_all@meta.data))
colnames(gmps_neus_all@meta.data)[idx] <- paste0(names(neu_modules), "_score")
```


```{r}
FeaturePlot(gmps_neus_all, reduction = "umap", features = "proNeu_score")
FeaturePlot(gmps_neus_all, reduction = "umap", features = "proNeu2_score")
FeaturePlot(gmps_neus_all, reduction = "umap", features = "preNeu_score")
FeaturePlot(gmps_neus_all, reduction = "umap", features = "immNeu_score")
FeaturePlot(gmps_neus_all, reduction = "umap", features = "matNeu_score")
FeaturePlot(gmps_neus_all, reduction = "umap", features = "cycling_score")

DimPlot(gmps_neus_all, reduction = "umap", group.by = "seurat_clusters")
```


```{r}
```

```{r}
DefaultAssay(gmps_neus_all) <- "RNA"
gmps_neus_all <- FindNeighbors(gmps_neus_all)

gmps_neus_all <- FindClusters(gmps_neus_all, resolution = 0.4)
DimPlot(gmps_neus_all, reduction = "umap", group.by = "seurat_clusters")
DimPlot(gmps_neus_all, reduction = "umap", group.by = "sex")
VlnPlot(gmps_neus_all, features = "proNeu_score")
VlnPlot(gmps_neus_all, features = "proNeu2_score")
VlnPlot(gmps_neus_all, features = "preNeu_score")
VlnPlot(gmps_neus_all, features = "immNeu_score")
VlnPlot(gmps_neus_all, features = "matNeu_score")

fine_neu_labels <- c(
  "5" = "GMPs",
  "0" = "proNeu",
  "6" = "proNeu",
  "2" = "proNeu2",
  "4" = "preNeu",
  "1" = "immature",
  "3" = "mature",
  "7" = "mature"
)

# Assuming your existing fine labels are in a metadata column, e.g. seurat_obj$immgen_fine
gmps_neus_all$fine_neu_labels <- unname(
  fine_neu_labels[as.character(gmps_neus_all$seurat_clusters)]
)

DimPlot(gmps_neus_all, reduction = "umap", group.by = "sex")
```

### Trajectory analysis

```{r}
library(monocle3)
library(SeuratWrappers)

find_traj_genes <- function(obj){
  orig_umap <- reducedDims(obj)$UMAP
  obj <- choose_graph_segments(obj)
  reducedDims(obj)$UMAP <- orig_umap[colnames(obj), ]
  obj <- cluster_cells(obj, reduction_method = "UMAP")
  obj <- learn_graph(obj, close_loop = FALSE, use_partition = FALSE)
  obj <- order_cells(obj, reduction_method = "UMAP")
  return(obj)
}
```

```{r}
compute_pseudotime_spearman <- function(cds) {
  
  pseudotime <- cds@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]
  
  expr_matrix <- exprs(cds)  # genes x cells, sparse matrix
  expr_matrix <- expr_matrix[, names(pseudotime)]  # align cell order
  
  # Remove cells with NA/Inf pseudotime (common at root-adjacent or disconnected cells)
  valid_cells <- is.finite(pseudotime)
  expr_matrix <- expr_matrix[, valid_cells]
  pt <- pseudotime[valid_cells]
  
  # Spearman correlation per gene (loop version, safest for sparse matrices)
  spearman_results <- apply(expr_matrix, 1, function(x) {
    test <- cor.test(as.numeric(x), pt, method = "spearman")
    c(rho = test$estimate, p_value = test$p.value)
  })
  
  spearman_df <- as.data.frame(t(spearman_results))
  colnames(spearman_df) <- c("rho", "p_value")
  spearman_df$gene <- rownames(spearman_df)
  spearman_df$padj <- p.adjust(spearman_df$p_value, method = "BH")
  
  # Sort by strongest correlation
  spearman_df <- spearman_df[order(-abs(spearman_df$rho)), ]
  
  return(spearman_df)
}
```

#### Male 

```{r}
neustem_3m_male_bm <- subset(gmps_neus_all, age == "3m" & sex == "male")
DimPlot(neustem_3m_male_bm, reduction = "umap", group.by = "age")

neustem_3m_male_cds <- as.cell_data_set(neustem_3m_male_bm, assay = "RNA")
neustem_3m_male_cds <- preprocess_cds(neustem_3m_male_cds, num_dim = 50)
neustem_3m_male_cds <- cluster_cells(neustem_3m_male_cds)

neustem_3m_male_cds <- learn_graph(
  neustem_3m_male_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    prune_graph = TRUE,
    minimal_branch_len = 10,
    ncenter = 300),
  verbose = FALSE)

neustem_3m_male_cds <- order_cells(neustem_3m_male_cds, reduction_method = "UMAP")
neustem_3m_male_cds <- find_traj_genes(neustem_3m_male_cds)

neustem_3m_male_genes <- graph_test(neustem_3m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_3m_male_genes <- neustem_3m_male_genes[rev(order(neustem_3m_male_genes$morans_test_statistic )), ]

spearman_3m  <- compute_pseudotime_spearman(neustem_3m_male_cds)
write.csv(spearman_3m, "male_bm_3m_spearman.csv")
```

```{r}
neustem_12m_male_bm <- subset(gmps_neus_all, age == "12m" & sex == "male")
DimPlot(neustem_12m_male_bm, reduction = "umap", group.by = "age")

neustem_12m_male_cds <- as.cell_data_set(neustem_12m_male_bm, assay = "RNA")
neustem_12m_male_cds <- preprocess_cds(neustem_12m_male_cds, num_dim = 50)
neustem_12m_male_cds <- cluster_cells(neustem_12m_male_cds)

neustem_12m_male_cds <- learn_graph(
  neustem_12m_male_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    prune_graph = TRUE,
    minimal_branch_len = 10),
  verbose = FALSE)

neustem_12m_male_cds <- order_cells(neustem_12m_male_cds, reduction_method = "UMAP")
neustem_12m_male_cds <- find_traj_genes(neustem_12m_male_cds)

neustem_12m_male_genes <- graph_test(neustem_12m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_12m_male_genes <- neustem_12m_male_genes[rev(order(neustem_12m_male_genes$morans_test_statistic )), ]

spearman_12m  <- compute_pseudotime_spearman(neustem_12m_male_cds)
write.csv(spearman_12m, "male_bm_12m_spearman.csv")
```

```{r}
neustem_18m_male_bm <- subset(gmps_neus_all, age == "18m" & sex == "male")
DimPlot(neustem_18m_male_bm, reduction = "umap", group.by = "age")

neustem_18m_male_cds <- as.cell_data_set(neustem_18m_male_bm, assay = "RNA")
neustem_18m_male_cds <- preprocess_cds(neustem_18m_male_cds, num_dim = 50)
neustem_18m_male_cds <- cluster_cells(neustem_18m_male_cds)

neustem_18m_male_cds <- learn_graph(
  neustem_18m_male_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    prune_graph = TRUE,
    minimal_branch_len = 10),
  verbose = FALSE)

neustem_18m_male_cds <- order_cells(neustem_18m_male_cds, reduction_method = "UMAP")
neustem_18m_male_cds <- find_traj_genes(neustem_18m_male_cds)

neustem_18m_male_genes <- graph_test(neustem_18m_male_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_18m_male_genes <- neustem_18m_male_genes[rev(order(neustem_18m_male_genes$morans_test_statistic )), ]

print(neustem_18m_male_genes)

intersect(rownames(neustem_3m_male_genes), rownames(neustem_18m_male_genes))

spearman_18m <- compute_pseudotime_spearman(neustem_18m_male_cds)
write.csv(spearman_18m, "male_bm_18m_spearman.csv")
```

```{r}
print(neustem_3m_male_genes)
print(neustem_12m_male_genes)
print(neustem_18m_male_genes)

# Genes unique to each timepoint (not found in either of the other two lists)
unique_3m  <- setdiff(rownames(neustem_3m_male_genes),  
                      union(rownames(neustem_12m_male_genes), rownames(neustem_18m_male_genes)))

unique_12m  <- setdiff(rownames(neustem_12m_male_genes),  
                      union(rownames(neustem_3m_male_genes), rownames(neustem_18m_male_genes)))

unique_18m  <- setdiff(rownames(neustem_18m_male_genes),  
                      union(rownames(neustem_12m_male_genes), rownames(neustem_3m_male_genes)))

write.csv(neustem_3m_male_genes, "male_trajectory_3m_genes.csv")
write.csv(neustem_12m_male_genes, "male_trajectory_12m_genes.csv")
write.csv(neustem_18m_male_genes, "male_trajectory_18m_genes.csv")
```

#### Female

```{r}
neustem_3m_female_bm <- subset(gmps_neus_all, age == "3m" & sex == "female")
DimPlot(neustem_3m_female_bm, reduction = "umap", group.by = "age")

neustem_3m_female_cds <- as.cell_data_set(neustem_3m_female_bm, assay = "RNA")
neustem_3m_female_cds <- preprocess_cds(neustem_3m_female_cds, num_dim = 50)
neustem_3m_female_cds <- cluster_cells(neustem_3m_female_cds)

neustem_3m_female_cds <- learn_graph(
  neustem_3m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    prune_graph = TRUE,
    minimal_branch_len = 10),
  verbose = FALSE)

neustem_3m_female_cds <- order_cells(neustem_3m_female_cds, reduction_method = "UMAP")
neustem_3m_female_cds <- find_traj_genes(neustem_3m_female_cds)

neustem_3m_female_genes <- graph_test(neustem_3m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_3m_female_genes <- neustem_3m_female_genes[rev(order(neustem_3m_female_genes$morans_test_statistic )), ]

spearman_3m  <- compute_pseudotime_spearman(neustem_3m_female_cds)
write.csv(spearman_3m, "female_bm_3m_spearman.csv")
```

```{r}
neustem_12m_female_bm <- subset(gmps_neus_all, age == "12m" & sex == "female")
DimPlot(neustem_12m_female_bm, reduction = "umap", group.by = "age")

neustem_12m_female_cds <- as.cell_data_set(neustem_12m_female_bm, assay = "RNA")
neustem_12m_female_cds <- preprocess_cds(neustem_12m_female_cds, num_dim = 50)
neustem_12m_female_cds <- cluster_cells(neustem_12m_female_cds)

neustem_12m_female_cds <- learn_graph(
  neustem_12m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_12m_female_cds <- order_cells(neustem_12m_female_cds, reduction_method = "UMAP")
neustem_12m_female_cds <- find_traj_genes(neustem_12m_female_cds)

neustem_12m_female_genes <- graph_test(neustem_12m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_12m_female_genes <- neustem_12m_female_genes[rev(order(neustem_12m_female_genes$morans_test_statistic )), ]

spearman_12m  <- compute_pseudotime_spearman(neustem_12m_female_cds)
write.csv(spearman_12m, "female_bm_12m_spearman.csv")
```

```{r}
neustem_18m_female_bm <- subset(gmps_neus_all, age == "18m" & sex == "female")
DimPlot(neustem_18m_female_bm, reduction = "umap", group.by = "age")

neustem_18m_female_cds <- as.cell_data_set(neustem_18m_female_bm, assay = "RNA")
neustem_18m_female_cds <- preprocess_cds(neustem_18m_female_cds, num_dim = 50)
neustem_18m_female_cds <- cluster_cells(neustem_18m_female_cds)

neustem_18m_female_cds <- learn_graph(
  neustem_18m_female_cds,
  use_partition = FALSE,
  close_loop = FALSE,  
  learn_graph_control = list(
    ncenter = 300,
    prune_graph = TRUE),
  verbose = FALSE)

neustem_18m_female_cds <- order_cells(neustem_18m_female_cds, reduction_method = "UMAP")
neustem_18m_female_cds <- find_traj_genes(neustem_18m_female_cds)

neustem_18m_female_genes <- graph_test(neustem_18m_female_cds, 
                        neighbor_graph = "principal_graph", 
                        reduction_method = "UMAP", 
                        cores = 4)

neustem_18m_female_genes <- neustem_18m_female_genes[rev(order(neustem_18m_female_genes$morans_test_statistic )), ]

print(neustem_18m_female_genes)

intersect(rownames(neustem_3m_female_genes), rownames(neustem_18m_female_genes))

spearman_18m <- compute_pseudotime_spearman(neustem_18m_female_cds)
write.csv(spearman_18m, "female_bm_18m_spearman.csv")
```

```{r}
print(neustem_3m_female_genes)
print(neustem_12m_female_genes)
print(neustem_18m_female_genes)

# Genes unique to each timepoint (not found in either of the other two lists)
unique_3m  <- setdiff(rownames(neustem_3m_female_genes),  
                      union(rownames(neustem_12m_female_genes), rownames(neustem_18m_female_genes)))

unique_12m  <- setdiff(rownames(neustem_12m_female_genes),  
                      union(rownames(neustem_3m_female_genes), rownames(neustem_18m_female_genes)))

unique_18m  <- setdiff(rownames(neustem_18m_female_genes),  
                      union(rownames(neustem_12m_female_genes), rownames(neustem_3m_female_genes)))

write.csv(neustem_3m_female_genes, "female_trajectory_3m_genes.csv")
write.csv(neustem_12m_female_genes, "female_trajectory_12m_genes.csv")
write.csv(neustem_18m_female_genes, "female_trajectory_18m_genes.csv")
```

#### Combined

```{r}
gmps_neus_all <- JoinLayers(gmps_neus_all)
combined_cds <- as.cell_data_set(gmps_neus_all, assay = "RNA")
combined_cds <- preprocess_cds(combined_cds, num_dim = 50)
combined_cds <- cluster_cells(combined_cds)

combined_cds <- learn_graph(
  combined_cds,
  use_partition = FALSE,
  close_loop = FALSE,
  learn_graph_control = list(
    prune_graph = TRUE,
    ncenter = 300),
  verbose = FALSE)

combined_cds <- order_cells(combined_cds, reduction_method = "UMAP")
```

```{r}
setwd("Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis")
neustem_3m_female_genes <- read_csv("female/female_bm_3m_trajectory_moran.csv")
neustem_3m_male_genes <- read_csv("male/male_bm_3m_trajectory_moran.csv")
neustem_12m_female_genes <- read_csv("female/female_bm_12m_trajectory_moran.csv")
neustem_12m_male_genes <- read_csv("male/male_bm_12m_trajectory_moran.csv")
neustem_18m_female_genes <- read_csv("female/female_bm_18m_trajectory_moran.csv")
neustem_18m_male_genes <- read_csv("male/male_bm_18m_trajectory_moran.csv")

gene_lists <- list(
  female_3m = neustem_3m_female_genes,
  male_3m = neustem_3m_male_genes,
  female_12m = neustem_12m_female_genes,
  male_12m = neustem_12m_male_genes,
  female_18m = neustem_18m_female_genes,
  male_18m = neustem_18m_male_genes
)

sig_gene_lists <- lapply(gene_lists, function(df) {
  df %>% dplyr::filter(q_value < 0.05) %>% pull(...1)
})

all_sig_genes <- unique(unlist(sig_gene_lists))
```

```{r}
# Subset the CDS by sex
combined_cds_female <- combined_cds[, combined_cds$sex == "female"] 
combined_cds_male   <- combined_cds[, combined_cds$sex == "male"]

# Female
counts_sub_female <- counts(combined_cds_female)[all_sig_genes, ]
pseudotime_female <- pseudotime(combined_cds_female)
cellWeights_female <- rep(1, ncol(combined_cds_female))

# Male
counts_sub_male <- counts(combined_cds_male)[all_sig_genes, ]
pseudotime_male <- pseudotime(combined_cds_male)
cellWeights_male <- rep(1, ncol(combined_cds_male))
```

```{r}
female_GAM <- fitGAM(counts = counts_sub_female,
                     pseudotime = pseudotime_female,
                     cellWeights = cellWeights_female,
                     nknots = 5,
                     conditions = factor(combined_cds_female@colData@listData[["age"]]))

plotSmoothers(female_GAM, counts_sub_female, gene = "Elane")

plotSmoothers(female_GAM, counts_sub_female, gene = "Elane",
              pointCol = colData(combined_cds_female)@listData[["fine_neu_labels"]])
```

```{r}
plotSmoothers(female_GAM, counts_sub_female, gene = "Fcer1g") +
  ylim(0, 4)
plotSmoothers(male_GAM, counts_sub_male, gene = "Fcer1g") +
  ylim(0, 4)

plotSmoothers(female_GAM, counts_sub_female, gene = "Lyz1") +
  ylim(0, 4)
plotSmoothers(male_GAM, counts_sub_male, gene = "Lyz1") +
  ylim(0, 4)

plotSmoothers(female_GAM, counts_sub_female, gene = "Lyz1")

plotSmoothers(female_GAM, counts_sub_female, gene = "Plbd1")

plotSmoothers(female_GAM, counts_sub_female, gene = "Ltf")

plotSmoothers(female_GAM, counts_sub_female, gene = "Lyz2")

plotSmoothers(female_GAM, counts_sub_female, gene = "Ifitm6")

plotSmoothers(female_GAM, counts_sub_female, gene = "Lcn2")

plotSmoothers(female_GAM, counts_sub_female, gene = "S100a9")

plotSmoothers(female_GAM, counts_sub_female, gene = "Cd177")

plotSmoothers(female_GAM, counts_sub_female, gene = "Adpgk")

plotSmoothers(female_GAM, counts_sub_female, gene = "Anxa1")

plotSmoothers(female_GAM, counts_sub_female, gene = "Ngp")

plotSmoothers(female_GAM, counts_sub_female, gene = "Cybb")

plotSmoothers(female_GAM, counts_sub_female, gene = "Srgn")
```

```{r}
male_GAM <- fitGAM(counts = counts_sub_male,
                     pseudotime = pseudotime_male,
                     cellWeights = cellWeights_male,
                     nknots = 5,
                     conditions = factor(combined_cds_male@colData@listData[["age"]]))

plotSmoothers(male_GAM, counts_sub_male, gene = "Elane")
plotSmoothers(male_GAM, counts_sub_male, gene = "Elane",
              pointCol = colData(neustem_3m_female_cds)@listData[["finel_label_readable"]])
```

```{r}
saveRDS(male_GAM, "male_bm_GAM.rds")
saveRDS(female_GAM, "female_bm_GAM.rds")
```

```{r}
plotSmoothers(female_GAM, counts_sub_female, gene = "Cybb",
              pointCol = colData(combined_cds_female)@listData[["fine_neu_labels"]]) +
  ylim(0, 4)
plotSmoothers(male_GAM, counts_sub_male, gene = "Cybb",
              pointCol = colData(combined_cds_male)@listData[["fine_neu_labels"]]) +
  ylim(0, 4)

plotSmoothers(female_GAM, counts_sub_female, gene = "Camp") +
  ylim(0, 8)
plotSmoothers(male_GAM, counts_sub_male, gene = "Camp") +
  ylim(0, 8)

plotSmoothers(female_GAM, counts_sub_female, gene = "Chil3") +
  ylim(0, 8)
plotSmoothers(male_GAM, counts_sub_male, gene = "Chil3") +
  ylim(0, 8)
```

```{r}
plotSmoothers(female_GAM, counts_sub_female, gene = "Elane") +
  ylim(0, 8)
plotSmoothers(male_GAM, counts_sub_male, gene = "Elane") +
  ylim(0, 8)

plotSmoothers(female_GAM, counts_sub_female, gene = "Mpo") +
  ylim(0, 8)
plotSmoothers(male_GAM, counts_sub_male, gene = "Mpo") +
  ylim(0, 8)

plotSmoothers(female_GAM, counts_sub_female, gene = "Cebpa") +
  ylim(0, 3)
plotSmoothers(male_GAM, counts_sub_male, gene = "Cebpa") +
  ylim(0, 3)

plotSmoothers(female_GAM, counts_sub_female, gene = "Ly6g") +
  ylim(0, 6)
plotSmoothers(male_GAM, counts_sub_male, gene = "Ms.Ly.6C") +
  ylim(0, 6)
```

```{r}
plotSmoothers(male_GAM, counts_sub_male, gene = "Il22ra2") 
```


```{r}
Default

FeaturePlot(gmps_neus_all, features = "nCount_RNA")
FeaturePlot(gmps_neus_all, features = "Cebpa")
FeaturePlot(gmps_neus_all, features = "Ms.CD62L", assay = "ADT")
```

```{r}
cd <- colData(combined_cds)
cond <- droplevels(interaction(cd$age, cd$sex, sep = "_"))
```



### Heatmap 

```{r}
neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "mature" & age != "9m")
neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "mature" & age != "9m") 

neus_all <- subset(gmps_neus_all, fine_neu_labels == "mature" & age != "9m")
```

```{r}
DefaultAssay(neus_all) <- "RNA"
neus_all <- NormalizeData(neus_all)
neus_all <- FindVariableFeatures(neus_all)
neus_all <- ScaleData(neus_all, 
                                  features = VariableFeatures(neus_all))
neus_all <- RunPCA(neus_all, assay = "RNA")

ElbowPlot(neus_all)
DimPlot(neus_all, reduction = "pca")

neus_all <- RunUMAP(neus_all,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(neus_all, reduction = "umap", group.by = "sex")
```

```{r}
DefaultAssay(neus_female) <- "RNA"
neus_female <- NormalizeData(neus_female)
neus_female <- FindVariableFeatures(neus_female)
neus_female <- ScaleData(neus_female, 
                                  features = VariableFeatures(neus_female))
neus_female <- RunPCA(neus_female, assay = "RNA")

ElbowPlot(neus_female)
DimPlot(neus_female, reduction = "pca")

neus_female <- RunUMAP(neus_female,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(neus_female, reduction = "umap", group.by = "age")
```

```{r}
DefaultAssay(neus_male) <- "RNA"
neus_male <- NormalizeData(neus_male)
neus_male <- FindVariableFeatures(neus_male)
neus_male <- ScaleData(neus_male, 
                                  features = VariableFeatures(neus_male))
neus_male <- RunPCA(neus_male, assay = "RNA")

ElbowPlot(neus_male)
DimPlot(neus_male, reduction = "pca")

neus_male <- RunUMAP(neus_male,
                        assay = "RNA",
                        reduction = "pca",
                        dims = 1:15,
                        reduction.name = "umap")

DimPlot(neus_male, reduction = "umap", group.by = "age")
```


```{r}
library(matrixStats)
library(ggplot2)

avg_expr <- AverageExpression(neus_all, group.by = "age_sex",
                               assays = "RNA", slot = "data")$RNA

print(colnames(avg_expr))  # confirm the exact strings before assuming anything

gene_var  <- rowVars(as.matrix(avg_expr))
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:2000]
pca_input <- t(avg_expr[top_genes, ])
pca <- prcomp(pca_input, scale. = TRUE, center = TRUE)

pca_df <- as.data.frame(pca$x[, 1:2])
pca_df$sample <- rownames(pca_df)

split_labels <- strsplit(pca_df$sample, "-")
pca_df$age <- sapply(split_labels, `[`, 1)
pca_df$sex <- sapply(split_labels, `[`, 2)

print(unique(pca_df$age))  # should now show g12m, g18m, g3m
print(unique(pca_df$sex))  # should now show F, M

pca_df$sex <- factor(pca_df$sex, levels = c("F", "M"), labels = c("female", "male"))
pca_df$age <- factor(pca_df$age, levels = c("g3m", "g12m", "g18m"))

var_explained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, group = sex)) +
  geom_point(size = 4) +
  geom_text(aes(label = age), vjust = -1, size = 3.5, show.legend = FALSE) +
  labs(x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"),
       color = "Sex") +
  theme_minimal(base_size = 13)
```

```{r}
avg_expr <- AverageExpression(gmps_neus_all, 
                              group.by = "age_sex",
                              assays = "RNA", 
                              slot = "data")$RNA

print(colnames(avg_expr))  # confirm the exact strings before assuming anything

gene_var  <- rowVars(as.matrix(avg_expr))
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:2000]
pca_input <- t(avg_expr[top_genes, ])
pca <- prcomp(pca_input, scale. = TRUE, center = TRUE)

pca_df <- as.data.frame(pca$x[, 1:2])
pca_df$sample <- rownames(pca_df)

split_labels <- strsplit(pca_df$sample, "-")
pca_df$age <- sapply(split_labels, `[`, 1)
pca_df$sex <- sapply(split_labels, `[`, 2)

print(unique(pca_df$age))  # should now show g12m, g18m, g3m
print(unique(pca_df$sex))  # should now show F, M

pca_df$sex <- factor(pca_df$sex, levels = c("F", "M"), labels = c("female", "male"))
pca_df$age <- factor(pca_df$age, levels = c("g3m", "g9m", "g12m", "g18m"))

var_explained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, group = sex)) +
  geom_point(size = 4) +
  geom_text(aes(label = age), vjust = -1, size = 3.5, show.legend = FALSE) +
  labs(x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"),
       color = "Sex") +
  theme_minimal(base_size = 13)
```

```{r}
avg_expr <- AverageExpression(subset(gmps_neus_all, age != "9m"), 
                              group.by = "age_sex",
                              assays = "RNA", 
                              slot = "data")$RNA

print(colnames(avg_expr))  # confirm the exact strings before assuming anything

gene_var  <- rowVars(as.matrix(avg_expr))
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:2000]
pca_input <- t(avg_expr[top_genes, ])
pca <- prcomp(pca_input, scale. = TRUE, center = TRUE)

pca_df <- as.data.frame(pca$x[, 1:2])
pca_df$sample <- rownames(pca_df)

split_labels <- strsplit(pca_df$sample, "-")
pca_df$age <- sapply(split_labels, `[`, 1)
pca_df$sex <- sapply(split_labels, `[`, 2)

print(unique(pca_df$age))  # should now show g12m, g18m, g3m
print(unique(pca_df$sex))  # should now show F, M

pca_df$sex <- factor(pca_df$sex, levels = c("F", "M"), labels = c("female", "male"))
pca_df$age <- factor(pca_df$age, levels = c("g3m", "g12m", "g18m"))

var_explained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, group = sex)) +
  geom_point(size = 4) +
  geom_text(aes(label = age), vjust = -1, size = 3.5, show.legend = FALSE) +
  labs(x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"),
       color = "Sex") +
  theme_minimal(base_size = 13)
```


```{r}
avg_expr <- AverageExpression(subset(gmps_neus_all, age != "9m" & fine_neu_labels == "proNeu"), 
                              group.by = "age_sex",
                              assays = "RNA", 
                              slot = "data")$RNA

print(colnames(avg_expr))  # confirm the exact strings before assuming anything

gene_var  <- rowVars(as.matrix(avg_expr))
top_genes <- names(sort(gene_var, decreasing = TRUE))[1:2000]
pca_input <- t(avg_expr[top_genes, ])
pca <- prcomp(pca_input, scale. = TRUE, center = TRUE)

pca_df <- as.data.frame(pca$x[, 1:2])
pca_df$sample <- rownames(pca_df)

split_labels <- strsplit(pca_df$sample, "-")
pca_df$age <- sapply(split_labels, `[`, 1)
pca_df$sex <- sapply(split_labels, `[`, 2)

print(unique(pca_df$age))  # should now show g12m, g18m, g3m
print(unique(pca_df$sex))  # should now show F, M

pca_df$sex <- factor(pca_df$sex, levels = c("F", "M"), labels = c("female", "male"))
pca_df$age <- factor(pca_df$age, levels = c("g3m","g9m", "g12m", "g18m"))

var_explained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, group = sex)) +
  geom_point(size = 4) +
  geom_text(aes(label = age), vjust = -1, size = 3.5, show.legend = FALSE) +
  labs(x = paste0("PC1 (", var_explained[1], "%)"),
       y = paste0("PC2 (", var_explained[2], "%)"),
       color = "Sex") +
  theme_minimal(base_size = 13)
```

```{r}
gmps_neus_all$fine_neu_labels <- factor(gmps_neus_all$fine_neu_labels, 
                                        levels = c("GMPs", "proNeu", "proNeu2", "preNeu", "immature", "mature"))

plot_pca_by_label <- function(seurat_obj, label, n_top_genes = 2000) {

  neu_subset <- subset(seurat_obj, age != "9m" & fine_neu_labels == label)

  avg_expr <- AverageExpression(neu_subset, group.by = "age_sex",
                                 assays = "RNA", slot = "data")$RNA

  # guard: need at least 3 samples for a PCA to mean anything at all,
  # and skip silently-incomplete groups rather than erroring the whole loop
  if (ncol(avg_expr) < 3) {
    message("Skipping '", label, "': only ", ncol(avg_expr), " age_sex groups present")
    return(NULL)
  }

  gene_var  <- rowVars(as.matrix(avg_expr))
  n_use     <- min(n_top_genes, sum(gene_var > 0))  # guard against too few variable genes
  top_genes <- names(sort(gene_var, decreasing = TRUE))[1:n_use]

  pca_input <- t(avg_expr[top_genes, ])
  pca <- prcomp(pca_input, scale. = TRUE, center = TRUE)

  pca_df <- as.data.frame(pca$x[, 1:2])
  pca_df$sample <- rownames(pca_df)

  split_labels <- strsplit(pca_df$sample, "-")
  pca_df$age <- sapply(split_labels, `[`, 1)
  pca_df$sex <- sapply(split_labels, `[`, 2)

  pca_df$sex <- factor(pca_df$sex, levels = c("F", "M"), labels = c("female", "male"))
  pca_df$age <- factor(pca_df$age, levels = c("g3m", "g12m", "g18m"))

  var_explained <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

  p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = sex, group = sex)) +
    geom_point(size = 4) +
    geom_text(aes(label = age), vjust = -1, size = 3.5, show.legend = FALSE) +
    labs(title = label,
         x = paste0("PC1 (", var_explained[1], "%)"),
         y = paste0("PC2 (", var_explained[2], "%)"),
         color = "Sex") +
    theme_minimal(base_size = 13)

  return(p)
}

# run across every fine_neu_labels value
all_labels <- unique(gmps_neus_all$fine_neu_labels)
pca_plots <- lapply(all_labels, function(lbl) plot_pca_by_label(gmps_neus_all, lbl))
names(pca_plots) <- all_labels
pca_plots <- Filter(Negate(is.null), pca_plots)  # drop the skipped ones

# view them
pca_plots[["proNeu"]]

# or all at once
library(patchwork)

custom_order <- c("GMPs", "proNeu", "proNeu2", "preNeu", "immature", "mature")  # your desired order — must match names(pca_plots) exactly
pca_plots_ordered <- pca_plots[custom_order]
wrap_plots(pca_plots_ordered)
```


```{r}
Idents(neus_female) <- neus_female$age
Idents(neus_male) <- neus_male$age

markers <- FindMarkers(neus_female, 
                       ident.1 = "18m", 
                       ident.2 = "3m",
                       logfc.threshold = 0, 
                       min.pct = 0.1)

VlnPlot(neus_female, features = "Cdc37l1")

markers <- FindMarkers(neus_male, 
                       ident.1 = "18m", 
                       ident.2 = "3m",
                       logfc.threshold = 0, 
                       min.pct = 0.1)

VlnPlot(neus_male, features = "Atp5k")
```


```{r}
neus_female$age <- factor(neus_female$age, levels = c("3m", "9m", "12m", "18m"))
avg_expr <- AverageExpression(neus_female, group.by = "age", assays = "RNA", slot = "data")$RNA

avg_expr <- avg_expr[apply(avg_expr, 1, max) > 0.5, ]  # drop near-zero genes
fc_range <- apply(avg_expr, 1, function(x) max(x) - min(x))
changing_genes <- avg_expr[order(-fc_range), ][1:1000, ]  # top movers, adjust N as needed

z_scored <- t(scale(t(changing_genes)))  # z-score per gene across timepoints

# hierarchical clustering into pattern groups
library(pheatmap)
clusters_female <- pheatmap(z_scored, 
                     cluster_cols = FALSE, 
                     cutree_rows = 7,
                     show_rownames = TRUE)

row_clusters <- cutree(clusters_female$tree_row, k = 7)

# Get the order rows are actually plotted in (top to bottom)
plot_order <- clusters_female$tree_row$order

# row_clusters in that plotted order
ordered_clusters <- row_clusters[plot_order]

# Relabel so cluster numbers increase top-to-bottom as displayed
cluster_levels_in_order <- unique(ordered_clusters)
relabel_map <- setNames(seq_along(cluster_levels_in_order), cluster_levels_in_order)

visual_clusters <- relabel_map[as.character(row_clusters)]
names(visual_clusters) <- names(row_clusters)

new_names <- c("1" = "U_shaped_aged_low", 
               "2" = "U_shaped_aged_high", 
               "3" = "Increase_middle_high", 
               "4" = "Increase_aged_high", 
               "5" = "Reverse_U_shaped", 
               "6" = "Sharp_aged_decrease",
               "7" = "Gradual_aged_decrease"
               )

# Apply to your (visually-ordered) cluster vector
renamed_clusters <- factor(new_names[as.character(visual_clusters)],
                            levels = new_names)
```

```{r}
neus_female$age <- factor(neus_female$age, levels = c("3m", "9m", "12m", "18m"))
avg_expr <- AverageExpression(neus_female, group.by = "age", assays = "RNA", slot = "data")$RNA
avg_expr <- avg_expr[apply(avg_expr, 1, max) > 0.5, ]  # drop near-zero genes
fc_range <- apply(avg_expr, 1, function(x) max(x) - min(x))
changing_genes <- avg_expr[order(-fc_range), ][1:1000, ]  # top movers, adjust N as needed
z_scored <- t(scale(t(changing_genes)))  # z-score per gene across timepoints

# hierarchical clustering into pattern groups
library(pheatmap)
clusters_female <- pheatmap(z_scored, 
                     cluster_cols = FALSE, 
                     cutree_rows = 7,
                     show_rownames = TRUE)

row_clusters <- cutree(clusters_female$tree_row, k = 7)

# Get the order rows are actually plotted in (top to bottom)
plot_order <- clusters_female$tree_row$order

# row_clusters in that plotted order
ordered_clusters <- row_clusters[plot_order]

# Relabel so cluster numbers increase top-to-bottom as displayed
cluster_levels_in_order <- unique(ordered_clusters)
relabel_map <- setNames(seq_along(cluster_levels_in_order), cluster_levels_in_order)
visual_clusters <- relabel_map[as.character(row_clusters)]
names(visual_clusters) <- names(row_clusters)

new_names <- c("1" = "U_shaped_aged_low", 
               "2" = "U_shaped_aged_high", 
               "3" = "Increase_middle_high", 
               "4" = "Increase_aged_high", 
               "5" = "Reverse_U_shaped", 
               "6" = "Sharp_aged_decrease",
               "7" = "Gradual_aged_decrease"
               )

# Apply to your (visually-ordered) cluster vector
renamed_clusters <- factor(new_names[as.character(visual_clusters)],
                            levels = new_names)

# Build a data frame with gene names and their cluster assignment
gene_cluster_df <- data.frame(
  gene = names(visual_clusters),
  cluster_number = visual_clusters,
  cluster_name = renamed_clusters,
  row.names = NULL
)

head(gene_cluster_df)

write.csv(gene_cluster_df, "female_neutrophil_age_gene_changes.csv")
```

### Rho trends

```{r}
age_trend_clusters <- function(obj,
                               age_col     = "age",
                               age_levels  = c("3m", "12m", "18m"),
                               rho_cutoff  = 0.1,
                               n_clusters  = 6,
                               tol         = 0.5,   # band for high/middle/low
                               assay       = "RNA",
                               plot        = TRUE,
                               out_csv     = NULL) {

  ## --- 1. Spearman trend per gene -------------------------------------
  age_char <- as.character(obj[[age_col]][, 1])
  if (!all(age_char %in% age_levels))
    stop("ages present that are not in age_levels: ",
         paste(setdiff(unique(age_char), age_levels), collapse = ", "))

  age_numeric <- as.numeric(factor(age_char, levels = age_levels))
  expr_mat <- GetAssayData(obj, assay = assay, layer = "data")

  trend_results <- apply(expr_mat, 1, function(x) {
    ct <- suppressWarnings(cor.test(x, age_numeric, method = "spearman",
                                    exact = FALSE))
    c(rho = unname(ct$estimate), pval = ct$p.value)
  })
  trend_df <- as.data.frame(t(trend_results))
  trend_df$padj <- p.adjust(trend_df$pval, method = "BH")

  changing_genes <- trend_df %>%
    dplyr::filter(abs(rho) > rho_cutoff) %>%
    dplyr::arrange(padj)

  message(sprintf("  %d of %d genes pass |rho| > %.2f",
                  nrow(changing_genes), nrow(trend_df), rho_cutoff))

  if (nrow(changing_genes) < n_clusters) {
    warning("fewer changing genes than n_clusters -- returning trend_df only")
    return(list(trend_df = trend_df, changing_genes = changing_genes,
               gene_cluster_df = NULL))
  }

  ## --- 2. Average expression per age, z-scored ------------------------
  avg_expr <- AverageExpression(obj, group.by = age_col,
                                assays = assay, layer = "data")[[assay]]

  ## AverageExpression prefixes numeric-leading group names with "g" (e.g.
  ## "3m" -> "g3m"). Match on the trailing age string rather than assuming
  ## the prefix, so this works whether or not Seurat mangles the name.
  col_age <- sub("^g", "", colnames(avg_expr))
  age_order <- match(age_levels, col_age)
  if (any(is.na(age_order)))
    stop("could not match AverageExpression columns to age_levels; ",
         "columns were: ", paste(colnames(avg_expr), collapse = ", "))
  avg_expr <- avg_expr[, age_order, drop = FALSE]
  colnames(avg_expr) <- age_levels

  avg_expr <- avg_expr[rownames(changing_genes), , drop = FALSE]
  z_scored <- t(scale(t(avg_expr)))
  z_scored <- z_scored[complete.cases(z_scored), , drop = FALSE]  # drop zero-var rows

  ## --- 3. Hierarchical clustering --------------------------------------
  hm <- pheatmap(z_scored, cluster_cols = FALSE, cutree_rows = n_clusters,
                show_rownames = TRUE, silent = !plot)
  row_clusters <- cutree(hm$tree_row, k = n_clusters)

  ## --- 4. Name each cluster from its OWN centroid ----------------------
  name_shape <- function(v, tol) {
    lab <- ifelse(v >  tol, "high", ifelse(v < -tol, "low", "middle"))
    paste(lab, collapse = "_")
  }

  centroids <- sapply(sort(unique(row_clusters)), function(k) {
    colMeans(z_scored[row_clusters == k, age_levels, drop = FALSE])
  })
  colnames(centroids) <- sort(unique(row_clusters))
  shape_by_cluster <- apply(centroids, 2, name_shape, tol = tol)

  if (any(duplicated(shape_by_cluster)))
    warning("two clusters got the same name (", 
            paste(shape_by_cluster[duplicated(shape_by_cluster) |
                                   duplicated(shape_by_cluster, fromLast = TRUE)],
                  collapse = ", "),
            ") -- consider a smaller tol or fewer clusters")

  cluster_name <- factor(shape_by_cluster[as.character(row_clusters)],
                         levels = unique(shape_by_cluster))

  gene_cluster_df <- data.frame(
    gene            = names(row_clusters),
    cluster_number  = row_clusters,
    cluster_name    = cluster_name,
    row.names       = NULL
  )

  list(trend_df = trend_df,
       changing_genes = changing_genes,
       z_scored = z_scored,
       heatmap = hm,
       centroids = centroids,
       gene_cluster_df = gene_cluster_df)
}
```

#### Mature neutrophils

```{r}
mat_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "mature" & age != "9m")
mat_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "mature" & age != "9m") 

DefaultAssay(mat_neus_female) <- "RNA"
mat_neus_female <- NormalizeData(mat_neus_female)
mat_neus_female <- FindVariableFeatures(mat_neus_female)
mat_neus_female <- ScaleData(mat_neus_female, 
                                  features = VariableFeatures(mat_neus_female))

DefaultAssay(mat_neus_male) <- "RNA"
mat_neus_male <- NormalizeData(mat_neus_male)
mat_neus_male <- FindVariableFeatures(mat_neus_male)
mat_neus_male <- ScaleData(mat_neus_male, 
                                  features = VariableFeatures(mat_neus_male))
```

```{r}
res_female_mature <- age_trend_clusters(mat_neus_female)

res_female_mature$changing_genes$gene <- rownames(res_female_mature$changing_genes)

mat_female_full <- merge(
  res_female_mature$gene_cluster_df,
  res_female_mature$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(mat_female_full, "mat_female_changing_genes.csv")
```

```{r}
res_male_mature <- age_trend_clusters(mat_neus_male)

res_male_mature$changing_genes$gene <- rownames(res_male_mature$changing_genes)

mat_male_full <- merge(
  res_male_mature$gene_cluster_df,
  res_male_mature$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(mat_male_full, "mat_male_changing_genes.csv")
```

```{r}
age_trend_clusters <- function(obj,
                               age_col     = "age",
                               age_levels  = c("3m", "12m", "18m"),
                               rho_cutoff  = 0.1,
                               n_clusters  = 6,
                               tol         = 0.5,
                               assay       = "RNA",
                               plot        = TRUE,
                               out_csv     = NULL,
                               do_cluster  = TRUE,     # NEW: FALSE = ranking only
                               min_p       = 1e-300) { # NEW: floor for pval before -log10

  suppressPackageStartupMessages({ library(Seurat); library(dplyr) })

  ## --- 1. Spearman trend per gene (ALL genes) ---
  age_char <- as.character(obj[[age_col]][, 1])
  if (!all(age_char %in% age_levels))
    stop("ages present that are not in age_levels: ",
         paste(setdiff(unique(age_char), age_levels), collapse = ", "))
  age_numeric <- as.numeric(factor(age_char, levels = age_levels))
  expr_mat <- GetAssayData(obj, assay = assay, layer = "data")

  trend_results <- apply(expr_mat, 1, function(x) {
    sx <- suppressWarnings(sd(as.numeric(x)))
    if (is.na(sx) || !is.finite(sx) || sx == 0)
      return(c(rho = NA_real_, pval = NA_real_))
    ct <- suppressWarnings(cor.test(as.numeric(x), age_numeric,
                                    method = "spearman", exact = FALSE))
    c(rho = unname(ct$estimate), pval = ct$p.value)
  })
  trend_df <- as.data.frame(t(trend_results))
  trend_df$gene <- rownames(trend_df)
  trend_df$padj <- p.adjust(trend_df$pval, method = "BH")

  ## --- 2. GSEA ranking metric ---
  p_floor <- ifelse(is.na(trend_df$pval), NA_real_, pmax(trend_df$pval, min_p))
  trend_df$rank_score <- sign(trend_df$rho) * (-log10(p_floor))
  trend_df$rank_score[is.na(trend_df$rank_score)] <- 0

  ord <- order(-trend_df$rank_score,
               -ifelse(is.na(trend_df$rho), 0, trend_df$rho),
               trend_df$gene)
  trend_df$rank_order <- NA_integer_
  trend_df$rank_order[ord] <- seq_len(nrow(trend_df))

  gsea_rank <- trend_df$rank_score[ord]
  names(gsea_rank) <- trend_df$gene[ord]
  gsea_rank <- gsea_rank[!duplicated(names(gsea_rank))]
  message(sprintf(" genome-wide ranking: %d genes", length(gsea_rank)))

  ## --- 3. Filtered changing genes (for clustering) ---
  changing_genes <- trend_df %>%
    dplyr::filter(!is.na(rho), abs(rho) > rho_cutoff) %>%
    dplyr::arrange(padj)
  message(sprintf(" %d of %d genes pass |rho| > %.2f",
                  nrow(changing_genes), nrow(trend_df), rho_cutoff))
  if (!is.null(out_csv)) { write.csv(trend_df, out_csv, row.names = FALSE) }

  if (!do_cluster)
    return(list(gsea_rank = gsea_rank, trend_df = trend_df,
                changing_genes = changing_genes, gene_cluster_df = NULL,
                z_scored = NULL, heatmap = NULL, centroids = NULL))

  if (nrow(changing_genes) < n_clusters) {
    warning("fewer changing genes than n_clusters -- returning ranking + trend_df only")
    return(list(gsea_rank = gsea_rank, trend_df = trend_df,
                changing_genes = changing_genes, gene_cluster_df = NULL,
                z_scored = NULL, heatmap = NULL, centroids = NULL))
  }

  ## --- 4. Average expression per age, z-scored ---
  avg_expr <- AverageExpression(obj, group.by = age_col,
                                assays = assay, layer = "data")[[assay]]
  col_age <- sub("^g", "", colnames(avg_expr))
  age_order <- match(age_levels, col_age)
  if (any(is.na(age_order)))
    stop("could not match AverageExpression columns to age_levels; columns were: ",
         paste(colnames(avg_expr), collapse = ", "))
  avg_expr <- avg_expr[, age_order, drop = FALSE]
  colnames(avg_expr) <- age_levels
  avg_expr <- avg_expr[changing_genes$gene, , drop = FALSE]
  z_scored <- t(scale(t(avg_expr)))
  z_scored <- z_scored[complete.cases(z_scored), , drop = FALSE]

  ## --- 5. Hierarchical clustering ---
  hm <- pheatmap::pheatmap(z_scored, cluster_cols = FALSE, cutree_rows = n_clusters,
                           show_rownames = TRUE, silent = !plot)
  row_clusters <- cutree(hm$tree_row, k = n_clusters)

  ## --- 6. Name each cluster from its OWN centroid ---
  name_shape <- function(v, tol) {
    lab <- ifelse(v > tol, "high", ifelse(v < -tol, "low", "middle"))
    paste(lab, collapse = "_")
  }
  centroids <- sapply(sort(unique(row_clusters)), function(k)
    colMeans(z_scored[row_clusters == k, age_levels, drop = FALSE]))
  colnames(centroids) <- sort(unique(row_clusters))
  shape_by_cluster <- apply(centroids, 2, name_shape, tol = tol)
  if (any(duplicated(shape_by_cluster)))
    warning("two clusters got the same name -- consider a smaller tol or fewer clusters")
  cluster_name <- factor(shape_by_cluster[as.character(row_clusters)],
                         levels = unique(shape_by_cluster))
  gene_cluster_df <- data.frame(gene = names(row_clusters),
                                cluster_number = row_clusters,
                                cluster_name = cluster_name, row.names = NULL)

  list(gsea_rank = gsea_rank, trend_df = trend_df,
       changing_genes = changing_genes, z_scored = z_scored,
       heatmap = hm, centroids = centroids, gene_cluster_df = gene_cluster_df)
}
```

```{r}
library(fgsea); library(msigdbr)

# one Seurat object per sex (mature neutrophils), run ranking only
res_f <- age_trend_clusters(mat_neus_female, do_cluster = FALSE)
res_m <- age_trend_clusters(mat_neus_male,   do_cluster = FALSE)

# GO:BP gene sets (mouse), restricted to the expressed background
bg <- union(names(res_f$gsea_rank), names(res_m$gsea_rank))
ms <- msigdbr(species = "mouse", db_species = "MM", collection = "M5", subcollection = "GO:BP")
pathways <- split(ms$gene_symbol, ms$gs_name)
pathways <- lapply(pathways, intersect, bg)
pathways <- pathways[lengths(pathways) >= 10 & lengths(pathways) <= 500]

set.seed(42)
fg_f <- fgsea(pathways, res_f$gsea_rank, minSize = 10, maxSize = 500)
fg_m <- fgsea(pathways, res_m$gsea_rank, minSize = 10, maxSize = 500)

write_csv(fg_f, "fg_f.csv")
write_csv(fg_m, "fg_m.csv")
```



#### Immature neutrophils

```{r}
immat_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "immature" & age != "9m")
immat_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "immature" & age != "9m") 

DefaultAssay(immat_neus_female) <- "RNA"
immat_neus_female <- NormalizeData(immat_neus_female)
immat_neus_female <- FindVariableFeatures(immat_neus_female)
immat_neus_female <- ScaleData(immat_neus_female, 
                                  features = VariableFeatures(immat_neus_female))

DefaultAssay(immat_neus_male) <- "RNA"
immat_neus_male <- NormalizeData(immat_neus_male)
immat_neus_male <- FindVariableFeatures(immat_neus_male)
immat_neus_male <- ScaleData(immat_neus_male, 
                                  features = VariableFeatures(immat_neus_male))
```

```{r}
res_female_immature <- age_trend_clusters(immat_neus_female)

res_female_immature$changing_genes$gene <- rownames(res_female_immature$changing_genes)

immat_female_full <- merge(
  res_female_immature$gene_cluster_df,
  res_female_immature$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(immat_female_full, "immat_female_changing_genes.csv")
```

```{r}
res_male_immature <- age_trend_clusters(immat_neus_male)

res_male_immature$changing_genes$gene <- rownames(res_male_immature$changing_genes)

immat_male_full <- merge(
  res_male_immature$gene_cluster_df,
  res_male_immature$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(immat_male_full, "immat_male_changing_genes.csv")
```

#### preNeu neutrophils

```{r}
preneu_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "preNeu" & age != "9m")
preneu_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "preNeu" & age != "9m") 

DefaultAssay(preneu_neus_female) <- "RNA"
preneu_neus_female <- NormalizeData(preneu_neus_female)
preneu_neus_female <- FindVariableFeatures(preneu_neus_female)
preneu_neus_female <- ScaleData(preneu_neus_female, 
                                  features = VariableFeatures(preneu_neus_female))

DefaultAssay(preneu_neus_male) <- "RNA"
preneu_neus_male <- NormalizeData(preneu_neus_male)
preneu_neus_male <- FindVariableFeatures(preneu_neus_male)
preneu_neus_male <- ScaleData(preneu_neus_male, 
                                  features = VariableFeatures(preneu_neus_male))
```

```{r}
res_female_preneuure <- age_trend_clusters(preneu_neus_female)

res_female_preneuure$changing_genes$gene <- rownames(res_female_preneuure$changing_genes)

preneu_female_full <- merge(
  res_female_preneuure$gene_cluster_df,
  res_female_preneuure$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(preneu_female_full, "preneu_female_changing_genes.csv")
```
#### proNeu2 neutrophils

```{r}
proneu2_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "proNeu2" & age != "9m")
proneu2_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "proNeu2" & age != "9m") 

DefaultAssay(proneu2_neus_female) <- "RNA"
proneu2_neus_female <- NormalizeData(proneu2_neus_female)
proneu2_neus_female <- FindVariableFeatures(proneu2_neus_female)
proneu2_neus_female <- ScaleData(proneu2_neus_female, 
                                  features = VariableFeatures(proneu2_neus_female))

DefaultAssay(proneu2_neus_male) <- "RNA"
proneu2_neus_male <- NormalizeData(proneu2_neus_male)
proneu2_neus_male <- FindVariableFeatures(proneu2_neus_male)
proneu2_neus_male <- ScaleData(proneu2_neus_male, 
                                  features = VariableFeatures(proneu2_neus_male))
```

```{r}
res_female_proneu2ure <- age_trend_clusters(proneu2_neus_female)

res_female_proneu2ure$changing_genes$gene <- rownames(res_female_proneu2ure$changing_genes)

proneu2_female_full <- merge(
  res_female_proneu2ure$gene_cluster_df,
  res_female_proneu2ure$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(proneu2_female_full, "proneu2_female_changing_genes.csv")
```

```{r}
res_male_proneu2 <- age_trend_clusters(proneu2_neus_male)

res_male_proneu2$changing_genes$gene <- rownames(res_male_proneu2$changing_genes)

proneu2_male_full <- merge(
  res_male_proneu2$gene_cluster_df,
  res_male_proneu2$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(proneu2_male_full, "proneu2_male_changing_genes.csv")
```

#### proNeu neutrophils

```{r}
proneu_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "proNeu" & age != "9m")
proneu_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "proNeu" & age != "9m") 

DefaultAssay(proneu_neus_female) <- "RNA"
proneu_neus_female <- NormalizeData(proneu_neus_female)
proneu_neus_female <- FindVariableFeatures(proneu_neus_female)
proneu_neus_female <- ScaleData(proneu_neus_female, 
                                  features = VariableFeatures(proneu_neus_female))

DefaultAssay(proneu_neus_male) <- "RNA"
proneu_neus_male <- NormalizeData(proneu_neus_male)
proneu_neus_male <- FindVariableFeatures(proneu_neus_male)
proneu_neus_male <- ScaleData(proneu_neus_male, 
                                  features = VariableFeatures(proneu_neus_male))
```

```{r}
res_female_proneuure <- age_trend_clusters(proneu_neus_female)

res_female_proneuure$changing_genes$gene <- rownames(res_female_proneuure$changing_genes)

proneu_female_full <- merge(
  res_female_proneuure$gene_cluster_df,
  res_female_proneuure$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(proneu_female_full, "proneu_female_changing_genes.csv")
```

```{r}
res_male_proneu <- age_trend_clusters(proneu_neus_male)

res_male_proneu$changing_genes$gene <- rownames(res_male_proneu$changing_genes)

proneu_male_full <- merge(
  res_male_proneu$gene_cluster_df,
  res_male_proneu$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(proneu_male_full, "proneu_male_changing_genes.csv")
```





#### GMPs

```{r}
gmps_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "GMPs" & age != "9m")
gmps_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "GMPs" & age != "9m") 

DefaultAssay(gmps_female) <- "RNA"
gmps_female <- NormalizeData(gmps_female)
gmps_female <- FindVariableFeatures(gmps_female)
gmps_female <- ScaleData(gmps_female, 
                                  features = VariableFeatures(gmps_female))

DefaultAssay(gmps_male) <- "RNA"
gmps_male <- NormalizeData(gmps_male)
gmps_male <- FindVariableFeatures(gmps_male)
gmps_male <- ScaleData(gmps_male, 
                                  features = VariableFeatures(gmps_male))
```

```{r}
res_female_gmpsure <- age_trend_clusters(gmps_female)

res_female_gmpsure$changing_genes$gene <- rownames(res_female_gmpsure$changing_genes)

gmps_female_full <- merge(
  res_female_gmpsure$gene_cluster_df,
  res_female_gmpsure$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(gmps_female_full, "gmps_female_changing_genes.csv")
```

```{r}
res_male_gmps <- age_trend_clusters(gmps_male)

res_male_gmps$changing_genes$gene <- rownames(res_male_gmps$changing_genes)

gmps_male_full <- merge(
  res_male_gmps$gene_cluster_df,
  res_male_gmps$changing_genes,
  by = "gene", sort = FALSE
)

write.csv(gmps_male_full, "gmps_male_changing_genes.csv")
```

### DeSEQ2

```{r}
DimPlot(gmps_neus_all, group.by = "CytoTRACE2_Potency")

neus_all <- subset(gmps_neus_all, CytoTRACE2_Potency == "Differentiated" & age != "9m")

DimPlot(neus_all, group.by = "sex")
```

```{r}
library(Seurat); library(glmGamPoi)

DefaultAssay(neus_all) <- "RNA"
neus_all$age <- factor(neus_all$age, levels = c("3m","12m","18m"))
neus_all$sex <- factor(neus_all$sex, levels = c("female","male"))

counts <- GetAssayData(neus_all, assay = "RNA", layer = "counts")  # slot= on Seurat v4
meta   <- neus_all@meta.data[, c("age","sex")]

# global detection filter — define ONCE on all cells, not per group
keep_genes <- rowSums(counts > 0) >= 10
counts <- counts[keep_genes, ]

run_lrt <- function(cts, md, full, reduced) {
  fit <- glm_gp(cts, design = full, col_data = md,
                size_factors = "normed_sum", on_disk = FALSE)
  res <- test_de(fit, reduced_design = reduced)
  list(fit = fit, res = res)
}

f <- meta$sex == "female"
m <- meta$sex == "male"

out_F <- run_lrt(counts[, f], meta[f, ], ~ age, ~ 1)
out_M <- run_lrt(counts[, m], meta[m, ], ~ age, ~ 1)

mu <- exp(out_F$fit$Beta)  
out_int <- run_lrt(counts, meta, ~ age * sex, ~ age + sex)

B <- out_F$fit$Beta
means_log <- cbind(`3m` = B[,1], `12m` = B[,1] + B[,2], `18m` = B[,1] + B[,3])

set.seed(42)
nperm <- 50

null_p <- unlist(replicate(nperm, {
  md <- meta
  md$age <- factor(ave(as.character(md$age), md$sex, FUN = sample),
                   levels = levels(meta$age))
  run_lrt(counts[, f], md[f, ], ~ age, ~ 1)$res$pval
}, simplify = FALSE))

# empirical FDR at a candidate cutoff
emp_fdr <- function(cut) (mean(null_p < cut) * nrow(counts)) /
                          max(1, sum(out_F$res$pval < cut))
thresh <- optimise(function(c) abs(emp_fdr(c) - 0.05), c(1e-8, 0.05))$minimum
```

```{r}
cd <- colData(combined_cds)

cd <- colData(cds); cd$age <- factor(cd$age, levels = c("3m","12m","18m"))
by_stage <- lapply(levels(cd$stage), function(s) {
  k <- cd$stage == s & cd$sex == "female"
  if (sum(k) < 50) return(NULL)
  run_lrt(counts(cds)[keep_genes, k], droplevels(cd[k, ]), ~ age, ~ 1)
})
```

```{r}
# ============================================================
# Age and sex×age DE in mature neutrophils — cell-level NB-GLM
# ============================================================
library(Seurat); library(glmGamPoi)

set.seed(42)
NPERM  <- 50     # raise to 100+ if thresholds move between seeds
TARGET <- 0.05   # empirical FDR

# ---- 1. metadata + counts ----------------------------------
DefaultAssay(neus_all) <- "RNA"
neus_all$age <- factor(neus_all$age, levels = c("3m","12m","18m"))
neus_all$sex <- factor(neus_all$sex, levels = c("female","male"))

counts <- GetAssayData(neus_all, assay = "RNA", layer = "counts")
meta   <- neus_all@meta.data[, c("age","sex")]

ok     <- !is.na(meta$age) & !is.na(meta$sex)
counts <- counts[, ok]
meta   <- droplevels(meta[ok, ])
print(table(meta$sex, meta$age))          # confirm against 422/330/560/96/332/248

keep_genes <- rowSums(counts > 0) >= 10   # global filter, defined once
counts     <- counts[keep_genes, ]
cat(nrow(counts), "genes,", ncol(counts), "cells\n")

f <- meta$sex == "female"
m <- meta$sex == "male"

# ---- 2. fit helper -----------------------------------------
run_lrt <- function(cts, md, full, reduced) {
  fit <- glm_gp(cts, design = full, col_data = md,
                size_factors = "normed_sum", on_disk = FALSE)
  list(fit = fit, res = test_de(fit, reduced_design = reduced))
}

out_F   <- run_lrt(counts[, f], droplevels(meta[f, ]), ~ age, ~ 1)
out_M   <- run_lrt(counts[, m], droplevels(meta[m, ]), ~ age, ~ 1)
out_int <- run_lrt(counts, meta, ~ age * sex, ~ age + sex)

stopifnot(all(out_F$res$df1 == 2), all(out_M$res$df1 == 2),
          all(out_int$res$df1 == 2))     # all three spend 2 df

# ---- 3. permutation nulls ----------------------------------
# per-sex: shuffle age within the subset (breaks age, keeps depth structure)
# interaction: shuffle sex within age (keeps the age main effect intact)
perm_null <- function(cts, md, full, reduced, shuffle) {
  unlist(lapply(seq_len(NPERM), function(i) {
    md2 <- shuffle(md)
    run_lrt(cts, md2, full, reduced)$res$pval
  }))
}

shuf_age <- function(md) { md$age <- sample(md$age); md }
shuf_sex <- function(md) {
  md$sex <- ave(as.character(md$sex), md$age, FUN = sample)
  md$sex <- factor(md$sex, levels = levels(meta$sex)); md
}

null_F   <- perm_null(counts[, f], droplevels(meta[f, ]), ~ age, ~ 1, shuf_age)
null_M   <- perm_null(counts[, m], droplevels(meta[m, ]), ~ age, ~ 1, shuf_age)
null_int <- perm_null(counts, meta, ~ age * sex, ~ age + sex, shuf_sex)

# ---- 4. empirical-FDR threshold ----------------------------
calib <- function(obs_p, null_p, target = TARGET) {
  grid <- 10^seq(-8, log10(0.05), length.out = 200)
  fdr  <- sapply(grid, function(c)
            (mean(null_p < c) * length(obs_p)) / max(1, sum(obs_p < c)))
  ok   <- which(fdr <= target)
  if (!length(ok)) return(list(thresh = 0, fdr = NA_real_))
  i <- max(ok); list(thresh = grid[i], fdr = fdr[i])
}

cal_F   <- calib(out_F$res$pval,   null_F)
cal_M   <- calib(out_M$res$pval,   null_M)
cal_int <- calib(out_int$res$pval, null_int)

# ---- 5. fitted means, z-scored (same axis as your patterns) --
group_means <- function(fit) {
  B <- fit$Beta
  cbind(`3m` = B[,1], `12m` = B[,1] + B[,2], `18m` = B[,1] + B[,3])
}
z_F <- t(scale(t(group_means(out_F$fit))))
z_M <- t(scale(t(group_means(out_M$fit))))

# ---- 6. tables ---------------------------------------------
build <- function(out, cal, z = NULL) {
  tb <- out$res[out$res$pval < cal$thresh, c("name","pval","adj_pval","f_statistic")]
  tb <- tb[order(tb$pval), ]
  if (!is.null(z)) tb <- cbind(tb, z[match(tb$name, rownames(z)), , drop = FALSE])
  tb
}
tab_F   <- build(out_F,   cal_F,   z_F)
tab_M   <- build(out_M,   cal_M,   z_M)
tab_int <- build(out_int, cal_int)

genes_F <- tab_F$name; genes_M <- tab_M$name; genes_int <- tab_int$name

# ---- 7. sensitivity: interaction without the 96-cell 12m male group
sub <- meta$age != "12m"
out_int_no12 <- run_lrt(counts[, sub], droplevels(meta[sub, ]),
                        ~ age * sex, ~ age + sex)
robust_int <- intersect(genes_int,
                        out_int_no12$res$name[out_int_no12$res$pval < 0.05])

# ---- 8. report ---------------------------------------------
cat(sprintf("female : thresh %.2e -> %d genes (BH<0.05: %d)\n",
            cal_F$thresh, length(genes_F), sum(out_F$res$adj_pval < 0.05)))
cat(sprintf("male   : thresh %.2e -> %d genes (BH<0.05: %d)\n",
            cal_M$thresh, length(genes_M), sum(out_M$res$adj_pval < 0.05)))
cat(sprintf("age*sex: thresh %.2e -> %d genes; %d survive dropping 12m\n",
            cal_int$thresh, length(genes_int), length(robust_int)))

write.csv(tab_F,   "omnibus_female.csv",      row.names = FALSE)
write.csv(tab_M,   "omnibus_male.csv",        row.names = FALSE)
write.csv(tab_int, "interaction_age_sex.csv", row.names = FALSE)
```


```{r}
# ==============================================================
# Genes differentially expressed anywhere along the neutrophil
# developmental trajectory, by age and/or sex.  3m vs 18m only.
#
# Null      : gene has a developmental profile, identical in all groups
# Alternative: that profile differs by age, by sex, or in shape
# ==============================================================

library(monocle3)
library(glmGamPoi)
library(splines)
library(SingleCellExperiment)

set.seed(42)

## ---- column names in colData(combined_cds) -------------------
AGE_COL   <- "age"
SEX_COL   <- "sex"
STAGE_COL <- "stage"      # set to NA if you have no stage labels
PT_DF     <- 3            # spline df for pseudotime
NPERM     <- 20           # start here; raise once thresholds look stable
TARGET    <- 0.05         # empirical FDR
MIN_CELLS <- 10           # gene detected in >= this many cells


# ==============================================================
# 1. Extract and clean
# ==============================================================
cd$age <- factor(as.character(cd[[AGE_COL]]), levels = c("3m", "12m", "18m"))
cd$sex <- factor(as.character(cd[[SEX_COL]]), levels = c("female", "male"))

# anything that failed to map is now NA — catch it here, loudly
if (anyNA(cd$age) || anyNA(cd$sex)) {
  cat("unmapped labels:\n")
  print(table(as.character(cd[[AGE_COL]])[is.na(cd$age)], useNA = "ifany"))
  print(table(as.character(cd[[SEX_COL]])[is.na(cd$sex)], useNA = "ifany"))
}

keep_cells <- is.finite(cd$pt) & !is.na(cd$age) & !is.na(cd$sex) &
              cd$age != "12m"
cds <- cds[, keep_cells]
cd  <- droplevels(cd[keep_cells, ])
cd$age <- droplevels(cd$age)   # drops the now-empty 12m level
cd$grp <- droplevels(interaction(cd$age, cd$sex, sep = "_"))

stopifnot(!anyNA(model.frame(f_lean, cd, na.action = na.pass)))


# ==============================================================
# 2. Restrict to the pseudotime span covered by ALL four groups
#
# Group-specific splines extrapolate freely outside their own
# cell range, and that divergence reads as a real difference.
# ==============================================================
rng <- tapply(cd$pt, cd$grp, range)
lo  <- max(vapply(rng, `[`, numeric(1), 1))
hi  <- min(vapply(rng, `[`, numeric(1), 2))
cat(sprintf("\nshared pseudotime span: %.3f - %.3f (full: %.3f - %.3f)\n",
            lo, hi, min(cd$pt), max(cd$pt)))

# density across the span, per group -- look for empty/near-empty cells
brk <- quantile(cd$pt, 0:5 / 5)
cat("\n== cells per pseudotime quintile x group (BEFORE trimming) ==\n")
print(table(cut(cd$pt, brk, include.lowest = TRUE), cd$grp))

in_span <- cd$pt >= lo & cd$pt <= hi
cat(sprintf("\ntrimming removes %d of %d cells (%.1f%%)\n",
            sum(!in_span), length(in_span), 100 * mean(!in_span)))
if (!is.na(STAGE_COL) && STAGE_COL %in% colnames(cd)) {
  cat("stages lost to trimming:\n")
  print(table(cd[[STAGE_COL]][!in_span]))
}

cds <- cds[, in_span]
cd  <- droplevels(cd[in_span, ])

cat("\n== final ==\n"); print(table(cd$sex, cd$age))
if (!is.na(STAGE_COL) && STAGE_COL %in% colnames(cd))
  print(table(cd[[STAGE_COL]], cd$grp))


# ==============================================================
# 3. Global gene filter -- defined once, on all cells
# ==============================================================
cts <- counts(cds)
cts <- cts[rowSums(cts > 0) >= MIN_CELLS, ]
cat(sprintf("\n%d genes x %d cells\n", nrow(cts), ncol(cts)))


# ==============================================================
# 4. Fit helper
# ==============================================================
run_lrt <- function(cts, md, full, reduced) {
  fit <- glm_gp(cts, design = full, col_data = md,
                size_factors = "normed_sum", on_disk = FALSE)
  list(fit = fit, res = test_de(fit, reduced_design = reduced))
}


# ==============================================================
# 5. Models
#
# LEAN is primary.  It lets both the age effect and the sex effect
# vary along pseudotime -- the "anywhere along the trajectory"
# part -- but omits the pt:age:sex term, which costs 3 df and is
# the worst-estimated piece of the model.
# FULL adds that three-way term; compare hit counts to see whether
# it earns its degrees of freedom.
# ==============================================================
f_lean <- ~ ns(pt, PT_DF) + age * sex +
            ns(pt, PT_DF):age + ns(pt, PT_DF):sex
f_full <- ~ ns(pt, PT_DF) * age * sex
f_null <- ~ ns(pt, PT_DF)                       # developmental profile only

# marginal decomposition: what remains when one factor is allowed
# to act freely in the reduced model
f_no_age <- ~ ns(pt, PT_DF) * sex               # -> tests any age effect
f_no_sex <- ~ ns(pt, PT_DF) * age               # -> tests any sex effect

out_any  <- run_lrt(cts, cd, f_lean, f_null)      # PRIMARY
out_full <- run_lrt(cts, cd, f_full, f_null)      # with three-way
out_age  <- run_lrt(cts, cd, f_full, f_no_age)    # age, marginal over sex
out_sex  <- run_lrt(cts, cd, f_full, f_no_sex)    # sex, marginal over age

cat("\ndf spent -- any:", out_any$res$df1[1],
    " full:", out_full$res$df1[1],
    " age:",  out_age$res$df1[1],
    " sex:",  out_sex$res$df1[1], "\n")


# ==============================================================
# 6. Permutation calibration
#
# Null = no dependence on age or sex, so age and sex are shuffled
# TOGETHER (keeping the pair intact) and pseudotime stays fixed.
# This preserves each cell's developmental position and library
# depth while destroying the group labels.
# ==============================================================
shuf_both <- function(md) {
  i <- sample(nrow(md)); md$age <- md$age[i]; md$sex <- md$sex[i]; md
}

perm_null <- function(cts, md, full, reduced, nperm = NPERM) {
  unlist(lapply(seq_len(nperm), function(i) {
    cat("."); if (i %% 10 == 0) cat(i, "\n")
    run_lrt(cts, shuf_both(md), full, reduced)$res$pval
  }))
}

cat("\npermuting (", NPERM, " x 4 fits, expect this to be slow)\n")
null_any <- perm_null(cts, cd, f_lean, f_null)
null_age <- perm_null(cts, cd, f_full, f_no_age)
null_sex <- perm_null(cts, cd, f_full, f_no_sex)

calib <- function(obs_p, null_p, target = TARGET) {
  grid <- 10^seq(-10, log10(0.05), length.out = 300)
  fdr  <- vapply(grid, function(c)
            (mean(null_p < c) * length(obs_p)) / max(1, sum(obs_p < c)),
            numeric(1))
  ok <- which(fdr <= target)
  if (!length(ok)) return(list(thresh = 0, fdr = NA_real_))
  i <- max(ok); list(thresh = grid[i], fdr = fdr[i])
}

cal_any <- calib(out_any$res$pval, null_any)
cal_age <- calib(out_age$res$pval, null_age)
cal_sex <- calib(out_sex$res$pval, null_sex)


# ==============================================================
# 7. Assemble and label
# ==============================================================
hits <- function(out, cal) out$res$name[out$res$pval < cal$thresh]

g_any <- hits(out_any, cal_any)
g_age <- hits(out_age, cal_age)
g_sex <- hits(out_sex, cal_sex)

tab <- out_any$res[out_any$res$name %in% g_any,
                   c("name", "pval", "adj_pval", "f_statistic")]
tab <- tab[order(tab$pval), ]
tab$age_effect <- tab$name %in% g_age
tab$sex_effect <- tab$name %in% g_sex
tab$driver <- with(tab, ifelse(age_effect & sex_effect, "both",
                        ifelse(age_effect, "age",
                        ifelse(sex_effect, "sex", "joint_only"))))

# where along the trajectory does each hit differ?  fitted values
# per group at five points across the span, for characterisation
pt_grid <- seq(min(cd$pt), max(cd$pt), length.out = 5)
newdat  <- expand.grid(pt = pt_grid, age = levels(cd$age), sex = levels(cd$sex))
mm      <- model.matrix(f_lean, newdat)
pred    <- out_any$fit$Beta[tab$name, , drop = FALSE] %*% t(mm)
colnames(pred) <- with(newdat, paste0("pt", round(pt, 2), "_", age, "_", sex))


# ==============================================================
# 8. Report
# ==============================================================
cat(sprintf("\nany age/or sex effect : thresh %.2e -> %d genes (BH<0.05: %d)\n",
            cal_any$thresh, length(g_any), sum(out_any$res$adj_pval < 0.05)))
cat(sprintf("  age-driven : %d\n  sex-driven : %d\n  both       : %d\n  joint only : %d\n",
            sum(tab$driver == "age"), sum(tab$driver == "sex"),
            sum(tab$driver == "both"), sum(tab$driver == "joint_only")))
cat(sprintf("\nthree-way model returns %d genes at BH<0.05 vs %d for lean\n",
            sum(out_full$res$adj_pval < 0.05), sum(out_any$res$adj_pval < 0.05)))

# does this recover the mature-compartment result, and what is new?
if (exists("genes_F")) {
  cat("\noverlap with the 56 mature-only female hits:",
      length(intersect(g_any, genes_F)), "\n")
  cat("trajectory-only (NOT in the mature analysis):",
      length(setdiff(g_any, genes_F)), "\n")
}

isg <- c("Isg15","Ifi27l2a","Oas1a","Oas1g","Oas3","Oasl2",
         "Slfn3","Slfn4","Ifitm3","H2-T24")
cat("\nISG module recovered:", paste(intersect(g_any, isg), collapse = ", "), "\n")

write.csv(tab, "pt_age_sex_hits.csv", row.names = FALSE)
write.csv(as.data.frame(pred), "pt_age_sex_fitted.csv")
saveRDS(list(any = out_any, full = out_full, age = out_age, sex = out_sex,
             cal = list(any = cal_any, age = cal_age, sex = cal_sex),
             cd = cd),
        "pt_age_sex_fits.rds")
```


### FitGAM

```{r}
library(tradeSeq)

age_num <- as.numeric(gsub("[^0-9.]", "", as.character(neus_all$age)))
pseudotime <- matrix(age_num, ncol = 1)
cellWeights <- matrix(1, nrow = ncol(mat_neus), ncol = 1)

counts <- GetAssayData(neus_all, assay = "RNA", layer = "counts")

sce <- fitGAM(
  counts      = as.matrix(counts),
  pseudotime  = pseudotime,
  cellWeights = cellWeights,
  conditions  = factor(neus_all$sex),   # this is the part that matters
  nknots      = 3
)

saveRDS(sce, "timecourse_all_mat_immat_GAM_model.rds")

assoRes <- associationTest(sce)
head(assoRes)
assoRes <- assoRes %>% na.omit() %>% subset(pvalue < 0.05)

startRes <- startVsEndTest(sce)
startRes <- startRes %>% na.omit() %>% subset(pvalue < 0.05)

res <- conditionTest(sce, l2fc = 0)
res$padj <- p.adjust(res$pvalue, "fdr")
res <- res %>% na.omit() %>% subset(padj < 0.05)

asso_per_sex  <- associationTest(sce, lineages = TRUE)
start_per_sex <- startVsEndTest(sce, lineages = TRUE)

endRes <- diffEndTest(sce)

add_fdr <- function(x) {
  x$padj <- p.adjust(x$pvalue, "fdr")
  x
}
asso_full <- add_fdr(na.omit(associationTest(sce, lineages = TRUE)))
asso_sig  <- subset(asso_full, padj < 0.05)
```

```{r}
library(tradeSeq)

age_num <- as.numeric(gsub("[^0-9.]", "", as.character(mat_neus$age)))
pseudotime <- matrix(age_num, ncol = 1)
cellWeights <- matrix(1, nrow = ncol(mat_neus), ncol = 1)

counts <- GetAssayData(mat_neus, assay = "RNA", layer = "counts")

sce <- fitGAM(
  counts      = as.matrix(counts),
  pseudotime  = pseudotime,
  cellWeights = cellWeights,
  conditions  = factor(mat_neus$sex),   # this is the part that matters
  nknots      = 3
)

saveRDS(sce, "timecourse_all_GAM_model.rds")

assoRes <- associationTest(sce)
head(assoRes)
assoRes <- assoRes %>% na.omit() %>% subset(pvalue < 0.05)

startRes <- startVsEndTest(sce)
startRes <- startRes %>% na.omit() %>% subset(pvalue < 0.05)

res <- conditionTest(sce, l2fc = 0)
res$padj <- p.adjust(res$pvalue, "fdr")
res <- res %>% na.omit() %>% subset(padj < 0.05)

asso_per_sex  <- associationTest(sce, lineages = TRUE)
start_per_sex <- startVsEndTest(sce, lineages = TRUE)

endRes <- diffEndTest(sce)

add_fdr <- function(x) {
  x$padj <- p.adjust(x$pvalue, "fdr")
  x
}
asso_full <- add_fdr(na.omit(associationTest(sce, lineages = TRUE)))
asso_sig  <- subset(asso_full, padj < 0.05)
```

```{r}
plotSmoothers(sce, counts, gene = "Mmp8")
```


### My GSEA idea

```{r}
rho_female_changing_genes <- read_csv('/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis/female/between_ages/mat_female_changing_genes.csv')
rho_male_changing_genes <- read.csv('/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis/male/between_ages/mat_male_changing_genes.csv')

degs_18m_sexes <- read_csv('/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis/sex_comparisons/between_ages_sexes/degs_matneu_18m_RNA.csv')
degs_12m_sexes <- read_csv('/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis/sex_comparisons/between_ages_sexes/degs_matneu_12m_RNA.csv')
degs_3m_sexes <- read_csv('/Users/costahn/Desktop/Aging Mouse Data/CITE-seq/analysis/aging_BM_analysis/sex_comparisons/between_ages_sexes/degs_matneu_3m_RNA.csv')

#ident.1 = male (positive, higher in males)
#ident.2 = female (negayive, higher in females)
```

```{r}
male_increase <- rho_male_changing_genes %>% dplyr::filter(rho > 0)
male_decrease <- rho_male_changing_genes %>% dplyr::filter(rho < 0)

male_18m_degs_up <- degs_18m_sexes %>% dplyr::filter(avg_log2FC > 0)
male_18m_degs_down <- degs_18m_sexes %>% dplyr::filter(avg_log2FC < 0)

male_increase_increase <- subset(male_increase, gene %in% male_18m_degs_up$gene)


male_decrease <- subset(male_decrease, gene %in% degs_18m_sexes$gene)

ora_male_increase_increase <- gost(male_increase_increase$gene,
                          organism = "mmusculus")
```


```{r}
degs_18m_sexes <- degs_18m_sexes %>% dplyr::filter(abs(avg_log2FC) > 0.5)
degs_12m_sexes <- degs_12m_sexes %>% dplyr::filter(abs(avg_log2FC) > 0.5)
degs_3m_sexes <- degs_3m_sexes %>% dplyr::filter(abs(avg_log2FC) > 0.5)


changing_genes <- union(rho_female_changing_genes$gene, rho_male_changing_genes$gene)

degs_18m_sexes <- subset(degs_18m_sexes,
                         gene %in% changing_genes)

degs_12m_sexes <- subset(degs_12m_sexes,
                         gene %in% changing_genes)

degs_3m_sexes <- subset(degs_3m_sexes,
                         gene %in% changing_genes)

male_18m_degs <- degs_18m_sexes %>% dplyr::filter(avg_log2FC > 0)
female_18m_degs <- degs_18m_sexes %>% dplyr::filter(avg_log2FC < 0)

ora_18m_male <- gost(male_18m_degs$gene,
                     organism = "mmusculus")
```


### Claude GSEA

```{r}
################################################################################
## Sex x age interaction GSEA for mature neutrophils  (v2 — descriptive labels)
##
## Ranking metric: Fisher-z contrast of per-cell Spearman's rho (expression vs
## numeric age), computed independently within each sex.
##
##   z_diff = (atanh(rho_M) - atanh(rho_F)) / sqrt(1/(n_M-3) + 1/(n_F-3))
##
## z_diff > 0 means ONLY: the age trajectory is more positive in males than in
## females. It does NOT mean the gene is higher in males, nor that males age
## "faster". The direction is resolved per pathway in section 5 by going back to
## the two underlying rho values for every leading-edge gene.
################################################################################

library(Seurat)
library(fgsea)
library(msigdbr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(ggplot2)
library(conflicted)

conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("slice",  "dplyr")
conflict_prefer("count",  "dplyr")

set.seed(42)

## ---------------------------------------------------------------------------
## 0. Parameters
## ---------------------------------------------------------------------------

SEX_COL       <- "sex"
AGE_COL       <- "age"
ASSAY         <- "RNA"
MIN_PCT_CELLS <- 0.05    # detected in >= 5% of cells within each sex
MIN_SET_SIZE  <- 15
MAX_SET_SIZE  <- 500
RHO_FLAT      <- 0.10    # |rho| below this is treated as no trend
                         # (matches the threshold used in the original pipeline)

SEX_CHR_GENES <- c("Xist", "Tsix", "Ddx3y", "Uty", "Eif2s3y", "Kdm5d", "Jpx", "Ftx")

GRANULE_GENES <- c("Ngp", "Lcn2", "Ltf", "Camp", "Chil3", "Serpinb1a",
                   "Wfdc21", "Mmp8", "Retnlg", "S100a8", "S100a9")

## ---------------------------------------------------------------------------
## 1. Per-sex rho on a shared gene universe
## ---------------------------------------------------------------------------

compute_rho <- function(obj, sex_value) {
  stopifnot(SEX_COL %in% colnames(obj@meta.data),
            AGE_COL %in% colnames(obj@meta.data))

  cells <- colnames(obj)[obj@meta.data[[SEX_COL]] == sex_value]
  if (length(cells) == 0) stop("No cells matched sex_value = ", sex_value)
  sub <- subset(obj, cells = cells)

  expr <- GetAssayData(sub, assay = ASSAY, layer = "data")   # log-normalised
  age  <- suppressWarnings(
            as.numeric(gsub("[^0-9.]", "", as.character(sub@meta.data[[AGE_COL]]))))
  if (anyNA(age)) stop("Could not parse age values for ", sex_value)
  stopifnot(length(age) == ncol(expr))

  pct  <- Matrix::rowMeans(expr > 0)
  keep <- names(pct)[pct >= MIN_PCT_CELLS]
  expr <- expr[keep, , drop = FALSE]

  age_r  <- rank(age)
  expr_r <- t(apply(as.matrix(expr), 1, rank))

  rho <- as.numeric(cor(t(expr_r), age_r))
  tibble(gene = rownames(expr), rho = rho, n = ncol(expr))
}

mat_neus <- subset(gmps_neus_all, fine_neu_labels %in% c("mature", "immature"))

rho_M <- compute_rho(mat_neus, "male")
rho_F <- compute_rho(mat_neus, "female")

## ---------------------------------------------------------------------------
## 2. Fisher-z interaction statistic, with a per-gene pattern label
## ---------------------------------------------------------------------------

CLAMP <- 0.999

rank_tbl <- inner_join(
    dplyr::select(rho_M, gene, rho_M = rho, n_M = n),
    dplyr::select(rho_F, gene, rho_F = rho, n_F = n),
    by = "gene"
  ) %>%
  mutate(
    z_M    = atanh(pmax(pmin(rho_M, CLAMP), -CLAMP)),
    z_F    = atanh(pmax(pmin(rho_F, CLAMP), -CLAMP)),
    se     = sqrt(1 / (n_M - 3) + 1 / (n_F - 3)),
    z_diff = (z_M - z_F) / se,

    trend_M = case_when(rho_M >  RHO_FLAT ~ "up",
                        rho_M < -RHO_FLAT ~ "down",
                        TRUE              ~ "flat"),
    trend_F = case_when(rho_F >  RHO_FLAT ~ "up",
                        rho_F < -RHO_FLAT ~ "down",
                        TRUE              ~ "flat"),

    pattern = case_when(
      trend_M == "up"   & trend_F == "down" ~ "opposing: up in M, down in F",
      trend_M == "down" & trend_F == "up"   ~ "opposing: down in M, up in F",
      trend_M == "flat" & trend_F == "down" ~ "female-only: declines in F",
      trend_M == "flat" & trend_F == "up"   ~ "female-only: rises in F",
      trend_M == "down" & trend_F == "flat" ~ "male-only: declines in M",
      trend_M == "up"   & trend_F == "flat" ~ "male-only: rises in M",
      trend_M == "up"   & trend_F == "up"   ~ "shared: rises in both",
      trend_M == "down" & trend_F == "down" ~ "shared: declines in both",
      TRUE                                  ~ "no trend in either"
    ),
    divergent = grepl("^opposing", pattern)
  ) %>%
  dplyr::filter(is.finite(z_diff)) %>%
  dplyr::arrange(desc(z_diff))

cat("Genes in shared universe:", nrow(rank_tbl), "\n")
cat("z_diff range:", round(range(rank_tbl$z_diff), 2), "\n\n")
print(dplyr::count(rank_tbl, pattern, sort = TRUE))

ranks <- setNames(rank_tbl$z_diff, rank_tbl$gene)

## ---------------------------------------------------------------------------
## 3. Gene sets
## ---------------------------------------------------------------------------

gs_df <- msigdbr(species = "Mus musculus", collection = "C5", subcollection = "GO:BP")
# msigdbr < 10:
# gs_df <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:BP")

pathways <- split(gs_df$gene_symbol, gs_df$gs_name)
pathways <- lapply(pathways, function(g) intersect(unique(g), names(ranks)))
pathways <- pathways[lengths(pathways) >= MIN_SET_SIZE]

cat("Gene sets tested:", length(pathways), "\n")

## ---------------------------------------------------------------------------
## 4. Preranked GSEA
## ---------------------------------------------------------------------------

res <- fgseaMultilevel(pathways, ranks,
                       minSize = MIN_SET_SIZE, maxSize = MAX_SET_SIZE, eps = 0)

sig <- res[order(res$pval)][res$padj[order(res$pval)] < 0.05, ]
main <- if (nrow(sig) > 0) collapsePathways(sig, pathways, ranks) else list(mainPathways = character(0))

## ---------------------------------------------------------------------------
## 5. Decompose each leading edge into what is actually happening
## ---------------------------------------------------------------------------
## This is the section that replaces the old "male-skewed / female-skewed"
## label. For each pathway it reports which per-gene pattern dominates its
## leading edge, so the sign of NES can be read correctly.

lookup <- rank_tbl %>% dplyr::select(gene, rho_M, rho_F, pattern, divergent)

describe_leading_edge <- function(le) {
  tab <- lookup %>% dplyr::filter(gene %in% le)
  if (nrow(tab) == 0) {
    return(tibble(dominant_pattern = NA_character_, dominant_n = NA_integer_,
                  dominant_frac = NA_real_, divergent_frac = NA_real_,
                  mean_rho_M = NA_real_, mean_rho_F = NA_real_))
  }
  counts <- dplyr::count(tab, pattern, sort = TRUE)
  tibble(
    dominant_pattern = counts$pattern[1],
    dominant_n       = counts$n[1],
    dominant_frac    = counts$n[1] / nrow(tab),
    divergent_frac   = mean(tab$divergent),
    mean_rho_M       = mean(tab$rho_M),
    mean_rho_F       = mean(tab$rho_F)
  )
}

le_summary <- map_dfr(res$leadingEdge, describe_leading_edge)

res_out <- res %>%
  as_tibble() %>%
  bind_cols(le_summary) %>%
  mutate(
    independent = pathway %in% main$mainPathways,

    # Neutral restatement of the NES sign — no claim about which sex is "worse"
    trajectory_more_positive_in = ifelse(NES > 0, "male", "female"),

    # Human-readable summary, safe to paste into a figure legend
    interpretation = sprintf(
      "trajectory more positive in %s; leading edge mostly '%s' (%d/%d genes); mean rho M=%.2f F=%.2f",
      trajectory_more_positive_in, dominant_pattern,
      dominant_n, lengths(leadingEdge), mean_rho_M, mean_rho_F
    ),

    # How confident the directional story is
    signal_type = case_when(
      divergent_frac >= 0.5                      ~ "opposing trajectories",
      grepl("^male-only",   dominant_pattern)    ~ "male-specific change",
      grepl("^female-only", dominant_pattern)    ~ "female-specific change",
      grepl("^shared",      dominant_pattern)    ~ "shared direction, differing magnitude",
      TRUE                                       ~ "mixed / weak"
    ),

    leadingEdge_n = lengths(leadingEdge),
    sex_chr_in_LE = sapply(leadingEdge, function(g) paste(intersect(g, SEX_CHR_GENES), collapse = ";")),
    granule_in_LE = sapply(leadingEdge, function(g) paste(intersect(g, GRANULE_GENES), collapse = ";")),
    leadingEdge   = sapply(leadingEdge, paste, collapse = ";")
  ) %>%
  dplyr::select(pathway, NES, pval, padj, independent, signal_type,
                trajectory_more_positive_in, dominant_pattern, dominant_n,
                dominant_frac, divergent_frac, mean_rho_M, mean_rho_F,
                leadingEdge_n, sex_chr_in_LE, granule_in_LE,
                interpretation, leadingEdge) %>%
  dplyr::arrange(padj, desc(abs(NES)))

readr::write_csv(res_out, "gsea_sex_by_age_GOBP.csv")

## Quick look at the reportable set
res_out %>%
  dplyr::filter(padj < 0.05, independent) %>%
  dplyr::select(pathway, NES, padj, signal_type, dominant_pattern, dominant_frac) %>%
  print(n = 30)

## Inspect any single pathway gene by gene
inspect <- function(pathway_name) {
  le <- strsplit(res_out$leadingEdge[res_out$pathway == pathway_name], ";")[[1]]
  rank_tbl %>%
    dplyr::filter(gene %in% le) %>%
    dplyr::select(gene, rho_M, rho_F, z_diff, pattern) %>%
    dplyr::arrange(desc(abs(z_diff)))
}

inspect("GOBP_RESPONSE_TO_TYPE_I_INTERFERON")
################################################################################
```


```{r}
library(Seurat)
library(fgsea)
library(msigdbr)
library(dplyr)
library(tibble)
library(ggplot2)

set.seed(42)

## ---------------------------------------------------------------------------
## 0. Parameters
## ---------------------------------------------------------------------------

SEX_COL       <- "sex"          # metadata column holding "M" / "F"
AGE_COL       <- "age"  # metadata column holding 3, 12, 18
ASSAY         <- "RNA"
MIN_PCT_CELLS <- 0.05           # gene must be detected in >= 5% of cells IN BOTH sexes
MIN_SET_SIZE  <- 15
MAX_SET_SIZE  <- 500

# Genes to flag (not auto-removed) in the leading edges
SEX_CHR_GENES <- c("Xist", "Tsix", "Ddx3y", "Uty", "Eif2s3y", "Kdm5d", "Jpx", "Ftx")

# Highly abundant granule transcripts, for the compositional sensitivity run
GRANULE_GENES <- c("Ngp", "Lcn2", "Ltf", "Camp", "Chil3", "Serpinb1a",
                   "Wfdc21", "Mmp8", "Retnlg", "S100a8", "S100a9")

## ---------------------------------------------------------------------------
## 1. Compute rho per sex on a SHARED gene universe
## ---------------------------------------------------------------------------
## This is the part that matters most. If rho_M and rho_F were computed on
## different gene sets (e.g. each filtered separately), the contrast is not
## comparable. Recomputing both here guarantees one universe.

compute_rho <- function(obj, sex_value) {
  cells <- colnames(obj)[obj@meta.data[[SEX_COL]] == sex_value]
  sub   <- subset(obj, cells = cells)

  expr <- GetAssayData(sub, assay = ASSAY, layer = "data")   # log-normalised
  age  <- as.numeric(sub@meta.data[[AGE_COL]])

  # Detection filter within this sex
  pct  <- Matrix::rowMeans(expr > 0)
  keep <- names(pct)[pct >= MIN_PCT_CELLS]
  expr <- expr[keep, , drop = FALSE]

  # Spearman = Pearson on ranks. Rank age once, rank each gene across cells.
  age_r <- rank(age)
  expr_r <- t(apply(as.matrix(expr), 1, rank))

  rho <- as.numeric(cor(t(expr_r), age_r, method = "pearson"))
  names(rho) <- rownames(expr)

  tibble(gene = names(rho), rho = rho, n = ncol(expr))
}

mat_neus <- subset(gmps_neus_all, fine_neu_labels == "mature")

rho_M <- compute_rho(mat_neus, "male")
rho_F <- compute_rho(mat_neus, "female")

## ---------------------------------------------------------------------------
## 2. Fisher-z interaction statistic
## ---------------------------------------------------------------------------

CLAMP <- 0.999  # atanh(+/-1) is infinite; clamp before transforming

rank_tbl <- inner_join(
    dplyr::select(rho_M, gene, rho_M = rho, n_M = n),
    dplyr::select(rho_F, gene, rho_F = rho, n_F = n),
    by = "gene"
  ) %>%
  mutate(
    z_M    = atanh(pmax(pmin(rho_M, CLAMP), -CLAMP)),
    z_F    = atanh(pmax(pmin(rho_F, CLAMP), -CLAMP)),
    se     = sqrt(1 / (n_M - 3) + 1 / (n_F - 3)),
    z_diff = (z_M - z_F) / se
  ) %>%
  dplyr::filter(is.finite(z_diff)) %>%
  dplyr::arrange(desc(z_diff))

cat("Genes in shared universe:", nrow(rank_tbl), "\n")
cat("z_diff range:", round(range(rank_tbl$z_diff), 2), "\n")

## Note: z_diff is an interaction z-score, but do NOT read it as per-gene
## evidence. With n_M ~ 191 and n_F ~ 651 the SE is ~0.083, so a rho difference
## of 0.10 gives z ~ 1.2. Cells are pseudoreplicates (pooled animals), so the
## nominal null is wrong anyway. The value of z_diff here is that it produces a
## sensibly scaled RANKING, which is all GSEA needs.

ranks <- setNames(rank_tbl$z_diff, rank_tbl$gene)

## ---------------------------------------------------------------------------
## 3. Gene sets: GO Biological Process, mouse
## ---------------------------------------------------------------------------

# msigdbr >= 10:
gs_df <- msigdbr(species = "Mus musculus", collection = "C5", subcollection = "GO:BP")
# msigdbr < 10, use instead:
# gs_df <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:BP")

pathways <- split(gs_df$gene_symbol, gs_df$gs_name)

# Restrict gene sets to the tested universe so set sizes are honest
pathways <- lapply(pathways, function(g) intersect(unique(g), names(ranks)))
pathways <- pathways[lengths(pathways) >= MIN_SET_SIZE]

cat("Gene sets tested:", length(pathways), "\n")

## ---------------------------------------------------------------------------
## 4. Preranked GSEA
## ---------------------------------------------------------------------------

res <- fgseaMultilevel(
  pathways  = pathways,
  stats     = ranks,
  minSize   = MIN_SET_SIZE,
  maxSize   = MAX_SET_SIZE,
  eps       = 0
)

## GO:BP is heavily redundant — collapse nested/overlapping sets
main <- collapsePathways(
  res[order(res$pval)][res$padj[order(res$pval)] < 0.05, ],
  pathways, ranks
)

res_out <- res %>%
  as_tibble() %>%
  mutate(
    independent   = pathway %in% main$mainPathways,
    direction     = ifelse(NES > 0, "male-skewed ageing", "female-skewed ageing"),
    leadingEdge_n = lengths(leadingEdge),
    sex_chr_in_LE = sapply(leadingEdge, function(g) paste(intersect(g, SEX_CHR_GENES), collapse = ";")),
    granule_in_LE = sapply(leadingEdge, function(g) paste(intersect(g, GRANULE_GENES), collapse = ";")),
    leadingEdge   = sapply(leadingEdge, paste, collapse = ";")
  ) %>%
  arrange(padj, desc(abs(NES)))

readr::write_csv(res_out, "gsea_sex_by_age_GOBP.csv")

## ---------------------------------------------------------------------------
## 5. Sensitivity run: drop abundant granule transcripts
## ---------------------------------------------------------------------------
## Guards against the 18m compositional-dilution signature driving secretory /
## translation sets. Pathways that survive both runs are the ones to report.

ranks_sens <- ranks[!names(ranks) %in% GRANULE_GENES]

res_sens <- fgseaMultilevel(
  pathways = lapply(pathways, setdiff, GRANULE_GENES),
  stats    = ranks_sens,
  minSize  = MIN_SET_SIZE,
  maxSize  = MAX_SET_SIZE,
  eps      = 0
)

comparison <- res_out %>%
  dplyr::select(pathway, NES_full = NES, padj_full = padj, independent) %>%
  left_join(
    res_sens %>% as_tibble() %>% dplyr::select(pathway, NES_sens = NES, padj_sens = padj),
    by = "pathway"
  ) %>%
  dplyr::mutate(
    robust = padj_full < 0.05 & padj_sens < 0.05 & sign(NES_full) == sign(NES_sens)
  ) %>%
  dplyr::arrange(desc(robust), padj_full)

readr::write_csv(comparison, "gsea_sensitivity_comparison.csv")

## ---------------------------------------------------------------------------
## 6. Plot the robust, non-redundant hits
## ---------------------------------------------------------------------------

top <- comparison %>%
  dplyr::slice_head(n = 25) %>%
  dplyr::mutate(
    label = gsub("^GOBP_", "", pathway) %>% gsub("_", " ", .) %>% tolower()
  )

ggplot(top, aes(x = reorder(label, NES_full), y = NES_full, fill = NES_full > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#2C6E9B", "FALSE" = "#B5482A"),
                    labels = c("TRUE" = "male-skewed", "FALSE" = "female-skewed"),
                    name = "Ageing trajectory") +
  labs(x = NULL, y = "NES (sex x age interaction)",
       title = "GO:BP pathways with sex-differential ageing in mature neutrophils") +
  theme_minimal(base_size = 11)

ggsave("gsea_sex_by_age_GOBP.pdf", width = 9, height = 7)

## ---------------------------------------------------------------------------
## 7. RECOMMENDED: empirical null by permuting sex labels
## ---------------------------------------------------------------------------
## fgsea's null permutes gene labels, which ignores inter-gene correlation and
## is anti-conservative for coexpressed modules. Since you already built
## permutation calibration into the corrected pipeline, reuse that approach:
## shuffle the sex label across cells WITHIN each age group, recompute rho for
## both pseudo-sexes, rebuild z_diff, rerun fgsea, and record the max |NES| per
## permutation. Compare observed NES against that distribution.
##
## for (i in 1:200) {
##   obj$sex_perm <- ave(obj$sex, obj[[AGE_COL]][,1], FUN = sample)
##   ... recompute rho for sex_perm == "M" / "F", rebuild ranks, fgsea ...
##   null_max[i] <- max(abs(perm_res$NES))
## }
## empirical_p <- mean(null_max >= abs(observed_NES))
################################################################################
```

```{r}
################################################################################
## Per-age sex contrasts + comparison against the sex x age interaction
##
## Part A: GSEA on M vs F within each of 3m, 12m, 18m  (descriptive)
## Part B: line up those NES against the interaction NES (interpretive)
##
## Assumes `pathways`, `ranks` (interaction z_diff) and `res` already exist
## from sex_by_age_gsea.R
################################################################################

library(Seurat)
library(fgsea)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("slice",  "dplyr")
conflict_prefer("rename", "dplyr")

SEX_COL <- "sex"
AGE_COL <- "age"
ASSAY   <- "RNA"
MIN_PCT <- 0.05
AGES    <- c(3, 12, 18)

## ---------------------------------------------------------------------------
## A1. Sex contrast within each age
## ---------------------------------------------------------------------------
## Rank on avg_log2FC, NOT on p-value. Cell numbers differ ~10-fold between the
## male and female arms at 12m; a Wilcoxon p is a function of that imbalance,
## whereas log2FC is not.

age_num <- suppressWarnings(
  as.numeric(gsub("[^0-9.]", "", as.character(mat_neus@meta.data[[AGE_COL]])))
)
mat_neus$age_num <- age_num

sex_de_at_age <- function(obj, this_age) {
  cells <- colnames(obj)[obj$age_num == this_age]
  sub   <- subset(obj, cells = cells)
  Idents(sub) <- sub@meta.data[[SEX_COL]]

  message("age ", this_age, "m: ", paste(table(Idents(sub)), collapse = " / "), " cells")

  FindMarkers(
    sub, ident.1 = "male", ident.2 = "female",
    assay = ASSAY, slot = "data",
    logfc.threshold = 0, min.pct = MIN_PCT,   # keep everything above detection
    test.use = "wilcox"
  ) %>% rownames_to_column("gene")
}

de_list <- lapply(setNames(AGES, paste0(AGES, "m")),
                  function(a) sex_de_at_age(mat_neus, a))

## ---------------------------------------------------------------------------
## A2. Shared gene universe across all three ages AND the interaction
## ---------------------------------------------------------------------------
## Without this the four NES columns are computed on different backgrounds and
## are not comparable — which is the whole point of the exercise.

universe <- Reduce(intersect, c(lapply(de_list, `[[`, "gene"), list(names(ranks))))
cat("Shared universe:", length(universe), "genes\n")

rank_at_age <- lapply(de_list, function(d) {
  d <- d %>% filter(gene %in% universe) %>% arrange(desc(avg_log2FC))
  setNames(d$avg_log2FC, d$gene)
})

pathways_u <- lapply(pathways, intersect, universe)
pathways_u <- pathways_u[lengths(pathways_u) >= 15]

## ---------------------------------------------------------------------------
## A3. Run GSEA per age
## ---------------------------------------------------------------------------

gsea_by_age <- lapply(names(rank_at_age), function(nm) {
  fgseaMultilevel(pathways_u, rank_at_age[[nm]],
                  minSize = 15, maxSize = 500, eps = 0) %>%
    as_tibble() %>%
    select(pathway, NES, pval, padj) %>%
    rename_with(~ paste0(.x, "_", nm), c(NES, pval, padj))
})
names(gsea_by_age) <- names(rank_at_age)

## Interaction result on the same restricted universe, for a fair comparison
res_int <- fgseaMultilevel(pathways_u, ranks[universe],
                           minSize = 15, maxSize = 500, eps = 0) %>%
  as_tibble() %>%
  select(pathway, NES_int = NES, padj_int = padj)

## ---------------------------------------------------------------------------
## B1. Assemble one table
## ---------------------------------------------------------------------------

combined <- Reduce(function(a, b) full_join(a, b, by = "pathway"), gsea_by_age) %>%
  full_join(res_int, by = "pathway") %>%
  mutate(
    # how much the sex difference moves across the lifespan
    delta_18_3 = NES_18m - NES_3m,
    # is the direction of the sex difference stable across all three ages?
    consistent_sign = sign(NES_3m) == sign(NES_12m) & sign(NES_12m) == sign(NES_18m),
    # does the per-age picture agree with the interaction statistic?
    agrees_with_int = sign(delta_18_3) == sign(NES_int),
    sig_any_age = pmin(padj_3m, padj_12m, padj_18m, na.rm = TRUE) < 0.05,
    sig_int     = padj_int < 0.05
  ) %>%
  arrange(padj_int)

readr::write_csv(combined, "gsea_perage_vs_interaction.csv")

## ---------------------------------------------------------------------------
## B2. Classify
## ---------------------------------------------------------------------------

classified <- combined %>%
  mutate(class = case_when(
    sig_int & sig_any_age & agrees_with_int  ~ "sex-differential ageing (supported)",
    sig_int & !agrees_with_int               ~ "interaction only - check 12m leverage",
    !sig_int & sig_any_age & consistent_sign &
      abs(delta_18_3) < 0.5                  ~ "constitutive sex difference",
    sig_any_age                              ~ "age-specific sex difference",
    TRUE                                     ~ "ns"
  ))

table(classified$class)

## ---------------------------------------------------------------------------
## B3. NES trajectory heatmap for the supported hits
## ---------------------------------------------------------------------------

top <- classified %>%
  filter(class == "sex-differential ageing (supported)") %>%
  slice_head(n = 30)

plot_df <- top %>%
  select(pathway, NES_3m, NES_12m, NES_18m) %>%
  pivot_longer(-pathway, names_to = "age", values_to = "NES") %>%
  mutate(
    age   = factor(gsub("NES_", "", age), levels = c("3m", "12m", "18m")),
    label = tolower(gsub("_", " ", gsub("^GOBP_", "", pathway)))
  )

ggplot(plot_df, aes(age, reorder(label, NES), fill = NES)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "#B5482A", mid = "grey95", high = "#2C6E9B",
                       midpoint = 0, name = "NES\n(male vs female)") +
  labs(x = NULL, y = NULL,
       title = "Sex difference by age, for pathways flagged by the interaction test") +
  theme_minimal(base_size = 10)

ggsave("nes_trajectories.pdf", width = 9, height = 8)

## ---------------------------------------------------------------------------
## B4. Overlap between the per-age significant sets
## ---------------------------------------------------------------------------

sig_sets <- list(
  `3m`  = combined$pathway[which(combined$padj_3m  < 0.05)],
  `12m` = combined$pathway[which(combined$padj_12m < 0.05)],
  `18m` = combined$pathway[which(combined$padj_18m < 0.05)],
  interaction = combined$pathway[which(combined$padj_int < 0.05)]
)
sapply(sig_sets, length)

jaccard <- outer(seq_along(sig_sets), seq_along(sig_sets), Vectorize(function(i, j) {
  a <- sig_sets[[i]]; b <- sig_sets[[j]]
  if (length(union(a, b)) == 0) return(NA)
  length(intersect(a, b)) / length(union(a, b))
}))
dimnames(jaccard) <- list(names(sig_sets), names(sig_sets))
round(jaccard, 2)
################################################################################
```


# Other

Redefining the lists change between sexes with Claude recommended approach

```{r}
avg_expr <- AverageExpression(subset(gmps_neus_all, age != "9m" & fine_neu_labels == "mature"), 
                              group.by = "age_sex",
                              assays = "RNA", 
                              slot = "data")$RNA

age_order <- c("g3m-M", "g12m-M", "g18m-M", "g3m-F", "g12m-F", "g18m-F")
avg_expr <- avg_expr[, age_order]   # reorder columns to true age order

avg_expr <- avg_expr[rownames(changing_genes), ]
z_scored <- t(scale(t(avg_expr)))
z_scored <- t(scale(t(avg_expr)))

shared_genes <- c("Adam19", "Adpgk", "Ccnd3", "Chil5", "Clec4a2", "Cpne2",
                 "Dstn", "Fpr2", "Il18rap", "Mospd1", "Oaz2", "Per1",
                 "R3hdm4", "Rabggta", "Retnlg", "Tmcc1", "Unc93b1", "Zfp236", 
                 "Anxa1", "Cd177", "Cdc37l1", "Lcn2", "Mgst1", "Ngp",
                 "S100a8", "S100a9", "Samd9l", "Wfdc21",
                 "Csf3r", "Ep300", "Fxyd5", "Hist1h2bc", "Hnrnpul1",
                 "Lst1", "Mindy2", "Plagl2", "Asprv1", "Atp6v1h", "Rassf2")

z <- subset(z_scored, rownames(z_scored) %in% shared_genes)

# hierarchical clustering into pattern groups
library(pheatmap)
clusters_all <- pheatmap(z, 
                     cluster_cols = FALSE, 
                     cutree_rows = 6,
                     show_rownames = TRUE)
```

```{r}
library(Seurat)

age_levels <- c("3m", "12m", "18m")   # 9m removed, final

# helper: Spearman age-trend per gene
trend_test <- function(obj) {
  age_num <- as.numeric(factor(obj$age, levels = age_levels))
  if (any(is.na(age_num)))
    stop("Some cells have an age not in ", paste(age_levels, collapse="/"),
         " -- check obj$age levels: ", paste(unique(obj$age), collapse=", "))
  expr <- GetAssayData(obj, layer = "data")
  res <- apply(expr, 1, function(x) {
    ct <- suppressWarnings(cor.test(x, age_num, method = "spearman", exact = FALSE))
    c(rho = unname(ct$estimate), pval = ct$p.value)
  })
  df <- as.data.frame(t(res))
  df$padj <- p.adjust(df$pval, method = "BH")
  df
}

tm <- trend_test(neus_male)     # male
tf <- trend_test(neus_female)   # female

tm_filt <- na.omit(tm)
tf_filt <- na.omit(tf)
write.csv(tm_filt, "male_BM_ages_rho.csv")
write.csv(tf_filt, "female_BM_ages_rho.csv")


cat("Male genes tested:  ", nrow(tm), " | Female:", nrow(tf), "
")
summary(abs(tm$rho)); summary(abs(tf$rho))

male_cut   <- quantile(abs(tm_filt$rho), 0.90, na.rm = TRUE)
female_cut <- quantile(abs(tf_filt$rho), 0.90, na.rm = TRUE)

male_genes   <- rownames(tm_filt)[abs(tm_filt$rho) > male_cut]
female_genes <- rownames(tf_filt)[abs(tf_filt$rho) > female_cut]
```

```{r}
library(ggplot2)

# tf = female trend_df with $rho, $padj  (from the diagnostic earlier)
# tm = male   trend_df with $rho, $padj

# 1. Compare the |rho| distributions directly
summary(abs(tm$rho)); summary(abs(tf$rho))

# 2. Is there a cliff right at 0.1? Count genes in bins near the threshold
brks <- c(0, 0.05, 0.08, 0.10, 0.12, 0.15, 0.2, 1)
cat("MALE |rho| bins:
");   print(table(cut(abs(tm$rho), brks)))
cat("FEMALE |rho| bins:
"); print(table(cut(abs(tf$rho), brks)))

# 3. Overlay the distributions
df <- rbind(data.frame(rho = abs(tm$rho), sex = "Male"),
            data.frame(rho = abs(tf$rho), sex = "Female"))
ggplot(df, aes(rho, fill = sex)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0.1, linetype = "dashed") +
  labs(x = "|Spearman rho| (age trend)", title = "Age-trend effect sizes by sex") +
  theme_minimal()

# Do males have systematically larger age-trend effect sizes than females?
wilcox.test(abs(tm$rho), abs(tf$rho), alternative = "greater")
```

FindMarkers() approach

```{r}
ages <- c("3m", "12m", "18m")
pairs <- combn(ages, 2, simplify = FALSE)   # all 3 pairwise comparisons

run_age_markers <- function(obj, label) {
  Idents(obj) <- "age"
  map_dfr(pairs, function(p) {
    FindMarkers(obj, ident.1 = p[2], ident.2 = p[1],   # older vs younger
               logfc.threshold = 0.25, min.pct = 0.1,
               test.use = "wilcox", verbose = FALSE) %>%
      rownames_to_column("gene") %>%
      mutate(comparison = paste0(p[1], "_vs_", p[2]), sex = label)
  })
}

male_markers   <- run_age_markers(neus_male,   "Male")
female_markers <- run_age_markers(neus_female, "Female")

write.csv(male_markers,   "male_age_markers.csv",   row.names = FALSE)
write.csv(female_markers, "female_age_markers.csv", row.names = FALSE)
```

Downsampling

```{r}
library(Seurat)

# Downsample every age group to the smallest group, within one sex.
# Idents must be set to age first -- subset(downsample=) works on Idents.
downsample_ages <- function(obj, age_col = "age", seed = 1) {
  set.seed(seed)
  Idents(obj) <- age_col
  n_min <- min(table(obj[[age_col]][, 1]))
  subset(obj, downsample = n_min)
}

male_ds   <- downsample_ages(neus_male)     # -> 26 per age, 78 cells
female_ds <- downsample_ages(neus_female)   # -> 185 per age, 555 cells
```

```{r}
# Age as numeric for trend testing
age_numeric <- as.numeric(factor(female_ds$age, levels = c("3m", "12m", "18m")))
# Spearman correlation per gene: tests for monotonic trend with age
# Run on log-normalized data slot
expr_mat <- GetAssayData(female_ds, layer = "data")
trend_results <- apply(expr_mat, 1, function(x) {
  ct <- cor.test(x, age_numeric, method = "spearman", exact = FALSE)
  c(rho = unname(ct$estimate), pval = ct$p.value)
})
trend_df <- as.data.frame(t(trend_results))
trend_df$padj <- p.adjust(trend_df$pval, method = "BH")
# Select genes with statistical evidence of a trend (not arbitrary top-N)
changing_genes <- trend_df |>
  dplyr::filter(abs(rho) > 0.1) |>
  arrange(padj)
# For trajectory clustering: use average expression per age, z-scored
avg_expr <- AverageExpression(female_ds, group.by = "age",
                              assays = "RNA", slot = "data")$RNA
age_order <- c("g3m", "g12m", "g18m")
avg_expr <- avg_expr[, age_order]   # reorder columns to true age order
avg_expr <- avg_expr[rownames(changing_genes), ]
z_scored <- t(scale(t(avg_expr)))
```

```{r}
# Age as numeric for trend testing
age_numeric <- as.numeric(factor(male_ds$age, levels = c("3m", "12m", "18m")))
# Spearman correlation per gene: tests for monotonic trend with age
# Run on log-normalized data slot
expr_mat <- GetAssayData(male_ds, layer = "data")
trend_results <- apply(expr_mat, 1, function(x) {
  ct <- cor.test(x, age_numeric, method = "spearman", exact = FALSE)
  c(rho = unname(ct$estimate), pval = ct$p.value)
})
trend_df <- as.data.frame(t(trend_results))
trend_df$padj <- p.adjust(trend_df$pval, method = "BH")
# Select genes with statistical evidence of a trend (not arbitrary top-N)
changing_genes <- trend_df |>
  dplyr::filter(abs(rho) > 0.1) |>
  arrange(padj)
# For trajectory clustering: use average expression per age, z-scored
avg_expr <- AverageExpression(male_ds, group.by = "age",
                              assays = "RNA", slot = "data")$RNA
age_order <- c("g3m", "g12m", "g18m")
avg_expr <- avg_expr[, age_order]   # reorder columns to true age order
avg_expr <- avg_expr[rownames(changing_genes), ]
z_scored <- t(scale(t(avg_expr)))
```

```{r}
downsample_all <- function(objs, age_col = "age", seed = 1) {
  n_min <- min(sapply(objs, function(o) min(table(o[[age_col]][, 1]))))
  lapply(objs, function(o) {
    set.seed(seed)
    Idents(o) <- age_col
    subset(o, downsample = n_min)
  })
}

ds <- downsample_all(list(male = neus_male, female = neus_female))
# -> 26 cells per age per sex, 156 cells total
```

Mature neutrophil DEGs

```{r}
library(Seurat)
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(Matrix)

AGES  <- c("3m", "12m", "18m")
PAIRS <- list(c("12m", "3m"), c("18m", "12m"), c("18m", "3m"))

CFG <- list(
  min_cells = 10,    # gene must be detected in >= this many cells ...
  min_frac  = 0.10,  # ... and >= this fraction, in EVERY age group
  n_perm    = 20,
  q         = 0.95   # threshold = this percentile of |log2FC| under shuffling
)

## ---------------------------------------------------------------------------
## 1. Global detection filter
##
## Requiring detection in all three age groups is what guarantees every gene
## has all three comparisons, and is what prevents a gene seen in 3 of 26
## cells from generating a 256-fold "change".
## ---------------------------------------------------------------------------

detectable_genes <- function(obj, cfg = CFG) {
  cnt  <- GetAssayData(obj, layer = "counts")
  ages <- as.character(obj$age)

  ok <- vapply(AGES, function(a) {
    idx <- ages == a
    need <- max(cfg$min_cells, ceiling(cfg$min_frac * sum(idx)))
    Matrix::rowSums(cnt[, idx, drop = FALSE] > 0) >= need
  }, logical(nrow(cnt)))

  keep <- rownames(cnt)[rowSums(ok) == length(AGES)]
  message(sprintf("  detection filter: %d of %d genes retained",
                  length(keep), nrow(cnt)))
  keep
}

## ---------------------------------------------------------------------------
## 2. Pairwise comparisons (used for both real and permuted labels)
##
## `features` is fixed from the real data so the permuted runs test exactly
## the same gene set -- otherwise the null would not be calibrated like for
## like. min.pct and logfc.threshold are 0 because filtering already happened.
## ---------------------------------------------------------------------------

run_pairwise <- function(obj, features, permute = FALSE, seed = 1) {
  ages <- as.character(obj$age)
  if (permute) { set.seed(seed); ages <- sample(ages) }
  obj$.age_use <- factor(ages, levels = AGES)
  Idents(obj) <- ".age_use"

  purrr::map_dfr(PAIRS, function(p) {
    FindMarkers(obj, ident.1 = p[1], ident.2 = p[2],
                features = features,
                logfc.threshold = 0, min.pct = 0,
                test.use = "wilcox", verbose = FALSE) %>%
      tibble::rownames_to_column("gene") %>%
      dplyr::transmute(gene,
                       comparison = paste0(p[2], "_to_", p[1]),
                       log2FC   = avg_log2FC,
                       pct_diff = pct.1 - pct.2)
  })
}

## ---------------------------------------------------------------------------
## 3. Permutation-calibrated thresholds, one per comparison per sex
## ---------------------------------------------------------------------------

calibrate <- function(obj, features, cfg = CFG) {
  purrr::map_dfr(seq_len(cfg$n_perm),
                 ~ run_pairwise(obj, features, permute = TRUE, seed = .x)) %>%
    dplyr::group_by(comparison) %>%
    dplyr::summarise(threshold = quantile(abs(log2FC), cfg$q, na.rm = TRUE),
                     .groups = "drop")
}

## ---------------------------------------------------------------------------
## 4. Trajectory shape, NA-safe
## ---------------------------------------------------------------------------

classify <- function(early, net, thr_e, thr_l, tol = NULL) {
  if (is.null(tol)) tol <- mean(c(thr_e, thr_l)) / 2

  P <- cbind(`3m` = 0, `12m` = early, `18m` = net)   # relative positions
  P <- P - rowMeans(P)                                # centre, as in z-scoring

  L <- ifelse(P >  tol, "high",
       ifelse(P < -tol, "low", "middle"))

  out <- apply(L, 1, paste, collapse = "_")
  out[is.na(early) | is.na(net)] <- NA_character_
  out
}

## ---------------------------------------------------------------------------
## 5. Main
## ---------------------------------------------------------------------------

age_changing <- function(obj, label, cfg = CFG) {
  message("\n=== ", label, " ===")
  feats <- detectable_genes(obj, cfg)
  if (length(feats) < 50)
    warning(label, ": only ", length(feats),
            " genes pass detection -- arm is likely underpowered")

  obs <- run_pairwise(obj, feats)
  thr <- calibrate(obj, feats, cfg)
  message("  permutation thresholds (|log2FC|):")
  print(as.data.frame(thr), row.names = FALSE)

  te <- thr$threshold[thr$comparison == "3m_to_12m"]
  tl <- thr$threshold[thr$comparison == "12m_to_18m"]
  tn <- thr$threshold[thr$comparison == "3m_to_18m"]

  wide <- obs %>%
    dplyr::select(gene, comparison, log2FC) %>%
    tidyr::pivot_wider(names_from = comparison, values_from = log2FC) %>%
    dplyr::rename(early = `3m_to_12m`, late = `12m_to_18m`, net = `3m_to_18m`)

  pcts <- obs %>%
    dplyr::group_by(gene) %>%
    dplyr::summarise(max_abs_pct_diff = max(abs(pct_diff)), .groups = "drop")

  res <- wide %>%
    dplyr::left_join(pcts, by = "gene") %>%
    dplyr::mutate(
      max_abs_log2FC = pmax(abs(early), abs(late), abs(net)),
      passes = abs(early) > te | abs(late) > tl | abs(net) > tn,
      shape = classify(early, net, te, tn),
      sex    = label
    ) %>%
    dplyr::filter(passes) %>%
    dplyr::arrange(dplyr::desc(max_abs_log2FC))

  ## sanity report -- sparsity artefacts show up as absurd fold-changes
  med <- median(res$max_abs_log2FC)
  message(sprintf("  %d genes | median |log2FC| = %.2f | n NA shapes = %d",
                  nrow(res), med, sum(is.na(res$shape))))
  if (med > 2)
    warning(label, ": median |log2FC| = ", round(med, 2),
            " -- still sparsity-driven; raise CFG$min_cells")

  res
}
```


```{r}
male_genes   <- age_changing(neus_male,   "Male")
female_genes <- age_changing(neus_female, "Female")

write.csv(male_genes,   "mature_male_age_changing_genes.csv",   row.names = FALSE)
write.csv(female_genes, "mature_female_age_changing_genes.csv", row.names = FALSE)

## ---------------------------------------------------------------------------
## 6. Cross-sex comparison
##
## Note the shared set is only interpretable if BOTH arms passed the sanity
## check above. A shared list built against a broken male arm is meaningless.
## ---------------------------------------------------------------------------

shared <- dplyr::inner_join(male_genes, female_genes, by = "gene",
                            suffix = c("_M", "_F")) %>%
  dplyr::select(gene, shape_M, shape_F,
                early_M, late_M, net_M, early_F, late_F, net_F,
                max_abs_log2FC_M, max_abs_log2FC_F) %>%
  dplyr::mutate(same_shape = shape_M == shape_F)

write.csv(shared, "mature_comparison_age_changing_genes.csv", row.names = FALSE)

cat(sprintf("\nMale: %d | Female: %d | Shared: %d\n",
            nrow(male_genes), nrow(female_genes), nrow(shared)))
print(table(male_genes$shape,   useNA = "ifany"))
print(table(female_genes$shape, useNA = "ifany"))
```

```{r}
immat_neus_female <- subset(gmps_neus_all, sex == "female" & fine_neu_labels == "immature" & age != "9m")
immat_neus_male <- subset(gmps_neus_all, sex == "male" & fine_neu_labels == "immature" & age != "9m") 

DefaultAssay(immat_neus_female) <- "RNA"
immat_neus_female <- NormalizeData(immat_neus_female)
immat_neus_female <- FindVariableFeatures(immat_neus_female)
immat_neus_female <- ScaleData(immat_neus_female, 
                                  features = VariableFeatures(immat_neus_female))

DefaultAssay(immat_neus_male) <- "RNA"
immat_neus_male <- NormalizeData(immat_neus_male)
immat_neus_male <- FindVariableFeatures(immat_neus_male)
immat_neus_male <- ScaleData(immat_neus_male, 
                                  features = VariableFeatures(immat_neus_male))

male_genes   <- age_changing(immat_neus_male,   "Male")
female_genes <- age_changing(immat_neus_female, "Female")

write.csv(male_genes,   "age_changing_male_v2.csv",   row.names = FALSE)
write.csv(female_genes, "age_changing_female_v2.csv", row.names = FALSE)

## ---------------------------------------------------------------------------
## 6. Cross-sex comparison
##
## Note the shared set is only interpretable if BOTH arms passed the sanity
## check above. A shared list built against a broken male arm is meaningless.
## ---------------------------------------------------------------------------

shared <- dplyr::inner_join(male_genes, female_genes, by = "gene",
                            suffix = c("_M", "_F")) %>%
  dplyr::select(gene, shape_M, shape_F,
                early_M, late_M, net_M, early_F, late_F, net_F,
                max_abs_log2FC_M, max_abs_log2FC_F) %>%
  dplyr::mutate(same_shape = shape_M == shape_F)

write.csv(shared, "age_changing_shared_v2.csv", row.names = FALSE)

cat(sprintf("\nMale: %d | Female: %d | Shared: %d\n",
            nrow(male_genes), nrow(female_genes), nrow(shared)))
print(table(male_genes$shape,   useNA = "ifany"))
print(table(female_genes$shape, useNA = "ifany"))
```

```{r}
CFG <- list(
  adt_min_median = 0,   # require median DSB-normalised signal > this,
                        # in EVERY age group -- placeholder, see note above
  n_perm = 20,
  q      = 0.95
)

## ---------------------------------------------------------------------------
## 1. Feature filter
## ---------------------------------------------------------------------------

adequate_adt <- function(obj, cfg = CFG) {
  dat  <- GetAssayData(obj, assay = "ADT", layer = "data")   # DSB-normalised
  ages <- as.character(obj$age)

  ok <- vapply(AGES, function(a) {
    idx <- ages == a
    apply(dat[, idx, drop = FALSE], 1, median) > cfg$adt_min_median
  }, logical(nrow(dat)))

  keep <- rownames(dat)[rowSums(ok) == length(AGES)]
  message(sprintf("  ADT filter: %d of %d antibodies retained",
                  length(keep), nrow(dat)))
  keep
}

## ---------------------------------------------------------------------------
## 2. Pairwise comparisons (used for both real and permuted labels)
## ---------------------------------------------------------------------------

run_pairwise <- function(obj, features, permute = FALSE, seed = 1) {
  ages <- as.character(obj$age)
  if (permute) { set.seed(seed); ages <- sample(ages) }
  obj$.age_use <- factor(ages, levels = AGES)
  Idents(obj) <- ".age_use"

  purrr::map_dfr(PAIRS, function(p) {
    FindMarkers(obj, ident.1 = p[1], ident.2 = p[2],
                assay = "ADT", features = features,
                logfc.threshold = 0, min.pct = 0,
                test.use = "wilcox", verbose = FALSE) %>%
      tibble::rownames_to_column("gene") %>%
      dplyr::transmute(gene,
                       comparison = paste0(p[2], "_to_", p[1]),
                       log2FC = avg_log2FC)
  })
}

## ---------------------------------------------------------------------------
## 3. Permutation-calibrated thresholds
## ---------------------------------------------------------------------------

calibrate <- function(obj, features, cfg = CFG) {
  purrr::map_dfr(seq_len(cfg$n_perm),
                 ~ run_pairwise(obj, features, permute = TRUE, seed = .x)) %>%
    dplyr::group_by(comparison) %>%
    dplyr::summarise(threshold = quantile(abs(log2FC), cfg$q, na.rm = TRUE),
                     .groups = "drop")
}

## ---------------------------------------------------------------------------
## 4. Trajectory shape, NA-safe
## ---------------------------------------------------------------------------

classify <- function(early, net, thr_e, thr_l, tol = NULL) {
  if (is.null(tol)) tol <- mean(c(thr_e, thr_l)) / 2

  P <- cbind(`3m` = 0, `12m` = early, `18m` = net)   # relative positions
  P <- P - rowMeans(P)                                # centre, as in z-scoring

  L <- ifelse(P >  tol, "high",
       ifelse(P < -tol, "low", "middle"))

  out <- apply(L, 1, paste, collapse = "_")
  out[is.na(early) | is.na(net)] <- NA_character_
  out
}

## ---------------------------------------------------------------------------
## 5. Main
## ---------------------------------------------------------------------------

age_changing_adt <- function(obj, label, cfg = CFG) {
  message("\n=== ", label, " / ADT ===")
  feats <- adequate_adt(obj, cfg)
  if (length(feats) < 5)
    warning(label, ": only ", length(feats), " antibodies pass filter")

  obs <- run_pairwise(obj, feats)
  thr <- calibrate(obj, feats, cfg)
  message("  permutation thresholds (|log2FC|):")
  print(as.data.frame(thr), row.names = FALSE)

  te <- thr$threshold[thr$comparison == "3m_to_12m"]
  tl_step <- thr$threshold[thr$comparison == "12m_to_18m"]
  tn <- thr$threshold[thr$comparison == "3m_to_18m"]

  res <- obs %>%
    dplyr::select(gene, comparison, log2FC) %>%
    tidyr::pivot_wider(names_from = comparison, values_from = log2FC) %>%
    dplyr::rename(early = `3m_to_12m`, late = `12m_to_18m`, net = `3m_to_18m`) %>%
    dplyr::mutate(
      max_abs_log2FC = pmax(abs(early), abs(late), abs(net)),
      passes = abs(early) > te | abs(late) > tl_step | abs(net) > tn,
      shape = classify(early, net, te, tn),
      sex    = label
    ) %>%
    dplyr::filter(passes) %>%
    dplyr::arrange(dplyr::desc(max_abs_log2FC))

  message(sprintf("  %d antibodies pass | median |log2FC| = %.2f",
                  nrow(res), median(res$max_abs_log2FC)))
  res
}

male_adt   <- age_changing_adt(neus_male,   "Male")
female_adt <- age_changing_adt(neus_female, "Female")

write.csv(male_adt,   "age_changing_male_adt.csv",   row.names = FALSE)
write.csv(female_adt, "age_changing_female_adt.csv", row.names = FALSE)

print(table(male_adt$shape,   useNA = "ifany"))
print(table(female_adt$shape, useNA = "ifany"))

shared <- dplyr::inner_join(male_adt, female_adt, by = "gene",
                            suffix = c("_M", "_F")) %>%
  dplyr::select(gene, shape_M, shape_F,
                early_M, late_M, net_M, early_F, late_F, net_F,
                max_abs_log2FC_M, max_abs_log2FC_F) %>%
  dplyr::mutate(same_shape = shape_M == shape_F)

write.csv(shared, "age_changing_shared_adt.csv", row.names = FALSE)

one_ab <- GetAssayData(neus_female, assay = "ADT", layer = "data")["Ms.Ly.6C", ]
hist(one_ab, breaks = 50)   # look for the valley between two modes
```

```{r}

```

