---
title: "Neutrophil ageing in mouse bone marrow"
subtitle: "CITE-seq, male and female, 3-18 months"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
source(here::here("R", "setup.R"))
cfg <- init_project(quiet = TRUE)
load_modules()

# The report reads what the pipeline wrote. It does not recompute anything:
# run `Rscript scripts/run_all.R` first.
have <- function(name) file.exists(table_path(cfg, name)) ||
                       file.exists(object_path(cfg, name))
read_result <- function(name) utils::read.csv(table_path(cfg, name))
```

## Design

Two hashed CITE-seq runs, one per sex, each multiplexing four ages
(`r paste(cfg$analysis$all_age_levels, collapse = ", ")`) across four hashtag
oligos. Every cross-age comparison below uses
`r paste(cfg$analysis$age_levels, collapse = ", ")` only: 9m is not populated
in every cell-type stratum, and including it would make the strata
non-comparable.

| Step | Script | Output |
|---|---|---|
| Preprocess (ADT/HTO, QC, doublets) | `01_preprocess.R` | `<sex>_filtered.rds` |
| Reductions, SingleR, stem/neutrophil subset | `02_annotate.R` | `<sex>_annotated.rds`, `<sex>_neustem.rds` |
| Merge sexes, stage the granulocytes | `03_merge.R` | `gmp_neutrophils.rds` |
| Male vs female within age | `04_sex_differences.R` | `degs_<age>_<assay>.csv` |
| Pseudotime trajectories | `05_trajectory.R` | `<sex>_<age>_trajectory_moran.csv` |
| Age trends per stage | `06_age_trends.R` | `age_trend_<stage>_<sex>.csv` |
| NB-GLM omnibus and interaction | `07_glm_models.R` | `omnibus_<sex>.csv` |
| Sex x age GSEA, per stage | `08_gsea.R` | `gsea_sex_by_age_GOBP_<stage>.csv` |
| Composition and trajectory shifts | `09_development_shifts.R` | `stage_age_trends.csv`, `pseudotime_shifts.csv` |

## Quality control

```{r qc}
#| eval: !expr file.exists(object_path(cfg, "gmp_neutrophils.rds"))
gmp_neu <- read_object(cfg, "gmp_neutrophils.rds")
knitr::kable(table(gmp_neu$fine_neu_labels, gmp_neu$sex),
             caption = "Cells per developmental stage and sex")
knitr::kable(table(gmp_neu$age, gmp_neu$sex),
             caption = "Cells per age and sex")
```

Per-sample QC distributions and doublet scores are in
`results/figures/<sex>_qc.pdf` and `<sex>_doublets.pdf`.

## Granulopoiesis

```{r stages}
#| eval: !expr exists("gmp_neu")
#| fig-width: 12
#| fig-height: 5
Seurat::DimPlot(gmp_neu, reduction = "umap", group.by = "fine_neu_labels") +
  Seurat::DimPlot(gmp_neu, reduction = "umap", group.by = "sex")
```

Stage labels come from clustering the merged GMP + neutrophil object and
mapping cluster ids to maturation stages via
`analysis.neutrophil_cluster_labels` in `config/config.yml`. **Cluster
numbering is not stable across Seurat versions**: if step 3 is re-run, check
`results/figures/gmp_neutrophil_modules.pdf` against
`gmp_neutrophil_clusters.pdf` before trusting the mapping.

## Does development itself shift with age?

Two things the expression analyses cannot see: whether the *mix* of maturation
stages changes, and whether cells sit further along the trajectory. Both are
compared across ages within a sex, which is the design-clean contrast.

```{r composition}
#| eval: !expr have("stage_age_trends.csv")
comp <- read_result("stage_composition.csv")
trends <- read_result("stage_age_trends.csv")

knitr::kable(trends[order(-abs(trends$log_odds_per_step)),
                    c("sex", "stage", "prop_first", "prop_last",
                      "log_odds_per_step", "padj")],
             row.names = FALSE, digits = 3,
             caption = "Change in each stage's share per timepoint step")
```

`log_odds_per_step` is the slope of a stage's share against age rank: positive
means that stage takes up more of the granulocytic compartment in older
animals. `results/figures/stage_composition.pdf` and `stage_trends.pdf` show
the same thing.

```{r pseudotime-shift}
#| eval: !expr have("pseudotime_shifts.csv")
shifts <- read_result("pseudotime_shifts.csv")
knitr::kable(shifts[, c("sex", "group", "n", "median_shift", "wasserstein",
                        "frac_beyond_reference_median", "ks_padj")],
             row.names = FALSE, digits = 3,
             caption = "Pseudotime distribution relative to the youngest age")
```

`frac_beyond_reference_median` is the share of that group's cells sitting past
the youngest group's median pseudotime; it is 0.5 for the reference by
construction, so anything above that is cells having moved further along.
`wasserstein` is the average shift in pseudotime units, which unlike the KS
statistic is comparable between rows.

The KS p-values are computed over cells, which are pseudoreplicates: read the
distances, not the p-values. `potency_shifts.csv` asks the same question of
CytoTRACE2 potency, which does not depend on where the trajectory root landed
— if both move the same way, the shift is not an artefact of the fit.

## Sex differences within each age

```{r sex-degs}
#| eval: !expr have(sprintf("degs_%s_RNA.csv", cfg$analysis$age_levels[1]))
degs <- lapply(stats::setNames(cfg$analysis$age_levels, cfg$analysis$age_levels),
               function(a) read_result(sprintf("degs_%s_RNA.csv", a)))

summary_tbl <- data.frame(
  age = names(degs),
  up_in_male   = vapply(degs, function(d) sum(d$p_val_adj < 0.05 & d$avg_log2FC > 0), integer(1)),
  up_in_female = vapply(degs, function(d) sum(d$p_val_adj < 0.05 & d$avg_log2FC < 0), integer(1))
)
knitr::kable(summary_tbl, row.names = FALSE,
             caption = "Significant genes by direction (ident.1 = male)")
```

## Age trends by stage

```{r age-trends}
#| eval: !expr have(sprintf("age_trend_mature_%s.csv", cfg$analysis$sex_levels[1]))
files <- list.files(cfg$paths$tables, pattern = "^age_trend_.*\\.csv$", full.names = TRUE)
trend_summary <- do.call(rbind, lapply(files, function(f) {
  parts <- strsplit(sub("\\.csv$", "", basename(f)), "_")[[1]]
  d <- utils::read.csv(f)
  data.frame(stage = parts[3], sex = parts[4], n_changing = nrow(d))
}))
knitr::kable(trend_summary, row.names = FALSE,
             caption = sprintf("Genes with |rho| > %.2f against age",
                               cfg$age_trend$rho_cutoff))
```

Effect-size distributions for the two sexes: `results/figures/rho_distributions.pdf`.
Pseudobulk PCA per stage: `results/figures/pseudobulk_pca_by_stage.pdf`.

## Sex x age interaction

Run separately within each maturation stage. Pooling stages would let a
composition shift masquerade as a within-cell age trend.

```{r gsea}
#| eval: !expr have(sprintf("gsea_sex_by_age_GOBP_%s.csv", cfg$gsea$stages[1]))
gsea <- read_result(sprintf("gsea_sex_by_age_GOBP_%s.csv", cfg$gsea$stages[1]))
top <- gsea[gsea$padj < 0.05 & gsea$independent, ]
knitr::kable(utils::head(top[, c("pathway", "NES", "padj", "signal_type",
                                 "dominant_pattern")], 20),
             row.names = FALSE, digits = 3,
             caption = sprintf("Independent GO:BP hits, sex x age interaction (%s)",
                               cfg$gsea$stages[1]))
```

A positive NES means the age trajectory is **more positive in males** than in
females. It does not mean the pathway is higher in males, and it does not mean
males age faster: `signal_type` and `dominant_pattern` say which of those the
leading edge actually supports. Pathways to report are the ones flagged
`robust` in `gsea_sensitivity_comparison.csv`, which survive removing the
abundant granule transcripts.

## Caveats

- Cells are pseudoreplicates: one hashtag per age means no biological
  replication, so per-cell p-values overstate the evidence. Everything selected on a
  permutation-calibrated threshold (steps 6 and 7) accounts for this; anything
  selected on a BH-adjusted Wilcoxon p-value does not.
- One 10x run per sex means sex is completely confounded with batch. Sex
  differences here cannot be separated from run-to-run technical variation.
- No ambient-RNA correction is applied. Highly expressed granule transcripts
  are the ones most likely to bleed between droplets, and they are also the
  ones carrying the maturation signal, so read stage-specific results for
  those genes with that in mind.

```{r session}
sessionInfo()
```
