---
title: "Neutrophil ageing in mouse bone marrow"
subtitle: "CITE-seq, one 10x run per sex, four ages (3, 9, 12, 18 months)"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
    fig-width: 10
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
# The report reads what the pipeline wrote and recomputes nothing. Run the
# pipeline (or at least step 13) first: `Rscript scripts/run_all.R`.
source(here::here("R", "setup.R"))
cfg <- load_config()
fig_dir <- file.path(cfg$paths$figures, "summary")

show_fig <- function(id) {
  f <- list.files(fig_dir, paste0("^", id, "_.*\\.png$"), full.names = TRUE)
  if (length(f)) knitr::include_graphics(f[1])
  else knitr::asis_output(sprintf(
    "\n> **%s not drawn.** Its table has not been produced; see `%s/FIGURES.md`.\n",
    id, fig_dir))
}
read_result <- function(name) {
  path <- table_path(cfg, name)
  if (file.exists(path)) utils::read.csv(path, stringsAsFactors = FALSE) else NULL
}
show_table <- function(df, cols = NULL, digits = 2, n = 20, caption = NULL) {
  if (is.null(df) || !nrow(df))
    return(knitr::asis_output("\n> *Table not produced by this run.*\n"))
  if (!is.null(cols)) df <- df[, intersect(cols, names(df)), drop = FALSE]
  knitr::kable(utils::head(df, n), digits = digits, caption = caption)
}
```

## What this can and cannot say

Each sex is one 10x library, so **every male-versus-female number is also a
library-versus-library number**. Ages are hashed *within* a library, so **age
comparisons within a sex are free of batch**. Each age is one hashtag pooling
six mice: cells are pseudoreplicates, intervals describe sampling of cells, and
no statistic here describes variation between mice. The results are a ranked
set of hypotheses for per-mouse validation, not findings.

## 1. What the data can be trusted for

```{r}
show_fig("fig01")
```

Depth by stage, age and library, on the counts the analyses read. Strata far
below the rest were not matched by the old single global target; matching is
now done inside each stratum (`fig02`).

```{r}
show_fig("fig02")
show_fig("fig03")
```

RNA and protein disagree most at the immature/mature boundary. Stage-resolved
results inherit that uncertainty.

## 2. How the stage mix changes with age

```{r}
show_fig("fig04")
show_fig("fig05")
show_table(read_result("stage_age_trends.csv"),
           c("sex", "stage", "prop_first", "prop_last", "log_odds_per_step", "padj"),
           caption = "Log-odds change in each stage's share per age step")
show_fig("fig06")
```

## 3. Did RNA QC remove the cells? (protein gate)

The male library loses immature and mature cells with age, and those are the
lowest-RNA cells. Surface protein does not depend on RNA capture, so step 12
stages every hashtagged singlet by Ly6G / CD11b / CXCR2, including the cells RNA
QC removed. **If protein-defined mature cells are present before QC and lost
after it, the RNA trend is QC loss.**

```{r}
show_fig("fig10")
show_fig("fig09")
show_table(read_result("protein_stage_age_trends.csv"),
           c("cell_set", "sex", "stage", "prop_first", "prop_last", "log_odds_per_step"),
           caption = "Protein-stage share per age step, before and after RNA QC")
```

## 4. Upstream: the progenitor compartment

```{r}
show_fig("fig07")
show_fig("fig08")
```

## 5. Where cells sit along the trajectory

```{r}
show_fig("fig11")
show_fig("fig12")
```

Read the pooled panel against the within-stage panels: a pooled shift with
flat within-stage panels is the stage mix moving, not the cells.

## 6. What changes in expression

```{r}
show_fig("fig13")
show_fig("fig14")
show_fig("fig15")
show_fig("fig16")
show_fig("fig17")
show_fig("fig18")
show_fig("fig19")
show_fig("fig20")
show_fig("fig21")
```

## 7. The sex contrast

Not a test of sex: a ranked list of candidates for orthogonal assays. The
deeper library is female, so a female-high result is what the confound
predicts; male-high results that survive depth matching rank first.

```{r}
show_fig("fig22")
show_fig("fig23")
```

## Figure index

```{r}
idx <- file.path(fig_dir, "FIGURES.md")
if (file.exists(idx)) knitr::asis_output(paste(readLines(idx)[-(1:2)], collapse = "\n"))
```
